---
title: "W7. C++ STL Iterators, SOLID Principles"
author: "Eugene Zouev, Munir Makhmutov"
date: "March 5, 2026"
format: html
engine: knitr
---

#### **1. Summary**

##### **1.1 Generic Programming and STL**

###### **1.1.1 What is Generic Programming?**

**Generic programming (GP)** is a programming paradigm that focuses on writing algorithms and data structures in the most general form possible — independent of any specific type. The two fundamental principles behind GP are **generality** (algorithms work for any type) and **efficiency** (no runtime overhead compared to hand-written type-specific code).

The **Standard Template Library (STL)** is the most successful implementation of generic programming. It was developed by Alexander Stepanov, Meng Lee, and David Musser in the early 1990s and became part of the C++ Standard in **1994**. By **1998**, it was formally standardized in Chapters 20, 22, 23, and 25 of the ISO C++ Standard.

<!-- DIAGRAM HERE -->

###### **1.1.2 GP vs. OOP — An Important Distinction**

STL does *not* use object-oriented programming (OOP). These are *orthogonal* paradigms — they solve different problems:

*   **OOP** bundles data and the algorithms that operate on that data inside a single **class**. The class is simultaneously a container *and* a collection of algorithms.
*   **GP** separates containers from algorithms completely:
    *   **Generic Containers** — data structures that know nothing about any specific algorithm.
    *   **Generic Algorithms** — functions that know nothing about any specific container.

The key insight is that GP requires no encapsulation or inheritance between containers and algorithms. The two halves are decoupled and communicate only through a thin interface: **iterators**.

###### **1.1.3 STL Components**

STL provides seven kinds of components, split into two groups:

**First-order components** (the core):

*   **Containers** — generic data structures: `vector`, `list`, `deque`, `set`, `map`, `multiset`, `multimap`, and their variants.
*   **Algorithms** — generic operations on containers: searching, sorting, copying, transforming, merging, and more.
*   **Iterators** — the glue between containers and algorithms (explained in detail below).

**Second-order components** (support infrastructure):

*   **Functional Objects** (functors) — callable objects that algorithms use as customization points.
*   **Adaptors** — wrappers that adapt containers, iterators, or functors to a different interface.
*   **Allocators** — policy objects controlling memory allocation.
*   **Traits** — compile-time information about types.

##### **1.2 Iterators: The Bridge Between Containers and Algorithms**

###### **1.2.1 The Motivation: Deriving Iterators from First Principles**

The best way to understand *why* iterators exist is to derive them from scratch. Consider the following concrete function that searches an array of integers for a given value:

```cpp
const int* find0(const int* array, int n, int x)
{
    const int* p = array;
    for (int i = 0; i < n; i++)
    {
        if (*p == x) return p;  // success
        p++;
    }
    return 0;  // fail
}
```

This works, but it has four hard-coded limitations:

1.  It searches only `int` values.
2.  It searches only in an *array* (a contiguous memory block).
3.  It requires the *address* of the first element.
4.  It requires the array *size*.

Let's remove each limitation one step at a time.

**Step 1 — Generalize the element type:**

```cpp
template <typename T>
T* find1(T* array, int n, const T& x)
{
    T* p = array;
    for (int i = 0; i < n; i++)
    {
        if (*p == x) return p;
        p++;
    }
    return 0;
}
```

Now the function works for any type `T`, as long as `T` supports `operator==`. (We pass `x` by reference to avoid expensive copies.)

**Step 2 — Replace size with a "past-the-end" pointer:**

Instead of passing the size `n`, we pass a pointer `beyond` that points to the memory location *one past* the last element. This removes the array-size dependency:

```cpp
template <typename T>
T* find2(T* array, T* beyond, const T& x)
{
    T* p = array;
    while (p != beyond)
    {
        if (*p == x) return p;
        p++;
    }
    return 0;
}
```

<!-- DIAGRAM HERE -->

The ISO C++ Standard explicitly guarantees this is safe: a pointer one past the last element of an array is a valid pointer for comparison purposes, even though you may not dereference it.

**Step 3 — Return `beyond` instead of `0` on failure:**

Returning a null pointer (`0`) ties us to pointers specifically. We generalize by returning `beyond` to signal "not found" — this makes the failure signal expressible in terms of *any* iterator type, not just pointers:

```cpp
template <typename T>
T* find3(T* array, T* beyond, const T& x)
{
    T* p = array;
    while (p != beyond)
    {
        if (*p == x) return p;
        p++;
    }
    return beyond;  // "not found" — valid for any iterator type
}
```

**Step 4 — Simplify and optimize:**

We merge the two loop conditions into one and rename `array` to `first` (since the algorithm no longer assumes a specific data structure):

```cpp
template <typename T>
T* find4(T* first, T* beyond, const T& x)
{
    T* p = first;
    while (p != beyond && *p != x)
        p++;
    return p;
}
```

**Step 5 — Remove pointers entirely (the final algorithm):**

The final leap: replace the pointer type `T*` with a new abstract type parameter `P`. This `P` can be *any* type that behaves like a pointer — including user-defined classes for non-array containers:

```cpp
template <typename T, typename P>
P find5(P first, P beyond, const T& x)
{
    P p = first;
    while (p != beyond && *p != x)
        p++;
    return p;
}
```

Type `P` is now the **iterator type**. Type `T` is the **value type** (the type of elements stored in the container). The algorithm requires only three operations from `P`: dereferencing (`operator*`), increment (`operator++`), and inequality comparison (`operator!=`).

###### **1.2.2 Using `find5`: The Half-Open Range Convention**

The pattern `find5(first, beyond, x)` establishes the **half-open range** convention: `[first, beyond)`. This means:

*   `first` points to the first element.
*   `beyond` points *one past* the last element (it is *not* dereferenceable).
*   The range contains all elements from `first` up to but *not including* `beyond`.

<!-- DIAGRAM HERE -->

For a plain C++ array this looks like:

```cpp
int A[100];
// ... initialize A ...
int* result = find5(A, &A[100], 7);
// T = int, P = int* (deduced by the compiler)
```

`&A[100]` is the canonical "past-the-end" pointer for an array of 100 elements.

###### **1.2.3 Iterators as a Formal Concept**

The derivation establishes two fundamental concepts:

*   **Value type** (`T`): the type of the elements stored in the container. Requirements: must support `operator!=`.
*   **Iterator type** (`P`): a type that provides access to a sequence of elements. Requirements: must support `operator*` (returning a value of the value type), `operator++` (advancing to the next element), and `operator!=`.

An **iterator** is any object whose type satisfies these requirements. Raw C++ pointers (`T*`) automatically satisfy all of them — which is why they work as iterators for arrays.

##### **1.3 Building a Custom Iterator: From Array to Linked List**

The power of iterators becomes clear when we apply `find5` to a *non-array* container — a singly-linked list.

###### **1.3.1 A Simple Linked List Node**

Here is a minimal `NODE` class for a singly-linked list of integers:

```cpp
class NODE {
    int value;   // the node's data
    NODE* next;  // pointer to the next node
public:
    NODE(int v) : value(v), next(0) { }
    void add(NODE* n) {
        n->next = this->next;
        this->next = n;
    }
};
```

<!-- DIAGRAM HERE -->

To use `find5` with this list, we need:

1.  `NODE` to satisfy the **value type** requirements (i.e., support `operator!=`).
2.  A separate **iterator class** that satisfies the iterator type requirements.

###### **1.3.2 Adding Value-Type Requirements to NODE**

The `find5` algorithm compares elements with `operator!=`. We add it to `NODE`:

```cpp
class NODE {
    int value;
    NODE* next;
public:
    NODE(int v) : value(v), next(0) { }
    void add(NODE* n) {
        n->next = this->next;
        this->next = n;
    }

    // Value type requirement: compare by value
    bool operator!=(NODE& n) { return this->value != n.value; }
    // Conversion to int for convenient value access
    operator int() { return value; }
};
```

###### **1.3.3 Writing the Iterator Class**

The iterator for `NODE`s must support `operator*`, `operator++`, and `operator!=`. It can be written as either a standalone class or as a **nested (inner) class** inside `NODE`. Using an inner class is usually better because it gives the iterator direct access to `NODE`'s private members.

```cpp
class NODE {
    int value;
    NODE* next;
public:
    NODE(int v)  : value(v), next(0) { }
    void add(NODE* n) { n->next = this->next; this->next = n; }
    bool operator!=(NODE& n) { return this->value != n.value; }
    operator int() { return value; }

    class iterator;
    friend class iterator;   // give iterator access to private members

    class iterator {
        NODE* p;             // current position
    public:
        iterator(NODE* node) : p(node) { }

        int operator*()  { return p->value; }   // dereference: get value
        void operator++() { p = p->next; }      // advance to next node
        bool operator!=(const iterator& other) const {
            return p != other.p;
        }
        operator NODE*() { return p; }          // conversion to raw pointer
    };
};
```

Now we can use `find5` on a linked list:

```cpp
NODE* list = new NODE(1);
// ... add more nodes ...
NODE::iterator it = find5(NODE::iterator(list), NODE::iterator(0), 7);
```

<!-- DIAGRAM HERE -->

###### **1.3.4 The Big Picture: Separation of Concerns**

The following illustrates what makes this design powerful:

*   The **array container** and the **list container** know *nothing* about the `find5` algorithm.
*   The **`find5` algorithm** knows *nothing* about arrays or lists.
*   The two sides simply agree on a common interface — the iterator requirements — and both fulfill it independently.

This means you can add new algorithms without touching containers, and add new containers without touching algorithms. The number of combinations grows *linearly* (containers + algorithms), not *quadratically* (containers × algorithms).

<!-- DIAGRAM HERE -->

##### **1.4 The `reverse` Algorithm: Introducing New Iterator Requirements**

Consider a `reverse` algorithm that reverses the order of elements in a container:

```cpp
template <typename P, typename T>
void reverse(P start, P beyond)
{
    while (start != beyond)
    {
        --beyond;
        if (start != beyond)
        {
            T t = *start;
            *start = *beyond;
            *beyond = t;
            ++start;
        }
    }
}
```

<!-- DIAGRAM HERE -->

This algorithm requires *more* from the iterator than `find5` did:

*   `operator++` — same as before.
*   `operator!=` — same as before.
*   `operator--` — **new**: the iterator must be able to move *backward*.
*   `operator*` returning a **reference** (not just a value) — **new**: we need to *write* through the iterator, not just read.

This observation leads to the formal classification of iterator categories.

##### **1.5 Iterator Categories**

Different algorithms require different capabilities from their iterators. The STL formalizes this as a **hierarchy of iterator categories**, where each level adds new operations on top of the previous:

<!-- DIAGRAM HERE -->

```{r, echo=FALSE, fig.cap="Iterator Category Hierarchy"}
#| fig-align: center
library(DiagrammeR)
grViz("
  digraph iterator_hierarchy {
    graph [rankdir=TB, bgcolor='#FAFAFA', fontname='Helvetica']
    node [shape=box, style='rounded,filled', fontname='Helvetica', fontsize=11]
    edge [fontname='Helvetica', fontsize=9]

    Input [label='Input Iterator\\n(x = *i; ++i; i++)', fillcolor='#D6EAF8']
    Output [label='Output Iterator\\n(*i = x; ++i; i++)', fillcolor='#D6EAF8']
    Forward [label='Forward Iterator\\n(Input + Output)', fillcolor='#D5F5E3']
    Bidir [label='Bidirectional Iterator\\n(Forward + --i; i--)', fillcolor='#FDEBD0']
    Random [label='Random Access Iterator\\n(Bidir + i+n, i-n, i[n], i<j, ...)', fillcolor='#FADBD8']
    Pointer [label='C++ Raw Pointer (T*)', fillcolor='#F9E79F', shape=ellipse]

    Input -> Forward
    Output -> Forward
    Forward -> Bidir
    Bidir -> Random
    Random -> Pointer [label=' is-a / subsumes']
  }
", width=600, height=380)
```

*   **Input Iterator** — can only be read and moved forward, one step at a time. Useful for single-pass algorithms like `find`. Operations: `x = *i`, `++i`, `i++`, `i == j`, `i != j`.
*   **Output Iterator** — can only be written to and moved forward. Operations: `*i = x`, `++i`, `i++`.
*   **Forward Iterator** — combines Input and Output. Supports multiple passes over the same range. Used by `forward_list`.
*   **Bidirectional Iterator** — Forward + backward movement (`--i`, `i--`). Used by `list`, `set`, `map`. Required by `reverse`.
*   **Random Access Iterator** — Bidirectional + arbitrary jumps (`i + n`, `i - n`, `i += n`, `i[n]`) and comparison (`i < j`, `i <= j`, etc.). Used by `vector`, `deque`, raw arrays. Required by `sort`.

All iterator categories support `operator==`, `operator!=`, and assignment (`=`).

**Key insight:** C++ raw pointers (`T*`) satisfy all Random Access Iterator requirements — they are the "perfect" iterator. Every STL algorithm that works on iterators also works on raw pointers.

##### **1.6 Iterator Behavior Across Containers**

Different containers provide different iterator categories, and their traversal order depends on the container's internal structure:

```cpp
int main() {
    std::map<int, std::string> orderedMap;
    std::unordered_map<int, std::string> unorderedMap;

    orderedMap.insert({3, "Three"}); unorderedMap.insert({3, "Three"});
    orderedMap.insert({1, "One"});   unorderedMap.insert({1, "One"});
    orderedMap.insert({2, "Two"});   unorderedMap.insert({2, "Two"});

    // std::map iterates in sorted key order: 1, 2, 3
    for (auto it = orderedMap.begin(); it != orderedMap.end(); ++it)
        std::cout << it->first << ": " << it->second << "\n";

    // std::unordered_map iterates in hash-bucket order (unpredictable)
    for (auto it = unorderedMap.begin(); it != unorderedMap.end(); ++it)
        std::cout << it->first << ": " << it->second << "\n";
}
```

*   `std::map` provides **Bidirectional Iterators** and always traverses in *sorted key order*.
*   `std::unordered_map` provides **Forward Iterators** and traverses in an *implementation-defined* order.

##### **1.7 Iterator Use-Case: Converting Between Containers**

A powerful use of iterators is constructing one container from another. Many STL containers accept a pair of iterators in their constructor, creating a copy of the given range:

```cpp
#include <iostream>
#include <vector>
#include <set>

int main() {
    std::vector<int> nums = {1, 2, 2, 3, 4, 4, 5, 6, 7, 8, 9, 9};

    // std::multiset: sorted, duplicates allowed
    std::multiset<int> multiSetNums(nums.begin(), nums.end());

    // std::set: sorted, duplicates removed automatically
    std::set<int> uniqueNums(nums.begin(), nums.end());

    for (int num : uniqueNums)
        std::cout << num << " ";   // 1 2 3 4 5 6 7 8 9
    std::cout << std::endl;

    for (int num : multiSetNums)
        std::cout << num << " ";   // 1 2 2 3 4 4 5 6 7 8 9 9
}
```

Here, `nums.begin()` and `nums.end()` are `vector` iterators, but the `set` and `multiset` constructors accept *any* Input Iterator. This is GP in action: the constructor algorithm is decoupled from the source container type.

##### **1.8 Defining a Custom Iterator Inside a Container Class**

The recommended pattern for providing iterators in your own container is to define the iterator as an **inner class** with access to the container's internals via `friend`:

```cpp
template<typename T, typename K /*, ... */>
class MyContainer {
    /* ... internal storage ... */
public:
    /* constructors, destructors, member functions */

    class iterator;
    friend class iterator;

    class iterator {
        /* ... */
    public:
        using pointer   = T*;
        using reference = T&;

        iterator& operator++();   // advance
        reference operator*();    // dereference
        bool operator!=(const iterator&) const;

    private:
        pointer ptr_;    // current position
        pointer end_;    // one past end
        pointer begin_;  // start of storage
        bool last_;      // sentinel flag
    };

    iterator begin();
    iterator end();
    /* ... */
};
```

The `friend class iterator` declaration gives the inner `iterator` class full access to `MyContainer`'s private members, which is necessary for efficient implementation.

##### **1.9 SOLID Principles**

Before diving into SOLID, it helps to know a few other common programming abbreviations you will encounter in the industry:

*   **KISS** — *Keep It Simple Stupid*: prefer simple solutions over complex ones.
*   **DRY** — *Don't Repeat Yourself*: avoid duplicating logic.
*   **RAII** — *Resource Acquisition Is Initialization*: tie resource lifetimes to object lifetimes (a core C++ idiom).
*   **SFINAE** — *Substitution Failure Is Not An Error*: a C++ template mechanism.

**SOLID** is a set of five design principles for building *flexible*, *maintainable*, and *extensible* object-oriented software. They were formalized by **Robert C. Martin** ("Uncle Bob") and are widely used across all OOP languages.

The five letters stand for:

```{r, echo=FALSE, fig.cap="SOLID Principles Overview"}
#| fig-align: center
library(DiagrammeR)
grViz("
  digraph solid {
    graph [bgcolor='#FAFAFA', fontname='Helvetica']
    node [shape=box, style='rounded,filled', fontname='Helvetica', fontsize=11, fillcolor='#D5F5E3']
    edge [style=invis]
    S [label='S — Single Responsibility\\nOne class, one reason to change']
    O [label='O — Open/Closed\\nOpen for extension, closed for modification']
    L [label='L — Liskov Substitution\\nSubtypes must be substitutable for supertypes']
    I [label='I — Interface Segregation\\nClients should not depend on unused methods']
    D [label='D — Dependency Inversion\\nDepend on abstractions, not concretions']
    S -> O -> L -> I -> D
  }
", width=600, height=320)
```

###### **1.9.1 S — Single Responsibility Principle (SRP)**

**A class should have only one reason to change.**

In practice, this means a class should perform only *one kind of task* — it should have only *one responsibility*. If a class handles both navigation *and* printing, then any change to the printing logic forces a change to the class, even if the navigation logic is untouched (and vice versa).

**Problem — SRP violated:**

```csharp
class Report
{
    public string text;

    public void GoToFirstPage() { ... }  // Navigation responsibility
    public void GoToLastPage()  { ... }
    public void GoToPage(int n) { ... }

    public void Print() { /* print text */ }  // Printing responsibility — wrong!
}
```

`Report` has two distinct reasons to change: if we change how navigation works, or if we add a new output format (XML, RTF, printer, console). This makes the class fragile and hard to reuse.

**Solution — extract the printing responsibility:**

```csharp
interface IOut {
    void print(string text);
}

class ToConsole : IOut {
    public void print(string text) { /* print to console */ }
}

class ToPrinter : IOut {
    public void print(string text) { /* send to printer */ }
}

// Report now has ONE responsibility: navigation
class Report
{
    public string text;

    public void GoToFirstPage() { ... }
    public void GoToLastPage()  { ... }
    public void GoToPage(int n) { ... }

    // Delegates printing to the injected IOut object
    public void Print(IOut p) {
        p.print(this.text);
    }
}
```

Now `Report` has a single reason to change: navigation. The `IOut` hierarchy can evolve independently.

<!-- DIAGRAM HERE -->

###### **1.9.2 O — Open/Closed Principle (OCP)**

**Software components should be open for extension but closed for modification.**

This means you should be able to add new behaviour to a system *without editing existing, tested code*. The classic violation is a class that must be *modified* every time a new variant is required.

**Problem — OCP violated:**

```csharp
class Cook
{
    public string name;

    public void MakeSoup()   { ... }
    // Need a salad? Must modify this class → VIOLATION
    // Need a dessert? Must modify this class → VIOLATION
}
```

Every new dish requires modifying `Cook`. This risks breaking existing behaviour and forces retesting of already-working code.

**Solution 1 — Use inheritance:**

```csharp
class Cook
{
    public string name;
    public void MakeSoup() { ... }  // closed for modification
}

class SmarterCook : Cook
{
    public void MakeSalad() { ... }  // extension without modification
}
```

**Solution 2 — Separate the algorithm from the entity (Strategy pattern):**

```csharp
interface IMeal
{
    void Make();
}

class Soup   : IMeal { void Make() { /* making soup */   } }
class Salad  : IMeal { void Make() { /* making salad */  } }
class Dessert: IMeal { void Make() { /* making dessert */} }

class Cook
{
    public string name;

    // Cook is CLOSED for modification: this method never changes
    public void Make(IMeal meal) {
        meal.Make();
    }
}
```

Adding a new dish is done by creating a new class that implements `IMeal` — `Cook` is never touched.

<!-- DIAGRAM HERE -->

###### **1.9.3 L — Liskov Substitution Principle (LSP)**

**A variable of a given type may be assigned a value of any subtype of that type, and a method with a parameter of a given type may be invoked with an argument of any subtype, without altering the correctness of the program.**

This principle, formulated by **Barbara Liskov**, is essentially a correctness condition on inheritance. It is *stronger* than simply having the right method signatures — it requires that a subtype *behave* as expected when used in place of its supertype.

**Formal statement:** If `S` is a subtype of `T`, then objects of type `S` may be substituted for objects of type `T` without altering the desirable properties of the program.

**Example of LSP holding:**

```java
Number n = new Integer(5);   // Integer is a subtype of Number — OK
Number m = new Double(3.14); // Double is a subtype of Number — OK
```

Any code that works on `Number` will work correctly when given an `Integer` or `Double`.

**Problem — LSP violated:**

```java
// Base class
public class Document {
    private String title;
    private int pages;

    public void setData(String title) {
        this.title = title;   // Only changes the title
    }
}

// Subclass that VIOLATES LSP
public class SpecialDocument extends Document {
    @Override
    public void setData(String title) {
        super.setData(title);
        setPages(20);   // Secretly changes pages too! Unexpected side effect.
    }
}
```

A caller that uses `Document doc = new SpecialDocument(...)` and calls `doc.setData("NewTitle")` will be *surprised* to find that the page count changed. The subclass silently breaks the contract of the parent class.

**Correct usage:**

```java
Document doc1 = new Document("MyDoc", 10);
Document doc2 = new SpecialDocument("MySpecialDoc", 10);

doc2.setData("MyDoc");
// doc1: pages = 10 (expected)
// doc2: pages = 20 (SURPRISE — LSP violation!)
```

The fix is to ensure `SpecialDocument.setData` has the *same observable effect* as `Document.setData` — or to reconsider whether `SpecialDocument` should inherit from `Document` at all.

<!-- DIAGRAM HERE -->

###### **1.9.4 I — Interface Segregation Principle (ISP)**

**Clients should not be forced to depend on methods they do not use.**

A large, monolithic interface forces every implementing class to provide implementations for methods it may not need, leading to either empty/stub implementations or artificially constrained designs.

**Problem — ISP violated:**

```csharp
interface IMessage
{
    void Send();
    string Text    { get; set; }
    string Subject { get; set; }
    string ToAddress   { get; set; }
    string FromAddress { get; set; }
    // byte[] Voice { get; set; }  ← may also be needed for voice messages
}

class SmsMessage : IMessage
{
    public void Send() { ... }
    public string Text    { get { ... } set { ... } }
    public string Subject { ... }  // SMS has no Subject — forced to implement it anyway!
    public string ToAddress   { get { ... } set { ... } }
    public string FromAddress { get { ... } set { ... } }
}

class VoiceMessage : IMessage
{
    public void Send() { ... }
    public string Text    { ... }  // Voice has no text — forced to implement it anyway!
    public string Subject { ... }  // Voice has no subject — forced to implement it anyway!
    public string ToAddress   { get { ... } set { ... } }
    public string FromAddress { get { ... } set { ... } }
}
```

**Solution — split the fat interface into focused ones:**

```csharp
interface IMessage
{
    void Send();
    string ToAddress   { get; set; }
    string FromAddress { get; set; }
}

interface IVoiceMessage : IMessage
{
    byte[] Voice { get; set; }
}

interface ITextMessage : IMessage
{
    string Text { get; set; }
}

interface IEmailMessage : ITextMessage
{
    string Subject { get; set; }
}
```

Now each implementing class only depends on the interface(s) it actually needs:

*   `VoiceMessage implements IVoiceMessage` — gets `Send`, `ToAddress`, `FromAddress`, `Voice`.
*   `SmsMessage implements ITextMessage` — gets `Send`, `ToAddress`, `FromAddress`, `Text`.
*   `EmailMessage implements IEmailMessage` — gets everything including `Subject`.

<!-- DIAGRAM HERE -->

###### **1.9.5 D — Dependency Inversion Principle (DIP)**

**High-level modules should not depend on low-level modules. Both should depend on abstractions. Abstractions should not depend on details; details should depend on abstractions.**

In practice, this means that a class should not directly instantiate the specific implementations it uses (e.g., `new SaveDocumentToPdf()`). Instead, it should work with *abstractions* (interfaces), and the concrete implementations are *injected* from the outside.

The OCP's Solution 2 above (the `Cook`/`IMeal` example) is also an instance of DIP: `Cook` depends on the `IMeal` *abstraction*, not on any concrete `Soup` or `Salad` class.

A common technique for DIP is **Dependency Injection (DI)**: the dependencies of a class are passed in (via constructor, method parameter, or a DI framework) rather than being created inside the class:

```java
// Without DI — high-level depends on low-level concrete class
public class DocumentProcessor {
    private SaveDocumentToPdf saver = new SaveDocumentToPdf(); // tightly coupled!
    public void process(Document doc) { saver.save("output", doc); }
}

// With DI — depends on abstraction (Saver interface)
public class DocumentProcessor {
    private final Saver saver;
    public DocumentProcessor(Saver saver) { this.saver = saver; } // injected!
    public void process(Document doc) { saver.save("output", doc); }
}
```

The second version can be tested with a mock `Saver`, can switch from PDF to DOCX without code changes, and follows both OCP and DIP.

#### **2. Definitions**

*   **Generic Programming (GP)**: A programming paradigm focused on writing algorithms and data structures in the most general form, parameterized over types, to achieve both generality and efficiency without runtime overhead.
*   **STL (Standard Template Library)**: The C++ implementation of generic programming, providing containers, algorithms, and iterators as decoupled, reusable components. Part of the C++ Standard since 1994.
*   **Container**: An STL data structure (e.g., `vector`, `list`, `set`, `map`) that stores elements but knows nothing about the algorithms that operate on it.
*   **Algorithm**: An STL function template (e.g., `find`, `sort`, `reverse`) that operates on element sequences but knows nothing about the specific container type.
*   **Iterator**: An object (or pointer) that provides a uniform interface for accessing and traversing the elements of a container. The bridge between containers and algorithms.
*   **Value type**: The type of elements stored in a container. In `find5<T, P>`, `T` is the value type.
*   **Iterator type**: The type of the iterator object itself. In `find5<T, P>`, `P` is the iterator type.
*   **Half-open range**: The convention `[first, beyond)` used by STL: `first` points to the first element; `beyond` points one past the last. `beyond` is valid to hold and compare against but must not be dereferenced.
*   **Input Iterator**: An iterator that supports one-pass forward traversal and reading only (`x = *i; ++i`). Used by algorithms like `find`.
*   **Output Iterator**: An iterator that supports one-pass forward traversal and writing only (`*i = x; ++i`). Used as output by algorithms like `copy`.
*   **Forward Iterator**: Combines Input and Output iterators; supports multiple passes. Used by `forward_list`.
*   **Bidirectional Iterator**: A Forward Iterator that also supports backward movement (`--i`). Used by `list`, `set`, `map`. Required by `reverse`.
*   **Random Access Iterator**: A Bidirectional Iterator that also supports arbitrary jumps (`i + n`, `i[n]`, `i < j`). Used by `vector`, arrays. Required by `sort`.
*   **Inner (nested) class**: A class declared inside another class. An iterator defined as an inner class of a container can access the container's private members via the `friend` declaration.
*   **SOLID**: An acronym for five object-oriented design principles: Single Responsibility, Open/Closed, Liskov Substitution, Interface Segregation, and Dependency Inversion.
*   **Single Responsibility Principle (SRP)**: A class should have only one reason to change — it should perform only one kind of task.
*   **Open/Closed Principle (OCP)**: Software components should be open for extension (new behaviour can be added) but closed for modification (existing code is not changed).
*   **Liskov Substitution Principle (LSP)**: Objects of a subtype can be substituted for objects of the supertype without altering the correctness of the program.
*   **Interface Segregation Principle (ISP)**: Clients should not be forced to depend on interfaces they do not use. Large interfaces should be split into smaller, focused ones.
*   **Dependency Inversion Principle (DIP)**: High-level modules should depend on abstractions (interfaces), not on concrete low-level modules. Concrete implementations are injected rather than instantiated directly.
*   **Dependency Injection (DI)**: A technique implementing DIP by passing (injecting) dependencies into a class from the outside, rather than creating them inside the class.

#### **3. Examples**

##### **3.1. Trace the `find5` Derivation Steps** (Lecture 7, Example 1)

Starting from the concrete `find0` function below, explain each of the five generalization steps that lead to the generic `find5` algorithm. For each step, state what was changed and why.

```cpp
const int* find0(const int* array, int n, int x) {
    const int* p = array;
    for (int i = 0; i < n; i++) {
        if (*p == x) return p;
        p++;
    }
    return 0;
}
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Each step removes one hard-coded dependency, making the algorithm progressively more general.

1.  **Step 1 — Generalize element type:** Replace `int` with a type parameter `T`. Now the function works for any type that supports `operator==`. The size `n` is still an `int`. Result: `find1<T>(T* array, int n, const T& x)`.
2.  **Step 2 — Replace size with past-the-end pointer:** Instead of `int n`, accept a second pointer `T* beyond` pointing one past the last element. The loop condition becomes `p != beyond`. Result: `find2<T>(T* array, T* beyond, const T& x)`. The dependency on "knowing the size" is removed.
3.  **Step 3 — Replace null pointer return with `beyond`:** Return `beyond` (not `0`) to signal "not found." This removes the dependency on null pointers being meaningful — any iterator type can express "not found" this way. Result: `find3` returns `beyond` on failure.
4.  **Step 4 — Simplify and rename:** Merge the two loop conditions (`p != beyond && *p != x`), eliminating one `return`. Rename `array` to `first` because the algorithm no longer assumes an array. Result: `find4<T>(T* first, T* beyond, const T& x)`.
5.  **Step 5 — Remove pointers:** Replace `T*` with a new abstract type parameter `P`. The function no longer mentions pointers anywhere. `P` can be a raw pointer *or* any user-defined class. Result: `find5<T, P>(P first, P beyond, const T& x)`. Now `P` is the **iterator type** and `T` is the **value type**.

**Answer:** The five steps systematically remove the dependencies on: (1) element type, (2) array size, (3) pointer arithmetic for "not found", (4) redundant code, (5) raw pointers. The final `find5` works on any container that provides an iterator satisfying three requirements: `operator*`, `operator++`, `operator!=`.

</details>

##### **3.2. Iterator Requirements for `find5` and `reverse`** (Lecture 7, Example 2)

For the two algorithms below, list all the requirements imposed on the iterator type `P` and identify which STL iterator category each algorithm requires as a minimum.

**Algorithm A** (`find5`):
```cpp
template <typename T, typename P>
P find5(P first, P beyond, const T& x) {
    P p = first;
    while (p != beyond && *p != x)
        p++;
    return p;
}
```

**Algorithm B** (`reverse`):
```cpp
template <typename P, typename T>
void reverse(P start, P beyond) {
    while (start != beyond) {
        --beyond;
        if (start != beyond) {
            T t = *start;
            *start = *beyond;
            *beyond = t;
            ++start;
        }
    }
}
```

<details>
<summary>Click to see the solution</summary>

**Algorithm A — `find5` requirements:**

*   `operator!=` — to compare `p != beyond` and `*p != x`.
*   `operator*` — to dereference `*p` and read the value (read-only, returns a value).
*   `operator++` — to advance to the next element (`p++`).

**Minimum category:** **Input Iterator** (read-only, forward-only traversal).

**Algorithm B — `reverse` requirements:**

*   `operator!=` — to compare `start != beyond`.
*   `operator++` — to advance `start` forward (`++start`).
*   `operator--` — to move `beyond` backward (`--beyond`). **New requirement!**
*   `operator*` returning a **reference** — to assign through the iterator (`*start = *beyond`). **New requirement!**

**Minimum category:** **Bidirectional Iterator** (forward + backward movement, read-write access).

**Answer:**

*   `find5` requires an **Input Iterator**.
*   `reverse` requires a **Bidirectional Iterator** (adds `--` and write-through dereference to the requirements of `find5`).

</details>

##### **3.3. Write the `NODE` Iterator** (Lecture 7, Example 3)

Given the `NODE` class below, complete the nested `iterator` class so that `find5` can search a singly-linked list.

```cpp
class NODE {
    int value;
    NODE* next;
public:
    NODE(int v) : value(v), next(0) { }
    void add(NODE* n) { n->next = this->next; this->next = n; }
    bool operator!=(NODE& n) { return this->value != n.value; }
    operator int() { return value; }

    class iterator;
    friend class iterator;

    class iterator {
        NODE* p;
    public:
        // TODO: implement iterator(NODE*), operator*(), operator++(), operator!=()
    };
};
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** The iterator wraps a `NODE*` pointer and implements the three required operations in terms of the linked list's pointer structure.

```cpp
class iterator {
    NODE* p;   // current node (nullptr = "past the end")
public:
    // Constructor: point to the given node
    iterator(NODE* node) : p(node) { }

    // operator*: return the value of the current node
    int operator*() { return p->value; }

    // operator++: advance to the next node in the list
    void operator++() { p = p->next; }

    // operator!=: compare by pointer identity (address of current node)
    bool operator!=(const iterator& other) const {
        return p != other.p;
    }

    // Conversion to NODE* (used when passing to find5 as the "beyond" sentinel)
    operator NODE*() { return p; }
};
```

**Usage:**

```cpp
NODE* list = new NODE(1);
NODE* n7 = new NODE(7); list->add(n7);
NODE* n3 = new NODE(3); list->add(n3);

// Search for value 7 in the list
NODE::iterator result = find5<int>(
    NODE::iterator(list),  // first: head of the list
    NODE::iterator(0),     // beyond: past-the-end sentinel (nullptr)
    7
);
```

**Answer:** The key insight is that `operator++` follows the `next` pointer, and `nullptr` (passed as `NODE*(0)`) serves as the "past-the-end" sentinel — equivalent to `&A[100]` for an array.

</details>

##### **3.4. Identify Iterator Categories for STL Containers** (Lab 7, Exercise 1)

For each STL container listed below, state the category of iterator it provides and give one algorithm that can be used with it but *not* with a weaker iterator category.

**(a)** `std::vector<int>`
**(b)** `std::list<int>`
**(c)** `std::forward_list<int>`
**(d)** `std::istream_iterator<int>`

<details>
<summary>Click to see the solution</summary>

**(a) `std::vector<int>`:**

*   Iterator category: **Random Access Iterator**
*   Algorithm requiring this: `std::sort` (requires random access for efficient partitioning: `i + n`, `i[n]`, `i < j`).

**(b) `std::list<int>`:**

*   Iterator category: **Bidirectional Iterator**
*   Algorithm requiring this: `std::reverse` (requires `--i` to move backward).

**(c) `std::forward_list<int>`:**

*   Iterator category: **Forward Iterator**
*   Algorithm requiring this: `std::find` (requires only `++i` and `*i`); `std::sort` cannot be used directly on `forward_list`.

**(d) `std::istream_iterator<int>`:**

*   Iterator category: **Input Iterator**
*   Algorithm: `std::find` (read-only, single-pass). Cannot be used with `std::reverse` (needs bidirectional) or `std::sort` (needs random access).

**Answer:**

| Container | Iterator Category | Exclusive Algorithm |
|-----------|------------------|---------------------|
| `vector` | Random Access | `sort` |
| `list` | Bidirectional | `reverse` |
| `forward_list` | Forward | `find` |
| `istream_iterator` | Input | `find` (read-only pass) |

</details>

##### **3.5. Circular Buffer with Custom Forward Iterator** (Lab 7, Exercise 2)

Build a **circular buffer** as a template class with a custom forward iterator.

**Requirements:**

*   The class must support: `push(value)`, `pop()` (returns the value), `empty()`, `size()`.
*   Use `std::vector` for internal storage.
*   The iterator must iterate over the buffer elements in **circular fashion**: after reaching the last element, it wraps around and continues from the first.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** A circular buffer uses modular arithmetic (`%`) to wrap indices. The forward iterator tracks how many elements it has visited so that it knows when a full circle is complete.

```cpp
#include <iostream>
#include <vector>
#include <stdexcept>

template <typename T>
class CircularBuffer {
    std::vector<T> data_;   // underlying storage
    size_t head_;           // index of the oldest element (front)
    size_t tail_;           // index where the next push will go
    size_t count_;          // number of elements currently stored
    size_t capacity_;       // total capacity

public:
    CircularBuffer(size_t capacity)
        : data_(capacity), head_(0), tail_(0),
          count_(0), capacity_(capacity) { }

    // Add an element at the back
    void push(const T& value) {
        if (count_ == capacity_)
            throw std::overflow_error("Buffer is full");
        data_[tail_] = value;
        tail_ = (tail_ + 1) % capacity_;
        ++count_;
    }

    // Remove and return the front element
    T pop() {
        if (count_ == 0)
            throw std::underflow_error("Buffer is empty");
        T value = data_[head_];
        head_ = (head_ + 1) % capacity_;
        --count_;
        return value;
    }

    bool   empty()    const { return count_ == 0; }
    size_t size()     const { return count_; }
    size_t capacity() const { return capacity_; }

    // ---- Forward Iterator ----
    class iterator {
        const CircularBuffer* buf_;   // the buffer being iterated
        size_t index_;                // physical index in data_
        size_t visited_;              // how many elements we have visited

    public:
        iterator(const CircularBuffer* buf, size_t start, size_t visited)
            : buf_(buf), index_(start), visited_(visited) { }

        T operator*() const {
            return buf_->data_[index_];
        }

        iterator& operator++() {
            index_ = (index_ + 1) % buf_->capacity_;
            ++visited_;
            return *this;
        }

        bool operator!=(const iterator& other) const {
            return visited_ != other.visited_;
        }
    };

    // begin(): start at head, 0 elements visited
    iterator begin() const {
        return iterator(this, head_, 0);
    }

    // end(): same position, count_ elements visited (signals "done")
    iterator end() const {
        return iterator(this, head_, count_);
    }
};

int main() {
    CircularBuffer<int> buf(5);
    buf.push(10);
    buf.push(20);
    buf.push(30);
    buf.push(40);
    buf.push(50);

    std::cout << "Forward pass: ";
    for (auto it = buf.begin(); it != buf.end(); ++it)
        std::cout << *it << " ";   // 10 20 30 40 50
    std::cout << std::endl;

    buf.pop();  // removes 10
    buf.push(60);

    std::cout << "After pop + push: ";
    for (int val : buf)
        std::cout << val << " ";   // 20 30 40 50 60
    std::cout << std::endl;

    return 0;
}
```

**How the circular wrapping works:**

*   `push` stores at `tail_` and advances `tail_ = (tail_ + 1) % capacity_`.
*   `pop` reads from `head_` and advances `head_ = (head_ + 1) % capacity_`.
*   The iterator advances via `index_ = (index_ + 1) % capacity_` and compares using `visited_` count rather than a physical index, because the "end" and "begin" physical positions coincide in a full buffer.

**Answer:** The iterator counts visited elements (0 to `count_`) to determine termination, using modular arithmetic to wrap around the physical storage.

</details>

##### **3.6. SRP Violation — Identify and Fix** (Tutorial 7, Problem 1)

The Java code below violates the Single Responsibility Principle. Identify the violation and refactor the code to comply with SRP.

```java
// ssad.srp.problem.Document
public class Document {
    private String title;
    private int pages;

    public Document(String title, int pages) {
        this.title = title;
        this.pages = pages;
    }

    public void save() {
        System.out.println("The document is saved to .docx file");
    }

    public void load() {
        System.out.println("The document is loaded from .docx file");
    }
}
```

```java
// ssad.srp.problem.Main
public class Main {
    public static void main(String[] args) {
        Document doc = new Document("MyDoc", 10);
        doc.save();
        doc.load();
    }
}
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** `Document` mixes two concerns: (1) storing document data, and (2) performing I/O (save/load). SRP demands these be separate classes.

**Violation:** `Document` has at least two reasons to change — if the document's data model changes (title, pages), *and* if the file formats change (e.g., adding PDF support). These are distinct responsibilities.

**Refactored solution:**

```java
// Document: only stores data
public class Document {
    private String title;
    private int pages;

    public Document(String title, int pages) {
        this.title = title;
        this.pages = pages;
    }

    @Override
    public String toString() {
        return "Document{title='" + title + "', pages=" + pages + "}";
    }
}
```

```java
// SaveDocument: responsible only for saving
public class SaveDocument {
    public void saveToDocx(String path, Document doc) {
        System.out.println("The document is saved to " + path + ".docx file");
        System.out.println(doc);
    }

    public void saveToPdf(String path, Document doc) {
        System.out.println("The document is saved to " + path + ".pdf file");
        System.out.println(doc);
    }
}
```

```java
// LoadDocument: responsible only for loading
public class LoadDocument {
    public Document loadFromDocx(String path, Document doc) {
        System.out.println("The document is loaded from " + path + ".docx file");
        return new Document(path + ".docx", 10);
    }

    public Document loadFromPdf(String path, Document doc) {
        System.out.println("The document is loaded from " + path + ".pdf file");
        return new Document(path + ".pdf", 10);
    }
}
```

```java
// Main: uses the separated classes
public class Main {
    public static void main(String[] args) {
        Document doc = new Document("MyDoc", 10);

        SaveDocument docSaver = new SaveDocument();
        docSaver.saveToPdf("MyDoc", doc);

        System.out.println();

        LoadDocument docLoader = new LoadDocument();
        Document anotherDoc = docLoader.loadFromPdf("AnotherDoc", doc);
    }
}
```

**Answer:** Extract `save()` and `load()` into dedicated classes `SaveDocument` and `LoadDocument`. `Document` now has a single responsibility: storing document state. Each new file format only requires adding a method to `SaveDocument`/`LoadDocument`, not modifying `Document`.

</details>

##### **3.7. OCP Violation — Identify and Fix** (Tutorial 7, Problem 2)

The Java code below violates the Open/Closed Principle. Identify the violation and refactor it to comply with OCP.

```java
// ssad.ocp.problem.SaveDocument
// TODO: OCP is violated — adding new types will require changes here
public class SaveDocument {
    public void saveToDocx(String path, Document doc) {
        System.out.println("The document is saved to " + path + ".docx file");
        System.out.println(doc);
    }

    public void saveToPdf(String path, Document doc) {
        System.out.println("The document is saved to " + path + ".pdf file");
        System.out.println(doc);
    }
}
```

```java
// ssad.ocp.problem.LoadDocument
public class LoadDocument {
    public Document loadFromDocx(String path, Document doc) {
        System.out.println("The document is loaded from " + path + ".docx file");
        return new Document(path + ".docx", 10);
    }

    public Document loadFromPdf(String path, Document doc) {
        System.out.println("The document is loaded from " + path + ".pdf file");
        return new Document(path + ".pdf", 10);
    }
}
```

```java
// ssad.ocp.problem.Main
public class Main {
    public static void main(String[] args) {
        Document doc = new Document("MyDoc", 10);
        SaveDocument docSaver = new SaveDocument();
        docSaver.saveToPdf("MyDoc", doc);

        LoadDocument docLoader = new LoadDocument();
        Document anotherDoc = docLoader.loadFromPdf("AnotherDoc", doc);
    }
}
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Every time a new file format (e.g., `.txt`, `.xml`) is needed, `SaveDocument` and `LoadDocument` must be *modified*. According to OCP, they should instead be *extended* via new implementations of an abstract interface.

**Violation:** Adding a new save format requires opening `SaveDocument.java` and modifying it — this is a textbook OCP violation. Similarly for `LoadDocument`.

**Refactored solution — introduce `Saver` and `Loader` interfaces:**

```java
// Saver interface: closed for modification, open for extension
public interface Saver {
    void save(String path, Document doc);
}
```

```java
// Loader interface
public interface Loader {
    Document load(String path, Document doc);
}
```

```java
// Concrete saver for DOCX
public class SaveDocumentToDocx implements Saver {
    @Override
    public void save(String path, Document doc) {
        System.out.println("The document is saved to " + path + ".docx file");
        System.out.println(doc);
    }
}
```

```java
// Concrete saver for PDF
public class SaveDocumentToPdf implements Saver {
    @Override
    public void save(String path, Document doc) {
        System.out.println("The document is saved to " + path + ".pdf file");
        System.out.println(doc);
    }
}
```

```java
// Concrete loader from DOCX
public class LoadDocumentFromDocx implements Loader {
    public Document load(String path, Document doc) {
        System.out.println("The document is loaded from " + path + ".docx file");
        return new Document(path + ".docx", 10);
    }
}
```

```java
// Concrete loader from PDF
public class LoadDocumentFromPdf implements Loader {
    public Document load(String path, Document doc) {
        System.out.println("The document is loaded from " + path + ".pdf file");
        return new Document(path + ".pdf", 10);
    }
}
```

```java
// Main: depends on abstractions (Saver, Loader)
public class Main {
    public static void main(String[] args) {
        Document doc = new Document("MyDoc", 10);

        Saver docSaver = new SaveDocumentToDocx();
        docSaver.save("MyDoc", doc);

        System.out.println();

        Loader docLoader = new LoadDocumentFromDocx();
        Document anotherDoc = docLoader.load("AnotherDoc", doc);
    }
}
```

**Answer:** Introduce `Saver` and `Loader` interfaces. Each concrete format is a separate class implementing the interface. Adding `.txt` format now means creating `SaveDocumentToTxt implements Saver` — no existing class is modified.

</details>

##### **3.8. LSP Violation — Identify and Fix** (Tutorial 7, Problem 3)

The Java code below violates the Liskov Substitution Principle. Identify the violation and explain its impact.

```java
// ssad.lsp.problem.Document
public class Document {
    private String title;
    private int pages;

    public Document(String title, int pages) {
        this.title = title;
        this.pages = pages;
    }

    public void setData(String title) {
        this.title = title;  // sets only the title
    }

    public String getTitle() { return title; }
    public int getPages()    { return pages; }
    public void setPages(int pages) { this.pages = pages; }

    @Override
    public String toString() {
        return "Document{title='" + title + "', pages=" + pages + "}";
    }
}

// ssad.lsp.problem.SpecialDocument
// TODO: LSP is violated — setData() has changed behaviour
public class SpecialDocument extends Document {
    public SpecialDocument(String title, int pages) {
        super(title, pages);
    }

    @Override
    public void setData(String title) {
        super.setData(title);
        setPages(20);  // unexpected side effect!
    }
}

// ssad.lsp.problem.Main
public class Main {
    public static void main(String[] args) {
        Document doc1 = new Document("MyDoc", 10);
        Document doc2 = new SpecialDocument("MySpecialDoc", 10);

        doc2.setData("MyDoc");

        System.out.println("doc1: " + doc1);
        System.out.println("doc2: " + doc2);
    }
}
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** LSP requires that a subclass can be used wherever the superclass is used *without surprising the caller*. The override must preserve the observable contract of the overridden method.

**Violation:** `Document.setData(title)` has the well-defined contract: *"change the title, leave everything else unchanged."* `SpecialDocument.setData(title)` breaks this contract by *also* changing `pages` to 20. A caller using `Document doc2 = new SpecialDocument(...)` would call `doc2.setData("MyDoc")` expecting only the title to change, and would be surprised when `pages` becomes 20.

**Demonstration of the violation:**

```java
Document doc1 = new Document("MyDoc", 10);
Document doc2 = new SpecialDocument("MySpecialDoc", 10);

doc2.setData("MyDoc");

System.out.println("doc1: " + doc1);
// Output: Document{title='MyDoc', pages=10}   ← expected

System.out.println("doc2: " + doc2);
// Output: Document{title='MyDoc', pages=20}   ← SURPRISE! Pages changed!
```

**Fix — make `SpecialDocument` honour the parent's contract:**

```java
public class SpecialDocument extends Document {
    public SpecialDocument(String title, int pages) {
        super(title, pages);
    }

    // Additional behaviour is added as a NEW method, not by overriding setData
    public void method() {
        System.out.println("Special document method");
    }

    // If setData MUST do something extra, it should not change observable state
    // that callers of Document.setData don't expect.
    // Best practice: do NOT override setData if you cannot preserve its contract.
}
```

Alternatively, reconsider the hierarchy: if `SpecialDocument` *must* change `pages` on every title update, it may not be a proper subtype of `Document` at all, and composition should be used instead of inheritance.

**Answer:** The violation is in `SpecialDocument.setData` calling `setPages(20)` as an unexpected side effect. The fix is to either preserve the parent's contract in the override, or redesign the hierarchy so that `SpecialDocument` is not forced to inherit `setData` with a different meaning.

</details>
