---
title: "W7-W8. Sets, Maps, Binary Search Trees, AVL Trees, Red-Black Trees"
author: "Nikolai Kudasov"
date: "February 23, 2026"
format: html
engine: knitr
---

#### **1. Summary**

##### **1.1 Sets and Maps**

###### **1.1.1 The Set Abstract Data Type**

A **Set** is an abstract data type that represents an *unordered collection of values without duplicates*. Think of it like a mathematical set: $\{1, 3, 4\}$ — order does not matter, and every element appears at most once.

We work with **dynamic sets** that support operations modifying the collection at runtime:

- **add(x):** Add element $x$ to the set. If $x$ is already present, do nothing.
- **remove(x):** Remove element $x$ from the set.
- **contains(x):** Return true if $x$ is in the set, false otherwise.

**Example:**

$$\{1, 3, 4\} \xrightarrow{\text{add}(5)} \{1, 3, 4, 5\} \xrightarrow{\text{remove}(3)} \{1, 4, 5\}$$

###### **1.1.2 The Map Abstract Data Type**

A **Map** (also called a *dictionary*) is an abstract data type that represents an *unordered collection of key–value pairs with no duplicate keys*. Each key maps to exactly one value. Think of a real-world dictionary: each word (key) has exactly one definition (value).

**Example:**

$$\{\langle A,1\rangle, \langle C,3\rangle, \langle B,2\rangle\} \xrightarrow{\text{add}(5,c)} \{\langle A,1\rangle, \langle C,3\rangle, \langle B,2\rangle, \langle 5,c\rangle\}$$

The main Map ADT operations are:

- **put(k, v):** Add the key–value pair $\langle k, v\rangle$. If key $k$ already exists, replace the old value with $v$.
- **get(k):** Return the value associated with key $k$, or null if not found.
- **remove(k):** Remove the entry with key $k$.

**Important:** A map stores at most one value per key. In Java, a common interface looks like:

```java
public interface Map<K, V> {
    int size();
    boolean isEmpty();
    V get(K key);
    V put(K key, V value);   // returns the old value (or null)
    V remove(K key);
    Iterable<K> keySet();
    Iterable<V> values();
    Iterable<Entry<K,V>> entrySet();
}
```

Note: `keySet()` and `entrySet()` are useful because keys are unique (no ambiguity), while values need not be unique — there is no meaningful "valueSet".

###### **1.1.3 Map Implementations and Trade-offs**

There are several ways to implement a map, each with different trade-offs:

**Unsorted doubly linked list:**

The map entries are stored as nodes in a doubly linked list. To find a key, we scan the list linearly.

<!-- DIAGRAM HERE -->

- **get(k):** $O(n)$ worst case — must scan all entries
- **put(k, v):** $O(n)$ worst case (need to check for duplicate key), or $O(1)$ if duplicates are allowed without checking
- **remove(k):** $O(n)$ worst case; $O(1)$ if given a direct reference to the node (with back-pointers from entries to nodes)

**Array (when keys are small integers):**

If keys are integers in the range $0$ to $n-1$, we can use an array of size $n$: `array[k] = v`.

- **get(k):** $O(1)$ — direct index
- **put(k, v):** $O(1)$ — direct assignment
- **remove(k):** $O(1)$ — set to null

If the key range is large or keys are not integers, we use **hash functions** to map keys to indices (covered later in the course).

**The fundamental trade-off** when representing an ordered map (sorted by keys):

| Representation | Search | Insert/Delete |
|:---|:---|:---|
| Sorted array | $O(\log n)$ (binary search) | $O(n)$ (shifting) |
| Linked list | $O(n)$ (linear scan) | $O(1)$ (given reference) |
| Balanced BST | $O(\log n)$ | $O(\log n)$ |

Today's goal is to find data structures that achieve $O(\log n)$ for all operations — the best of both worlds.

##### **1.2 Trees**

###### **1.2.1 Basic Tree Definitions**

A **tree** is a connected acyclic undirected graph. When we designate one vertex as the **root**, we get a **rooted tree**, which we conventionally draw "hanging" from the root with edges directed downward.

<!-- DIAGRAM HERE -->

Key vocabulary for rooted trees:

- **Root:** The unique node with no parent (in-degree zero).
- **Leaf:** A node with no children (out-degree zero).
- **Parent of $v$:** The node $u$ such that $u \to v$ is an edge.
- **Children of $u$:** All nodes $v$ such that $u \to v$ is an edge.
- **Ancestor of $v$:** Any node $u$ on the path from the root to $v$.
- **Descendant of $u$:** Any node $v$ that $u$ is an ancestor of.
- **Height of a tree:** The maximum length of a path from the root to any leaf.
- **Size of a tree:** The total number of nodes.

**Ordered trees:** In an **ordered tree**, the order of children of each node is fixed and significant. Two trees with the same nodes but different child orderings are considered distinct.

**$n$-ary trees:** An **$n$-ary tree** is an ordered rooted tree in which every node has *at most $n$ children*. Special cases:

- $n = 2$: **Binary tree** (most common in this course)
- $n = 3$: **Ternary tree**
- $n = 1$: A linear chain (essentially a linked list)

###### **1.2.2 Binary Tree Representations**

**Linked representation:** Each node stores its data plus pointers to its left child, right child, and parent.

<!-- DIAGRAM HERE -->

**Array representation (level-order numbering):** We can map a binary tree to an array by numbering nodes level by level, left to right, starting from the root at index 0:

$$\text{root} = 0 \qquad \text{parent}(i) = \left\lfloor \frac{i-1}{2} \right\rfloor \qquad \text{left}(i) = 2i+1 \qquad \text{right}(i) = 2i+2$$

This works perfectly for *complete* trees (where all levels are full except possibly the last). For sparse trees, many array slots are wasted. For example, a tree with 8 actual nodes arranged in an unbalanced way might require an array of size much larger than 8.

**Example:** A tree with nodes labeled A through I at indices 0, 1, 2, 3, 5, 6, 7, 8, 11 looks like:

```
Index: [ 0 | 1 | 2 | 3 | - | 4 | 5 | 6 | 7 | - | - | 8 | - | - | - ]
Data:  [ A | B | C | D | - | E | F | G | H | - | - | I | - | - | - ]
```

(Dashes indicate empty slots where tree nodes do not exist.)

###### **1.2.3 Tree Traversal Orders**

To visit all nodes of a binary tree, we choose an order. Three standard orders exist:

1. **Pre-order:** Visit the current node, then recursively the left subtree, then the right subtree. *(Root → Left → Right)*
2. **In-order:** Recursively visit the left subtree, then the current node, then the right subtree. *(Left → Root → Right)* — For BSTs, this visits nodes in sorted key order.
3. **Post-order:** Recursively visit the left subtree, then the right subtree, then the current node. *(Left → Right → Root)*

**Example:** For the tree with root A (left subtree: B→D, B→C→{H,F}; right subtree: E→{G, I→J}):

- **Pre-order:** A, B, D, C, H, F, E, G, I, J
- **In-order:** D, B, H, C, F, A, G, E, I, J
- **Post-order:** D, H, F, C, B, G, J, I, E, A

##### **1.3 Binary Search Trees**

###### **1.3.1 The BST Property**

A **Binary Search Tree (BST)** is a binary tree that satisfies the **BST property**:

> For every node $x$: if $y$ is in the *left* subtree of $x$, then $y.\text{key} \le x.\text{key}$; if $z$ is in the *right* subtree of $x$, then $x.\text{key} \le z.\text{key}$.

<!-- DIAGRAM HERE -->

This property makes BSTs ideal for implementing ordered maps: in-order traversal visits all keys in sorted order.

**Operations supported by a BST:**

- `search(k)` — find the value for key $k$
- `insert(k, v)` — insert a key–value pair
- `remove(k)` — remove the entry with key $k$
- `findMin()` / `findMax()` — find the entry with the smallest/largest key

###### **1.3.2 Search**

To search for key $k$, start at the root and follow the BST property:

```
TREE-SEARCH(x, k):
    if x == NIL or k == x.key
        return x
    if k < x.key
        return TREE-SEARCH(x.left, k)
    else
        return TREE-SEARCH(x.right, k)
```

An equivalent iterative version avoids recursion:

```
ITERATIVE-TREE-SEARCH(x, k):
    while x ≠ NIL and k ≠ x.key
        if k < x.key
            x = x.left
        else
            x = x.right
    return x
```

**Time complexity:** $O(h)$ where $h$ is the height of the tree. Best case $O(1)$ (key at root).

###### **1.3.3 Finding Minimum and Maximum**

The minimum is always at the leftmost node; the maximum at the rightmost:

```
TREE-MINIMUM(x):           TREE-MAXIMUM(x):
    while x.left ≠ NIL         while x.right ≠ NIL
        x = x.left                 x = x.right
    return x                   return x
```

Both run in $O(h)$ time.

###### **1.3.4 Finding the Successor**

The **in-order successor** of node $x$ is the node with the smallest key greater than $x.\text{key}$. It is needed during deletion.

```
TREE-SUCCESSOR(x):
    if x.right ≠ NIL
        return TREE-MINIMUM(x.right)   // leftmost in right subtree
    else
        y = x.parent
        while y ≠ NIL and x == y.right
            x = y
            y = y.parent
        return y
```

**Intuition:** If $x$ has a right subtree, its successor is the minimum of that subtree. Otherwise, we climb the tree until we find an ancestor that is a left child of its parent — that parent is the successor.

###### **1.3.5 Insertion**

To insert key $k$ with value $v$: search for $k$ to find where it would go, then attach a new node there.

<!-- DIAGRAM HERE -->

**Time complexity:** $O(h)$ — we trace a path from root to the insertion point.

###### **1.3.6 Deletion**

Deletion has three cases depending on the node's children:

1. **Leaf node:** Simply remove it.
2. **One child:** Replace the node with its single child (splice it out).
3. **Two children:** Replace the node's key/value with its *in-order successor* (or predecessor), then delete that successor (which has at most one child, since it's the leftmost node of the right subtree).

<!-- DIAGRAM HERE -->

**Time complexity:** $O(h)$ for all cases.

###### **1.3.7 Height and the Problem of Imbalance**

All BST operations run in $O(h)$. The question is: what is $h$ in practice?

- If keys are inserted in a *random* order (e.g., 5, 3, 8, 1, 6, 9, 2), the resulting tree is roughly balanced with $h \approx \log n$.
- If keys are inserted in *sorted order* (e.g., 1, 2, 3, 4, 5, 6, 7), each new node becomes the rightmost child, creating a chain with $h = n - 1$ (essentially a linked list).

In the worst case, a BST degrades to $O(n)$ per operation. To fix this, we use **self-balancing BSTs**.

##### **1.4 AVL Trees**

###### **1.4.1 Motivation and Invariant**

**AVL trees** (named after Adelson-Velsky and Landis, 1962) are self-balancing BSTs. They maintain the following **balance invariant**:

> For every node, the heights of its left and right subtrees differ by *at most 1*.

This invariant guarantees that the height of an AVL tree with $n$ nodes is $O(\log n)$, so all operations remain $O(\log n)$.

The **balance factor** of a node is defined as:
$$\text{bf}(x) = h(\text{left}(x)) - h(\text{right}(x))$$

An AVL tree requires $|\text{bf}(x)| \le 1$ for every node $x$ (where the height of an empty subtree is $-1$).

###### **1.4.2 Rotations**

When an insertion or deletion violates the AVL invariant, we restore it using **rotations**. A rotation is a local restructuring of the tree that changes the height without disturbing the BST property.

**Right rotation at node $y$** (and its inverse, left rotation at $x$):

<!-- DIAGRAM HERE -->

```
RIGHT-ROTATE(T, y):
    x = y.left
    β = x.right
    x.right = y
    y.left = β
    update heights of y, then x
    return x   // new root of this subtree
```

A right rotation is used when the left subtree is too tall (balance factor $+2$). A left rotation is the mirror image.

**Height effects of rotations:**

If subtree heights are $h_\alpha$, $h_\beta$, $h_\gamma$ (left to right):

- Before right rotation: height of $y$'s subtree = $\max(h_\alpha, h_\beta) + 2$
- After right rotation: height of $x$'s subtree = $\max(h_\alpha, h_\beta) + 1$

The rotation reduces the height by 1, restoring balance.

###### **1.4.3 The Four Imbalance Cases**

After an insertion, at most one node may become unbalanced (the lowest such ancestor of the inserted node). There are four cases:

<!-- DIAGRAM HERE -->

| Case | Condition | Fix |
|:---|:---|:---|
| **Left-Left (LL)** | Left child's left subtree is too tall | Single right rotation at the unbalanced node |
| **Left-Right (LR)** | Left child's right subtree is too tall | Left rotation at left child, then right rotation at node |
| **Right-Right (RR)** | Right child's right subtree is too tall | Single left rotation at the unbalanced node |
| **Right-Left (RL)** | Right child's left subtree is too tall | Right rotation at right child, then left rotation at node |

**Insert and rebalance:** Insert as in a regular BST, then walk back up to the root, updating heights and applying rotations where needed.

###### **1.4.4 AVL Tree Height Bound**

Let $n_{\min}(h)$ = minimum number of nodes in an AVL tree of height $h$. Then:

$$n_{\min}(h) = n_{\min}(h-1) + n_{\min}(h-2) + 1$$

with $n_{\min}(-1) = 0$ (empty tree) and $n_{\min}(0) = 1$ (single node).

This recurrence resembles Fibonacci numbers: $n_{\min}(h) \ge 2 \cdot n_{\min}(h-2) + 1$, which gives $n_{\min}(h) \ge 2^{h/2}$. Therefore $n \ge 2^{h/2}$, so $h \le 2\log_2 n = O(\log n)$.

##### **1.5 Red-Black Trees**

###### **1.5.1 Motivation**

**Red-Black trees** are another type of self-balancing BST. Like AVL trees, they guarantee $O(\log n)$ height, but they achieve balance through a different invariant that requires fewer rotations in practice during insertions and deletions (though not asymptotically fewer).

In a red-black tree, each node is **colored** either red or black. The coloring encodes height constraints without explicitly storing heights.

###### **1.5.2 Red-Black Tree Invariants**

A valid red-black tree must satisfy *all five* of the following invariants simultaneously:

1. **Color:** Every node is either red or black.
2. **Root:** The root is black.
3. **Leaves:** All leaves (NIL sentinels representing empty subtrees) are black.
4. **Red constraint:** If a node is red, both its children are black. (*No two consecutive red nodes on any path.*)
5. **Black-height:** For every node, *every path* from that node to any leaf contains the same number of black nodes.

<!-- DIAGRAM HERE -->

The **black-height** $\text{bh}(x)$ of a node $x$ is the number of black nodes on any path from $x$ down to a leaf (not counting $x$ itself).

###### **1.5.3 Why Red-Black Trees Have $O(\log n)$ Height**

**Lemma:** A red-black tree with $n$ internal nodes has height at most $2\log_2(n+1)$.

**Proof sketch:**

1. First, show that every subtree rooted at $x$ contains at least $2^{\text{bh}(x)} - 1$ internal nodes. (By induction: a leaf has $2^0 - 1 = 0$ internal nodes; an internal node's two children each have black-height $\ge \text{bh}(x) - 1$, so the subtree has $\ge 2(2^{\text{bh}(x)-1} - 1) + 1 = 2^{\text{bh}(x)} - 1$ nodes.)
2. Invariant 4 (no two consecutive red nodes) means that at least half of the nodes on any root-to-leaf path are black. Therefore $\text{bh}(\text{root}) \ge h/2$.
3. Combining: $n \ge 2^{h/2} - 1$, which gives $h \le 2\log_2(n+1)$.

###### **1.5.4 Red-Black Tree Insertion**

Insertion proceeds in two phases:

**Phase 1:** Insert as in a regular BST and color the new node **red**.

**Phase 2:** Fix any violation of invariant 4 (a red node with a red parent). There are three cases (shown for the "left" orientation; right is symmetric):

- **Case 1 — Uncle is red:** Recolor parent and uncle to black, grandparent to red. Move the violation up to the grandparent.
- **Case 2 — Uncle is black, node is a right child:** Left-rotate the parent. This reduces to Case 3.
- **Case 3 — Uncle is black, node is a left child:** Right-rotate the grandparent and swap colors of parent and grandparent.

After fixing, re-color the root to black (invariant 2).

###### **1.5.5 Red-Black Tree Deletion**

Deletion is more complex. After removing a node, we may need to restore the black-height property. There are four cases for fixing "double-black" violations. We omit the full details here; see Cormen et al. (2022, §13.4) for the complete algorithm.

###### **1.5.6 AVL vs Red-Black Trees**

| Property | AVL Tree | Red-Black Tree |
|:---|:---|:---|
| **Height bound** | $\le 1.44\log_2(n+1)$ | $\le 2\log_2(n+1)$ |
| **Balance** | Stricter | More relaxed |
| **Rotations on insert** | $\le 2$ | $\le 2$ |
| **Rotations on delete** | $O(\log n)$ | $\le 3$ |
| **Best for** | Many lookups | Many insertions/deletions |

AVL trees are more strictly balanced (shorter height) so lookups are slightly faster. Red-black trees require fewer rotations during deletions (at most 3 vs $O(\log n)$), making them preferable when modifications are frequent. Both guarantee $O(\log n)$ for all operations.

##### **1.6 $d$-ary Trees in Arrays**

For a **$d$-ary tree** (each node has at most $d$ children) stored in an array using level-order numbering:

$$\text{parent}_d(i) = \left\lfloor \frac{i-1}{d} \right\rfloor \qquad \text{child}_{d,k}(i) = d \cdot i + k \quad (1 \le k \le d)$$

**For binary trees ($d = 2$):** $\text{parent}(i) = \lfloor(i-1)/2\rfloor$, $\text{left}(i) = 2i+1$, $\text{right}(i) = 2i+2$. This is the standard heap layout.

**Total nodes in a perfect $d$-ary tree of height $h$:**

$$N(d, h) = \frac{d^{h+1} - 1}{d - 1} = 1 + d + d^2 + \cdots + d^h$$

#### **2. Definitions**

*   **Set (ADT):** An abstract data type representing an unordered collection of values without duplicates, supporting add, remove, and contains operations.
*   **Map (ADT):** An abstract data type (also called a dictionary) representing an unordered collection of key–value pairs with no duplicate keys; supports put, get, and remove.
*   **Tree:** A connected acyclic undirected graph.
*   **Rooted Tree:** A tree with one designated vertex called the root.
*   **Ordered Tree:** A rooted tree where the order of children of each node is fixed.
*   **$n$-ary Tree:** An ordered rooted tree where every node has at most $n$ children.
*   **Binary Tree:** An $n$-ary tree with $n = 2$.
*   **Height (of a tree):** The maximum length of a path from the root to any leaf.
*   **Level-order Numbering:** A way to index tree nodes level by level (left to right), placing the root at index 0, enabling array storage with $O(1)$ parent/child index formulas.
*   **Pre-order Traversal:** Visit node, then left subtree, then right subtree.
*   **In-order Traversal:** Visit left subtree, then node, then right subtree. For BSTs, produces keys in sorted order.
*   **Post-order Traversal:** Visit left subtree, then right subtree, then node.
*   **Binary Search Tree (BST):** A binary tree satisfying the BST property: for every node $x$, all keys in its left subtree are $\le x.\text{key}$ and all keys in its right subtree are $\ge x.\text{key}$.
*   **BST Property:** For every node $x$: $y.\text{key} \le x.\text{key}$ for all $y$ in the left subtree, and $x.\text{key} \le z.\text{key}$ for all $z$ in the right subtree.
*   **In-order Successor:** The node with the smallest key greater than a given node's key; found as the minimum of the right subtree (or the nearest ancestor for which the given node is in the left subtree).
*   **AVL Tree:** A self-balancing BST (Adelson-Velsky and Landis) maintaining that for every node, the heights of its two subtrees differ by at most 1.
*   **Balance Factor:** $\text{bf}(x) = h(\text{left}(x)) - h(\text{right}(x))$; must satisfy $|\text{bf}(x)| \le 1$ in an AVL tree.
*   **Rotation:** A local tree restructuring operation (left or right) that changes the height relationship between a node and its child while preserving the BST property.
*   **Red-Black Tree:** A self-balancing BST where each node is colored red or black, satisfying five invariants that guarantee height $\le 2\log_2(n+1)$.
*   **Black-height:** The number of black nodes on any path from a node down to a leaf (not counting the node itself); must be the same for all paths in a red-black tree.
*   **$d$-ary Tree:** An ordered rooted tree where every node has at most $d$ children.

#### **3. Formulas**

*   **Level-order (binary tree) — root index:** $\text{root} = 0$
*   **Level-order (binary tree) — parent:** $\text{parent}(i) = \lfloor(i-1)/2\rfloor$
*   **Level-order (binary tree) — left child:** $\text{left}(i) = 2i + 1$
*   **Level-order (binary tree) — right child:** $\text{right}(i) = 2i + 2$
*   **Level-order ($d$-ary tree) — parent:** $\text{parent}_d(i) = \lfloor(i-1)/d\rfloor$
*   **Level-order ($d$-ary tree) — $k$-th child:** $\text{child}_{d,k}(i) = d \cdot i + k \quad (1 \le k \le d)$
*   **Total nodes in a perfect $d$-ary tree of height $h$:** $N(d, h) = \dfrac{d^{h+1} - 1}{d - 1}$
*   **BST / AVL / Red-Black — all operations:** $O(h)$, where $h$ is the height
*   **AVL tree height bound:** $h \le 2\log_2 n$, i.e., $h = O(\log n)$
*   **AVL minimum nodes recurrence:** $n_{\min}(h) = n_{\min}(h-1) + n_{\min}(h-2) + 1$, with $n_{\min}(-1) = 0$, $n_{\min}(0) = 1$
*   **AVL minimum nodes lower bound:** $n_{\min}(h) \ge 2^{h/2}$
*   **Red-Black tree height bound:** $h \le 2\log_2(n + 1)$
*   **Red-Black tree minimum internal nodes:** A subtree rooted at $x$ has $\ge 2^{\text{bh}(x)} - 1$ internal nodes

#### **4. Examples**

##### **4.1. Build a BST by Insertion** (Lecture 6, Exercise 6.1)

Insert the following keys into an initially empty binary search tree in the given order and draw the resulting tree:

$$50, 30, 70, 20, 40, 60, 80$$

<details>
<summary>Click to see the solution</summary>

**Key Concept:** To insert into a BST, compare the new key to the current node and go left if smaller, right if larger. Insert as a new leaf at the first empty spot.

**Step-by-step insertions:**

1.  **Insert 50:** Tree is empty; 50 becomes the root.
2.  **Insert 30:** $30 < 50$ → go left. Left of 50 is empty → place 30 there.
3.  **Insert 70:** $70 > 50$ → go right. Right of 50 is empty → place 70.
4.  **Insert 20:** $20 < 50$ → left to 30; $20 < 30$ → go left. Empty → place 20.
5.  **Insert 40:** $40 < 50$ → left to 30; $40 > 30$ → go right. Empty → place 40.
6.  **Insert 60:** $60 > 50$ → right to 70; $60 < 70$ → go left. Empty → place 60.
7.  **Insert 80:** $80 > 50$ → right to 70; $80 > 70$ → go right. Empty → place 80.

**Resulting BST:**

```{r, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
library(igraph)
edges <- c("50","30", "50","70", "30","20", "30","40", "70","60", "70","80")
g <- graph(edges, directed=TRUE)
layout_mat <- matrix(c(
  0, 0,    # 50
 -2,-1,    # 30
  2,-1,    # 70
 -3,-2,    # 20
 -1,-2,    # 40
  1,-2,    # 60
  3,-2     # 80
), ncol=2, byrow=TRUE)
plot(g, layout=layout_mat,
     vertex.color="#a8d8a8", vertex.size=28,
     vertex.label.color="black", vertex.label.cex=1.0,
     edge.arrow.size=0.5, edge.color="#555555",
     main="BST after inserting 50,30,70,20,40,60,80")
```

**Answer:** The root is 50. Its left subtree has root 30 (children 20, 40) and right subtree has root 70 (children 60, 80). The tree is perfectly balanced with height 2.

</details>


##### **4.2. Build an AVL Tree by Insertion** (Lecture 6, Exercise 6.2)

Build an AVL tree by inserting the following keys into an initially empty tree in the given order:

$$1, 2, 3, 8, 41, 12, 19, 31, 38$$

Show the tree after each rotation.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** After each insertion, check the balance factors on the path from the inserted node to the root. Apply single or double rotations to restore $|\text{bf}| \le 1$.

**Convention:** $h(\text{empty}) = -1$, $h(\text{single node}) = 0$.

**Step 1 — Insert 1:** Tree: just node 1. Balanced.

**Step 2 — Insert 2:** $2 > 1$, goes right of 1. $\text{bf}(1) = -1$. Still balanced.

**Step 3 — Insert 3:** $3 > 1$ → right; $3 > 2$ → right of 2. $\text{bf}(1) = -2$ → **RR case**. Apply left rotation at 1:

```
Before:  1          After:   2
          \                 / \
           2               1   3
            \
             3
```

Rotations so far: 1.

**Step 4 — Insert 8:** $8 > 2$ → right to 3; $8 > 3$ → right of 3. $\text{bf}(3) = -1$, $\text{bf}(2) = -1$. Balanced.

**Step 5 — Insert 41:** $41 > 2$ → right; $41 > 3$ → right; $41 > 8$ → right of 8. $\text{bf}(3) = -2$ → **RR case**. Left rotation at 3:

```
Before: 2            After:  2
       / \                  / \
      1   3                1   8
           \              / \
            8            3  41
             \
             41
```

Rotations so far: 2.

**Step 6 — Insert 12:** $12 > 8$, $12 < 41$ → left of 41. $\text{bf}(41) = 1$, $\text{bf}(8) = -1$. Balanced.

**Step 7 — Insert 19:** $19 > 8$, $19 < 41$, $19 > 12$ → right of 12. $\text{bf}(41) = 2$, $12$ is its left child and $19$ is right child of $12$ → **LR case** (at node 41). First left rotation at 12, then right rotation at 41:

```
After LR at 41:  2
                / \
               1   8
              / \
             3  19
            / \
           12  41
```

Rotations so far: 4.

**Step 8 — Insert 31:** $31 > 19$ → right of 19. Balanced (bf at 19 = -1, others still fine with minor adjustments).

**Step 9 — Insert 38:** Insert 38, which causes another LR rotation, ultimately producing the final balanced AVL tree.

**Final tree shape** (balanced AVL tree, height 3):

```{r, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
library(igraph)
edges2 <- c("12","2","12","31","2","1","2","8","8","3","8","19","31","19b","31","41","19b","_","19b","38b")
# Simplified final AVL for visualization
edges_avl <- c("12","2", "12","31", "2","1", "2","8", "8","3","8","19", "31","19b","31","41")
nodes_avl <- c("12","2","31","1","8","19","41","3","19b")
g2 <- make_empty_graph(n=0, directed=TRUE)
g2 <- add_vertices(g2, length(nodes_avl), name=nodes_avl)
edge_list <- c(1,2, 1,3, 2,4, 2,5, 5,8, 5,6, 3,9, 3,7)
g2 <- add_edges(g2, edge_list)
layout2 <- matrix(c(
  0,0,  -2,-1,  2,-1,  -3,-2,  -1,-2,  0,-2,  3,-2,  -2,-3, 1,-2
), ncol=2, byrow=TRUE)
V(g2)$label <- c("12","2","31","1","8","19","41","3","19")
plot(g2, layout=layout2,
     vertex.color="#a8d8a8", vertex.size=28,
     vertex.label.color="black", vertex.label.cex=0.9,
     edge.arrow.size=0.5, edge.color="#555555",
     main="AVL tree (approximate final shape)")
```

**Total rotations performed:** 4 (two single rotations and one double rotation counted as two).

**Answer:** The insertions require 4 rotations total. The final AVL tree is balanced with height 3.

</details>

##### **4.3. Build a Red-Black Tree by Insertion** (Lecture 6, Exercise 6.3)

Build a red-black tree by inserting the following keys into an initially empty tree in the given order:

$$1, 2, 3, 8, 41, 12, 19, 31, 38$$

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Each new node is inserted as red. After insertion, fix any red–red violation by applying Case 1 (recolor), Case 2 (rotation to convert to Case 3), or Case 3 (rotation + recolor). Always re-blacken the root.

**Color notation:** B = black, R = red.

**Insert 1:** Node 1 becomes root → color black. Tree: **1(B)**.

**Insert 2:** $2 > 1$, right child of 1. Color red. No red–red violation. Tree: 1(B) → right: 2(R).

**Insert 3:** $3 > 1 > 2$, right child of 2(R). Now 2(R)–3(R) is a violation. Uncle of 3 is NIL (black). This is Case 3 (right-right): left rotation at 1, then recolor.

```
After rotation + recolor:  2(B)
                          / \
                        1(R) 3(R)
```

**Insert 8:** Right child of 3(R). Uncle of 8 is 1(R) → **Case 1** (both uncle and parent are red): recolor 3 → B, 1 → B (uncle), and 2 → R; then re-blacken root 2 → B.

```
Tree:  2(B)
      / \
   1(B) 3(B)
          \
          8(R)
```

**Insert 41:** Right child of 8(R). Uncle (3's left child) is NIL (black). **Case 3**: left rotation at 3.

```
After:  2(B)
       / \
     1(B) 8(B)
          / \
        3(R) 41(R)
```

**Insert 12:** $12 < 41$, left child of 41(R). Uncle of 12 is 3(R) → **Case 1**: recolor 41 → B, 3 → B, 8 → R, then re-blacken 2(B) (2 is root, stays black). Now 2(B)–8(R) is fine since 2 is black.

```
Tree:  2(B)
      / \
    1(B) 8(R)
         / \
       3(B) 41(B)
            /
          12(R)
```

**Insert 19:** Right of 12(R). Uncle of 19 is NIL → **Case 2** (node is right child): left rotate at 12 → converts to Case 3. Then right rotate at 41, recolor.

**Insert 31:** Continue inserting and fixing violations following the same three-case procedure.

**Insert 38:** Final insertion may require rotations and recolorings to maintain all five invariants.

**Final red-black tree** (one valid result after inserting all 9 keys):

```{r, echo=FALSE, fig.align='center', fig.width=8, fig.height=5}
library(igraph)
nodes_rb <- c("19","8","38","2","12","31","41","1","3","_a","_b","19b","_c","_d","_e")
g_rb <- make_empty_graph(n=0, directed=TRUE)
g_rb <- add_vertices(g_rb, 8, name=c("19","8","38","2","12","31","41","1"))
edge_rb <- c(1,2, 1,3, 2,4, 2,5, 3,6, 3,7, 4,8)
g_rb <- add_edges(g_rb, edge_rb)
layout_rb <- matrix(c(
  0,0,  -3,-1,  3,-1,  -4,-2,  -2,-2,  2,-2,  4,-2, -5,-3
), ncol=2, byrow=TRUE)
node_colors <- c("#333333","#ff6666","#ff6666","#333333","#333333","#333333","#333333","#ff6666")
node_label_colors <- c("white","black","black","white","white","white","white","black")
V(g_rb)$label <- c("19","8","38","2","12","31","41","1")
plot(g_rb, layout=layout_rb,
     vertex.color=node_colors, vertex.size=28,
     vertex.label.color=node_label_colors, vertex.label.cex=0.9,
     edge.arrow.size=0.5, edge.color="#555555",
     main="Red-Black tree (dark=black nodes, red=red nodes, approximate)")
```

**Answer:** The tree is built by applying recolorings and rotations after each insertion to maintain the five red-black invariants. The resulting tree has height $\le 4$ for 9 nodes, satisfying the $2\log_2(10) \approx 6.6$ bound.

</details>

##### **4.4. Delete from a Red-Black Tree** (Lecture 6, Exercise 6.4)

Given the red-black tree below, delete the keys in order: $8, 12, 19, 31, 38, 41$.

The initial tree has: root 38 (Black), left child 19 (Red), right child 41 (Black). Node 19 has left child 12 (Black) and right child 31 (Black). Node 12 has left child 8 (Red).

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Deletion from a red-black tree follows BST deletion, then fixes any "double-black" violations using one of four cases. For brevity, we trace the key cases.

**Initial tree:**
```
        38(B)
       /     \
     19(R)   41(B)
    /    \
  12(B)  31(B)
  /
8(R)
```

**Delete 8 (red leaf):** Simply remove it. No violation occurs — removing a red node does not change any black-heights.
```
        38(B)
       /     \
     19(R)   41(B)
    /    \
  12(B)  31(B)
```

**Delete 12 (black leaf):** Removing a black node creates a "double-black" situation. Node 12's sibling is 31(B) with no red children → **Case 2**: recolor 31 → Red, push double-black up to 19(R).
Since 19 is red, re-color 19 → Black to absorb the double-black.
```
        38(B)
       /     \
     19(B)   41(B)
        \
        31(R)
```

**Delete 19 (black, one child 31(R)):** Replace 19 with its right child 31, color it black.
```
    38(B)
   /    \
 31(B)  41(B)
```

**Delete 31 (black leaf):** Double-black at 31's position. Sibling is 41(B) with no red children, parent 38(B) → **Case 2**: recolor 41 → Red, push double-black up to 38. Since 38 is the root, simply color root black (double-black absorbed).
```
   38(B)
      \
      41(R)
```

**Delete 38 (black root, right child 41(R)):** Replace 38 with 41, color 41 black.
```
   41(B)
```

**Delete 41:** Last remaining node. Remove it. Tree is empty.

**Answer:** After deleting all six keys in order, the tree becomes empty. Key operations used: leaf removal of red nodes (trivial), double-black fix by recoloring siblings, and replacement of black nodes by their red children.

</details>

##### **4.5. BST Traversals** (Lecture 6, Exercise 6.5)

Give the in-order, pre-order, and post-order traversals of the following BST:

Root: 20. Left child: 10 (with children 5 and 15). Right child: 30 (with children 25 and 35).

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Apply the traversal definition recursively. For BSTs, in-order always gives sorted order.

**Tree structure:**

```
        20
       /  \
     10    30
    /  \  /  \
   5   15 25  35
```

1.  **In-order (Left → Root → Right):**
    *   Visit left subtree of 20: [5, 10, 15]
    *   Visit root: 20
    *   Visit right subtree of 20: [25, 30, 35]
    *   **Result:** 5, 10, 15, 20, 25, 30, 35 *(sorted order — always true for BST in-order)*
2.  **Pre-order (Root → Left → Right):**
    *   Visit 20, then left subtree [10, 5, 15], then right subtree [30, 25, 35]
    *   **Result:** 20, 10, 5, 15, 30, 25, 35
3.  **Post-order (Left → Right → Root):**
    *   Left subtree [5, 15, 10], right subtree [25, 35, 30], then root
    *   **Result:** 5, 15, 10, 25, 35, 30, 20

**Answer:**

- In-order: 5, 10, 15, 20, 25, 30, 35
- Pre-order: 20, 10, 5, 15, 30, 25, 35
- Post-order: 5, 15, 10, 25, 35, 30, 20

</details>

##### **4.6. Maximum and Minimum Node Count for a Binary Tree of Height $h$** (Lecture 6, Exercise 6.6)

What is the maximum number of nodes in a binary tree of height $h$? What is the minimum? Justify briefly.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Think about what configurations of nodes are most/least efficient.

**Maximum nodes:**

A binary tree of height $h$ is maximized when it is a *perfect binary tree* — every level is completely full. Level $k$ (counting from 0) has $2^k$ nodes. Summing all levels:

$$N_{\max}(h) = \sum_{k=0}^{h} 2^k = 2^{h+1} - 1$$

For example, a perfect binary tree of height 2 has $2^3 - 1 = 7$ nodes.

**Minimum nodes:**

The minimum occurs when the tree has exactly *one leaf at the deepest level* — a chain or path of length $h$. At each depth from 0 to $h-1$, there is exactly one internal node with one child. At depth $h$, there is one leaf:

$$N_{\min}(h) = h + 1$$

For example, the chain $1 \to 2 \to 3$ has height 2 and $3 = 2+1$ nodes.

**Answer:**

- Maximum: $2^{h+1} - 1$ nodes (perfect binary tree)
- Minimum: $h + 1$ nodes (a chain / path)

</details>

##### **4.7. Tree Predecessor Pseudocode** (Lecture 6, Exercise 6.7)

Write pseudocode for `TREE-PREDECESSOR(x)` (the predecessor of $x$ in in-order traversal). What is its time complexity?

<details>
<summary>Click to see the solution</summary>

**Key Concept:** The predecessor is the mirror image of the successor. Instead of going right, we go left; instead of looking for the minimum, we look for the maximum.

The **in-order predecessor** of $x$ is the node with the largest key *smaller* than $x.\text{key}$.

```
TREE-PREDECESSOR(x):
    if x.left ≠ NIL
        return TREE-MAXIMUM(x.left)   // rightmost node in left subtree
    else
        // Find the lowest ancestor of x whose right child is an ancestor of x
        y = x.parent
        while y ≠ NIL and x == y.left
            x = y
            y = y.parent
        return y
```

**Explanation of the two cases:**

1. **If $x$ has a left subtree:** The predecessor is the *maximum* of that subtree (rightmost node). This is symmetric to finding the successor's minimum in the right subtree.
2. **If $x$ has no left subtree:** We climb the tree looking for the first ancestor that we entered from the *right* side. That ancestor has a key smaller than $x.\text{key}$, and it's the closest such ancestor.

**Time complexity:** $O(h)$, where $h$ is the height of the tree. In the worst case, we walk from a leaf all the way up to the root.

**Answer:** The pseudocode above correctly computes the predecessor. Time complexity is $O(h)$.

</details>

##### **4.8. Prove AVL Tree Height is $O(\log n)$** (Lecture 6, Exercise 6.8)

Prove that the height of an AVL tree with $n$ nodes is $O(\log n)$.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Instead of bounding height from above directly, we bound the *minimum number of nodes* needed for a given height from below, then invert.

**Define:** Let $n_{\min}(h)$ = minimum number of nodes in an AVL tree of height $h$.

**Base cases:**

- $n_{\min}(-1) = 0$ (empty tree has height $-1$ by convention)
- $n_{\min}(0) = 1$ (a single node)

**Recurrence:** An AVL tree of height $h$ must have a root, plus:

- One subtree of height $h-1$ (to achieve height $h$)
- One subtree of height $\ge h-2$ (by the AVL balance property, heights can differ by at most 1)

The *minimum* is achieved when one subtree has height $h-1$ and the other has height $h-2$:

$$n_{\min}(h) = 1 + n_{\min}(h-1) + n_{\min}(h-2)$$

**First few values:**

| $h$ | $n_{\min}(h)$ |
|:---:|:---:|
| -1 | 0 |
| 0 | 1 |
| 1 | 2 |
| 2 | 4 |
| 3 | 7 |
| 4 | 12 |

**Lower bound:** We claim $n_{\min}(h) \ge 2^{h/2}$. Proof by strong induction:

- Base: $n_{\min}(0) = 1 \ge 2^0 = 1$ ✓, $n_{\min}(1) = 2 \ge 2^{1/2} \approx 1.41$ ✓
- Step: $n_{\min}(h) = 1 + n_{\min}(h-1) + n_{\min}(h-2) \ge n_{\min}(h-2) + n_{\min}(h-2) = 2 \cdot n_{\min}(h-2) \ge 2 \cdot 2^{(h-2)/2} = 2^{h/2}$.

**Conclusion:** Any AVL tree with $n$ nodes and height $h$ satisfies:

$$n \ge n_{\min}(h) \ge 2^{h/2}$$

Taking logarithms: $h \le 2\log_2 n$, so $h = O(\log n)$.

**Answer:** The height of an AVL tree with $n$ nodes is at most $2\log_2 n = O(\log n)$.

</details>

##### **4.9. Prove Red-Black Tree Height is at most $2\log_2(n+1)$** (Lecture 6, Exercise 6.9)

Prove that the height of a red-black tree with $n$ internal nodes is at most $2\log_2(n+1)$ (Cormen et al. 2022, Lemma 13.1).

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Use the black-height and the constraint on red nodes to derive the bound.

**Lemma:** A subtree rooted at any node $x$ contains at least $2^{\text{bh}(x)} - 1$ internal nodes.

*Proof by induction on height of $x$:*

- **Base case:** If $x$ is a leaf (NIL), it has 0 internal nodes. $\text{bh}(\text{NIL}) = 0$, and $2^0 - 1 = 0$. ✓
- **Inductive step:** If $x$ is an internal node with children $l$ and $r$:
  - Each child has black-height $\ge \text{bh}(x) - 1$ (they may be red, contributing 0, but the black nodes below them satisfy the black-height of $x$ minus the contribution of $x$ itself if black).
  - By induction, each child's subtree has $\ge 2^{\text{bh}(x)-1} - 1$ internal nodes.
  - Total for $x$'s subtree: $1 + (2^{\text{bh}(x)-1} - 1) + (2^{\text{bh}(x)-1} - 1) = 2^{\text{bh}(x)} - 1$.

**Applying to the root:**

Let $h$ = height of the tree and $\text{bh}$ = black-height of the root.

By invariant 4 (no two consecutive red nodes), at least half of the nodes on any root-to-leaf path are black. Therefore:

$$\text{bh}(\text{root}) \ge h/2$$

From the lemma:

$$n \ge 2^{\text{bh}(\text{root})} - 1 \ge 2^{h/2} - 1$$

Solving for $h$:

$$n + 1 \ge 2^{h/2} \implies h/2 \le \log_2(n+1) \implies h \le 2\log_2(n+1)$$

**Answer:** The height of a red-black tree with $n$ internal nodes is at most $2\log_2(n+1) = O(\log n)$.

</details>


##### **4.10. Compare AVL and Red-Black Trees** (Lecture 6, Exercise 6.10)

Compare AVL trees and red-black trees: when would you prefer one over the other?

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Both structures guarantee $O(\log n)$ operations, but they differ in their height constants and the cost of rebalancing.

**Height comparison:**

- AVL trees: height $\le 1.44\log_2(n+1)$ — strictly tighter bound
- Red-black trees: height $\le 2\log_2(n+1)$ — up to ~39% taller

**Rotation cost comparison:**

- **Insert:** Both require $O(1)$ rotations ($\le 2$ for AVL, $\le 2$ for red-black)
- **Delete:** AVL requires up to $O(\log n)$ rotations; red-black requires at most **3 rotations**

**Practical guidance:**

| Scenario | Preferred structure |
|:---|:---|
| **Read-heavy workload** (many lookups, few modifications) | **AVL tree** — lower height means fewer comparisons per lookup |
| **Write-heavy workload** (frequent insertions and deletions) | **Red-black tree** — fewer rotations on delete, lower overhead |
| **Mixed workload** | Red-black tree (used in most standard library implementations, e.g., Java `TreeMap`, C++ `std::map`) |
| **Memory-constrained** | AVL tree (only needs height per node; red-black needs 1-bit color per node) |

**Real-world use:** Most language standard libraries (Java, C++, Linux kernel's interval trees) use red-black trees because writes are common in practice and the simpler deletion fix (at most 3 rotations) outweighs the slightly taller tree.

**Answer:** Prefer **AVL trees** when lookups dominate; prefer **red-black trees** when insertions and deletions are frequent.

</details>

##### **4.11. Red-Black Tree Height — Recurrence for $B(h)$** (Problem Set 6, Problem 6.1)

Let $B(h)$ denote the minimum possible number of internal nodes in a red-black tree of black-height $h$ (black-height is the number of black nodes from root to leaf; empty tree has $\text{bh} = 0$, singleton has $\text{bh} = 1$).

1. Derive a recurrence relation for $B(h)$ and compute $B(h)$ for $h = 0, 1, 2, 3, 4, 5, 6, 7, 8$.
2. Derive a lower bound of the form $B(h) \ge c \cdot 2^{h/2}$.
3. Give a big-$O$ upper bound on the height of a red-black tree as a function of the number of keys $n$.
4. For $n = 10^6$, estimate the maximal possible height.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** To minimize internal nodes at a given black-height, maximize the use of red nodes (which contribute to node count without increasing black-height).

**Part 1 — Recurrence:**

A red-black tree of black-height $h$ has a root (black) and two subtrees. To minimize node count, we want the smallest possible subtrees. Each subtree must have black-height $h-1$. To minimize further, we can insert one red node at the top of each subtree (maximizing "structural efficiency").

The minimum is achieved with two children that are themselves minimum red-black trees of black-height $h-1$, optionally with red roots:

$$B(h) = 2B(h-1) + 1$$

with base case $B(0) = 0$ (empty tree, black-height 0) and considering black-height increments:

More precisely: a red-black tree of black-height $h \ge 1$ has a black root, and each child subtree must have black-height $h-1$. But each child may be red (adding one node "for free" before black-height increases). The minimum subtree of black-height $h-1$ with a possible red root has $B(h-1)$ nodes (if the root is black) or we can add a red root on top of a $B(h-1)$ tree. The minimum structure:

$$B(h) = 2B(h-1) + 1 \quad \text{with } B(0) = 0$$

**Computed values:**

| $h$ | $B(h) = 2B(h-1)+1$ |
|:---:|:---:|
| 0 | 0 |
| 1 | 1 |
| 2 | 3 |
| 3 | 7 |
| 4 | 15 |
| 5 | 31 |
| 6 | 63 |
| 7 | 127 |
| 8 | 255 |

Note: $B(h) = 2^h - 1$.

**Part 2 — Lower bound:**

From the closed form $B(h) = 2^h - 1 \ge 2^h / 2 = \frac{1}{2} \cdot 2^h$. Thus:

$$B(h) \ge \frac{1}{2} \cdot 2^h$$

This satisfies $B(h) \ge c \cdot 2^{h/2}$ as well (since $2^h \ge 2^{h/2}$ for $h \ge 0$), with $c = 1/2$.

**Part 3 — Height bound:**

A tree with $n$ internal nodes has black-height $\text{bh}$ satisfying $n \ge B(\text{bh}) = 2^{\text{bh}} - 1$, so $\text{bh} \le \log_2(n+1)$.

The total height $h \le 2 \cdot \text{bh}$ (since at most half the nodes on any path are red), giving:

$$h \le 2\log_2(n+1) = O(\log n)$$

**Part 4 — Estimate for $n = 10^6$:**

$$h \le 2\log_2(10^6 + 1) \approx 2 \times \log_2(10^6) = 2 \times 19.93 \approx \mathbf{40}$$

**Answer:**
1. $B(h) = 2B(h-1) + 1$ with $B(0) = 0$; closed form $B(h) = 2^h - 1$.
2. $B(h) \ge \frac{1}{2} \cdot 2^h$ (equivalently $B(h) \ge 1 \cdot 2^{h/2}$ for all $h \ge 0$ since $2^h \ge 2^{h/2}$).
3. $h \le 2\log_2(n+1) = O(\log n)$.
4. For $n = 10^6$, maximum height $\approx 40$.

</details>

##### **4.12. Build and Delete from a BST** (Problem Set 6, Problem 6.2)

Consider the BST built by inserting the following keys in that order:

$$7, 3, 5, 10, 4, 2, 6, 8, 9, 12, 1, 11$$

1. Draw the resulting BST.
2. Delete keys $10, 9, 8, 7, 6$ in that order. For each deletion, state which case applies and draw the tree after each deletion. Specify whether you used the successor or predecessor when deleting a node with two children.
3. List all possible keys that could be in the root after deleting $10, 9, 8, 7, 6$ in order (considering all choices of successor or predecessor).

<details>
<summary>Click to see the solution</summary>

**Part 1 — Build the BST:**

Insert each key following BST rules (go left if smaller, right if larger):

- **7:** root
- **3:** $3 < 7$ → left of 7
- **5:** $5 < 7$ → left; $5 > 3$ → right of 3
- **10:** $10 > 7$ → right of 7
- **4:** $4 < 7$ → left; $4 > 3$ → right; $4 < 5$ → left of 5
- **2:** $2 < 7$ → left; $2 < 3$ → left of 3
- **6:** $6 < 7$ → left; $6 > 3$ → right; $6 > 5$ → right of 5
- **8:** $8 > 7$ → right; $8 < 10$ → left of 10
- **9:** $9 > 7$ → right; $9 < 10$ → left; $9 > 8$ → right of 8
- **12:** $12 > 7$ → right; $12 > 10$ → right of 10
- **1:** ... → left of 2
- **11:** $11 > 7$ → right; $11 > 10$ → right; $11 < 12$ → left of 12

**Resulting BST:**
```
              7
            /   \
           3    10
          / \   / \
         2   5 8  12
        /   / \ \  /
       1   4   6 9 11
```

**Part 2 — Deletions:**

**Delete 10 (two children: 8 and 12):** Case — two children. Use **successor** (smallest in right subtree = 11). Replace 10 with 11; delete 11 from its original position (leaf).
```
              7
            /   \
           3    11
          / \   / \
         2   5 8  12
        /   / \ \
       1   4   6 9
```

**Delete 9 (leaf):** Case — leaf. Simply remove.
```
              7
            /   \
           3    11
          / \   / \
         2   5 8  12
        /   / \
       1   4   6
```

**Delete 8 (leaf):** Case — leaf. Simply remove.
```
              7
            /   \
           3    11
          / \     \
         2   5    12
        /   / \
       1   4   6
```

**Delete 7 (two children: 3 and 11):** Case — two children. Use **successor** (smallest in right subtree = 11). But 11 has a right child 12. Replace 7's key with 11; remove 11, promoting 12.
```
              11
            /    \
           3     12
          / \
         2   5
        /   / \
       1   4   6
```

**Delete 6 (leaf):** Case — leaf. Simply remove.
```
              11
            /    \
           3     12
          / \
         2   5
        /   /
       1   4
```

**Part 3 — All possible roots after the deletions:**

The only node with two children that required a choice is **7** (deleted in step 4). Its successor is 11 and its predecessor is 6 (but 6 still exists at that point).

If **successor (11)** is used: root becomes **11** (as shown above).

If **predecessor (6)** is used when deleting 7: Replace 7 with 6; then we still have 6 in the tree at its original position — but wait, 6 is the predecessor of 7 in the current tree (it's the rightmost node of 7's left subtree, namely right child of 5). Replace 7 with 6, then delete 6 from its position under 5.

In that case the root becomes **6**.

Also note: when deleting 10 (the first deletion), the successor of 10 is 11 and predecessor is 9. If we chose predecessor: replace 10 with 9, delete 9 (leaf). This gives a different tree shape going forward, which might change what ends up as root when 7 is deleted. Tracing: after deleting 10→9 (by predecessor), tree has 9 at position where 10 was; 9 has right child 12. Then delete original-9: already gone (used as predecessor). After deleting 8 (leaf), tree: root 7, left 3, right 9 (with right 12). Delete 7: successor of 7 is 8 — but 8 is already deleted! So successor is 9. Root becomes **9**. If predecessor is used: predecessor of 7 (in this tree) is 6. Root becomes **6**.

**All possible keys that can end up as root:** $\{6, 9, 11\}$.

**Answer:**
1. BST as drawn above.
2. Deletions use cases as described; tree evolves through the shown states.
3. The possible root values after all deletions are **6, 9, or 11**.

</details>

##### **4.13. $d$-ary Tree in an Array** (Problem Set 6, Problem 6.3)

Consider storing an ordered rooted $d$-ary tree in an array using level-order numbering (root at index 0).

1. For a perfect $d$-ary tree of height $h$, derive a closed-form formula for the total number of nodes $N(d, h)$.
2. Derive formulas for (a) the index of the parent of node at index $i > 0$, and (b) the index of the $k$-th child ($1 \le k \le d$) of the node at index $i$.
3. For a quaternary tree ($d = 4$) in an array of length $n = 70$, list all children of the node at index $i = 17$ that exist in the array.

<details>
<summary>Click to see the solution</summary>

**Part 1 — Total nodes in a perfect $d$-ary tree of height $h$:**

Level 0 (root) has 1 node. Level 1 has $d$ nodes. Level $k$ has $d^k$ nodes. Total:

$$N(d, h) = \sum_{k=0}^{h} d^k = \frac{d^{h+1} - 1}{d - 1}$$

For $d = 2$ (binary): $N(2, h) = 2^{h+1} - 1$. For $d = 4$, $h = 2$: $N = (4^3 - 1)/3 = 63/3 = 21$.

**Part 2 — Index formulas:**

**(a) Parent of node at index $i > 0$:**

The children of node at index $j$ occupy indices $dj+1$ through $dj+d$. Node $i$ is a child of node $j$ if $dj+1 \le i \le dj+d$, i.e., $j = \lfloor(i-1)/d\rfloor$.

$$\text{parent}_d(i) = \left\lfloor \frac{i-1}{d} \right\rfloor \quad (i > 0)$$

**(b) $k$-th child of node at index $i$:**

Node at index $i$ has children at indices $di+1, di+2, \ldots, di+d$. The $k$-th child ($1 \le k \le d$) is at:

$$\text{child}_{d,k}(i) = d \cdot i + k$$

**Verification for $d = 2$:** $\text{child}_{2,1}(i) = 2i+1$ (left), $\text{child}_{2,2}(i) = 2i+2$ (right). Matches the standard binary tree formulas. ✓

**Part 3 — Children of node at $i = 17$ for $d = 4$, $n = 70$:**

$$\text{child}_{4,k}(17) = 4 \times 17 + k = 68 + k$$

So the children are at indices $68+1=69$, $68+2=70$, $68+3=71$, $68+4=72$.

The array has indices $0$ to $n-1 = 69$. So:

- Index 69: **exists** (valid, $69 \le 69$)
- Index 70: **does not exist** ($70 > 69$)
- Index 71: does not exist
- Index 72: does not exist

**Answer:**
1. $N(d,h) = \dfrac{d^{h+1}-1}{d-1}$
2. (a) $\text{parent}_d(i) = \lfloor(i-1)/d\rfloor$; (b) $\text{child}_{d,k}(i) = d \cdot i + k$
3. Only child at index **69** exists in the array.

</details>

##### **4.14. Recurrence Relation and Master Theorem** (Midterm Spring 2025, Question A)

Compute the asymptotic worst-case time complexity of the following `solve` procedure:

```
solve(A, n):
    return helper(A, 0, n - 1)

helper(A, l, r):
    if r - l <= 1
        return 1
    else
        k := ceil((r - l + 1) / 9)
        a = helper(A, l, min(l + 3*k, r))
        b = helper(A, l + 2*k, min(l + 5*k, r))
        c = helper(A, l + 4*k, min(l + 7*k, r))
        d = helper(A, l + 6*k, r)
        for i from l to r:
            for j from l to r:
                if A[i] > A[j]:
                    exchange A[i] with A[j]
        return a
```

1. Express the running time $T(n)$ as a recurrence relation.
2. Find the asymptotic complexity using the master method.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Identify the number of recursive calls, their sizes, and the non-recursive work per call.

**Part 1 — Recurrence:**

The input size is $n = r - l + 1$. Setting $k = \lceil n/9 \rceil \approx n/9$:

- `helper(A, l, l + 3k)` works on approximately $3k \approx n/3$ elements
- `helper(A, l+2k, l+5k)` works on approximately $3k \approx n/3$ elements
- `helper(A, l+4k, l+7k)` works on approximately $3k \approx n/3$ elements
- `helper(A, l+6k, r)` works on approximately $3k \approx n/3$ elements

There are **4 recursive calls**, each of size $\approx n/3$.

The double loop runs in $\Theta(n^2)$ (the range $[l, r]$ has $n$ elements).

$$T(n) = 4T(n/3) + \Theta(n^2)$$

**Part 2 — Master theorem:**

Parameters: $a = 4$, $b = 3$, $f(n) = \Theta(n^2)$.

Compute $n^{\log_b a} = n^{\log_3 4} \approx n^{1.26}$.

Check Case 3: $f(n) = n^2 = \Omega(n^{\log_3 4 + \epsilon})$ for $\epsilon = 2 - \log_3 4 \approx 0.74 > 0$. ✓

Regularity: $a \cdot f(n/b) = 4 \cdot (n/3)^2 = 4n^2/9 \le (4/9) \cdot f(n)$ with $c = 4/9 < 1$. ✓

**Case 3 applies:** $T(n) = \Theta(f(n)) = \Theta(n^2)$.

**Answer:** $T(n) = 4T(n/3) + \Theta(n^2)$. By Master Theorem Case 3, $T(n) = \Theta(n^2)$.

</details>

##### **4.15. Radix Sort with Different Inner Sorts** (Midterm Spring 2025, Question B)

Consider sorting $p$ phrases using RADIX-SORT, treating each word as a "digit":

- Each phrase is a list of at most $w$ words.
- Each word is a list of at most $s$ symbols.
- Each symbol comes from an alphabet of size $a$.

What is the worst-case time complexity for sorting $p$ phrases if words are sorted using:

1. INSERTION-SORT
2. COUNTING-SORT
3. RADIX-SORT (treating each symbol as a digit of the word)

<details>
<summary>Click to see the solution</summary>

**Key Concept:** RADIX-SORT on phrases makes $w$ passes, one per word-position. Each pass sorts $p$ phrases by one word-digit using a stable sort of words.

**Case 1 — Words sorted by INSERTION-SORT:**

Comparing two words takes $O(s)$ (up to $s$ character comparisons). INSERTION-SORT on $p$ items makes $O(p^2)$ comparisons.

One pass: $O(p^2 \cdot s)$. Total for $w$ passes: $\mathbf{O(w \cdot p^2 \cdot s)}$.

**Case 2 — Words sorted by COUNTING-SORT:**

Map each word to an integer key in $[0, a^s)$. COUNTING-SORT with range $a^s$ on $p$ items: $O(p + a^s)$.

Total for $w$ passes: $\mathbf{O(w \cdot (p + a^s))}$.

**Case 3 — Words sorted by RADIX-SORT (each symbol as a digit):**

RADIX-SORT on $p$ words of length $s$ from alphabet of size $a$: $s$ passes each costing $O(p + a)$. One word-sort: $O(s(p + a))$.

Total for $w$ outer passes: $\mathbf{O(w \cdot s \cdot (p + a))}$.

**Answer:**
1. $O(w \cdot p^2 \cdot s)$
2. $O(w \cdot (p + a^s))$
3. $O(w \cdot s \cdot (p + a))$

</details>

##### **4.16. List Intersection Complexity** (Midterm Spring 2025, Question C)

Consider the intersection of $k$ non-empty sorted lists of integers with $m$ total elements. For each representation, determine whether an $O(mk)$ algorithm is possible:

1. **Dynamic Array List**
2. **Singly-Linked List (no tail pointer)**
3. **Doubly-Linked List**

<details>
<summary>Click to see the solution</summary>

**Key Concept:** The algorithm compares the fronts of all $k$ lists and advances the minimum-front list. Each step costs $O(k)$ comparisons plus $O(1)$ advancement per element.

**Case 1 — Dynamic Array:** Maintain current index per list; advancing is $O(1)$. Total: $O(mk)$. **YES**.

**Case 2 — Singly-Linked List:** Maintain current-node pointer per list; advancing via `next` is $O(1)$. Total: $O(mk)$. **YES**.

**Case 3 — Doubly-Linked List:** Same as singly-linked; `next` pointer gives $O(1)$ advancement. Total: $O(mk)$. **YES**.

**Answer:** All three representations support an $O(mk)$ intersection algorithm.

1. Dynamic Array: **YES**
2. Singly-Linked List: **YES**
3. Doubly-Linked List: **YES**

</details>

##### **4.17. Asymptotic Comparisons Table** (Midterm Spring 2025, Question D)

For each pair $(A, B)$ below, determine whether $A = O(B)$, $A = o(B)$, $A = \Omega(B)$, $A = \omega(B)$, $A = \Theta(B)$:

| $A$ | $B$ |
|:---|:---|
| $n^n$ | $n!$ |
| $n^{2025/2024}$ | $(2025/2024)^n$ |
| $\log^{2025} n$ | $n^{1/2025}$ |
| $n \log_2 n$ | $n^2 / \log_2 n$ |
| $\log_2(n^n)$ | $\log_2(n!)$ |

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Use growth hierarchies: $\log^k n \ll n^\epsilon \ll n^c \ll c^n \ll n^n$. Apply Stirling's approximation $n! \approx (n/e)^n \sqrt{2\pi n}$.

**Row 1: $n^n$ vs $n!$**

By Stirling: $n^n / n! \approx n^n / (n/e)^n = e^n \to \infty$. So $n^n = \omega(n!)$.

$O$: NO, $o$: NO, $\Omega$: YES, $\omega$: YES, $\Theta$: NO.

**Row 2: $n^{2025/2024}$ vs $(2025/2024)^n$**

Any polynomial is $o$ of any exponential: $n^c = o(r^n)$ for $r > 1$.

$O$: YES, $o$: YES, $\Omega$: NO, $\omega$: NO, $\Theta$: NO.

**Row 3: $\log^{2025} n$ vs $n^{1/2025}$**

Any power of $\log n$ is $o$ of any positive power of $n$.

$O$: YES, $o$: YES, $\Omega$: NO, $\omega$: NO, $\Theta$: NO.

**Row 4: $n \log_2 n$ vs $n^2 / \log_2 n$**

Ratio: $(n \log n) / (n^2/\log n) = \log^2 n / n \to 0$.

$O$: YES, $o$: YES, $\Omega$: NO, $\omega$: NO, $\Theta$: NO.

**Row 5: $\log_2(n^n)$ vs $\log_2(n!)$**

$\log_2(n^n) = n\log_2 n$. By Stirling: $\log_2(n!) = n\log_2 n - n\log_2 e + O(\log n) \sim n\log_2 n$. So they are $\Theta$ of each other.

$O$: YES, $o$: NO, $\Omega$: YES, $\omega$: NO, $\Theta$: YES.

**Answer table:**

| $A$ | $B$ | $O$ | $o$ | $\Omega$ | $\omega$ | $\Theta$ |
|:---|:---|:---:|:---:|:---:|:---:|:---:|
| $n^n$ | $n!$ | NO | NO | YES | YES | NO |
| $n^{2025/2024}$ | $(2025/2024)^n$ | YES | YES | NO | NO | NO |
| $\log^{2025} n$ | $n^{1/2025}$ | YES | YES | NO | NO | NO |
| $n \log_2 n$ | $n^2/\log_2 n$ | YES | YES | NO | NO | NO |
| $\log_2(n^n)$ | $\log_2(n!)$ | YES | NO | YES | NO | YES |

</details>

##### **4.18. Master Theorem Application** (Midterm Spring 2025, Question E)

Consider the recurrence $T(n) = 64T(n/4) + 7n^3 + 49$. Which applies?

1. $T(n) = \Theta(n^3 \log n)$
2. $T(n) = \Theta(n^4)$
3. Master theorem is not applicable
4. None of the above

Justify your answer.

<details>
<summary>Click to see the solution</summary>

**Parameters:** $a = 64$, $b = 4$, $f(n) = 7n^3 + 49 = \Theta(n^3)$.

**Compute $n^{\log_b a}$:**

$$n^{\log_4 64} = n^{\log_4 4^3} = n^3$$

**Compare:** $f(n) = \Theta(n^3) = \Theta(n^{\log_4 64} \cdot \log^0 n)$ → **Case 2** with $k = 0$.

$$T(n) = \Theta(n^3 \log^{0+1} n) = \Theta(n^3 \log n)$$

**Answer:** Option **1**: $T(n) = \Theta(n^3 \log n)$. Master Theorem Case 2 applies since $f(n) = \Theta(n^{\log_4 64}) = \Theta(n^3)$.

</details>

##### **4.19. Hash Table with Linear Probing** (Midterm Spring 2025, Question F)

Consider a hash table with 11 slots and $h(k) = (k + 1) \bmod 11$ with linear probing:

| Index | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Key | — | 23 | — | 35 | 2 | 15 | 5 | 13 | — | — | 9 |

Answer YES/NO:

1. Could key 9 have been inserted before 23?
2. Could key 23 have been inserted before 2?
3. Could key 13 have been inserted before 5?
4. Could key 15 have been inserted before 35?

<details>
<summary>Click to see the solution</summary>

**Home positions** (where $h(k)$ maps to):

- $h(9) = 10$ → stored at **10** (home) ✓
- $h(23) = 24 \bmod 11 = 2$ → stored at **1** (wrapped past 2,3,...,10,0 → all occupied at insertion time)
- $h(35) = 36 \bmod 11 = 3$ → stored at **3** (home) ✓
- $h(2) = 3 \bmod 11 = 3$ → stored at **4** (home 3 taken by 35 → probed to 4)
- $h(15) = 16 \bmod 11 = 5$ → stored at **5** (home) ✓
- $h(5) = 6 \bmod 11 = 6$ → stored at **6** (home) ✓
- $h(13) = 14 \bmod 11 = 3$ → stored at **7** (home 3 taken by 35; 4 by 2; 5 by 15; 6 by 5; probed to 7)

**Analysis:**

1. **9 before 23?** Key 9 sits at its home (10). It doesn't block any slot between index 2 and index 0 that 23 needs — it *does* occupy slot 10, which 23 needs to probe through (when wrapping). So 9 at 10 is *required* for 23 to end up at 1. Therefore, 9 **must** have been inserted before 23 (or at minimum was present when 23 was inserted). **YES**.
2. **23 before 2?** Key 2 needs home 3 to be occupied (by 35). Key 23 lives at slot 1 and has home 2, not 3. Inserting 23 before 2 does not affect whether slot 3 is available for 2; slot 3 must be occupied by 35 regardless. **YES**.
3. **13 before 5?** Key 13 is at slot 7, home 3. Linear probe: 3→4→5→6→7. For 13 to land at 7, slots 3,4,5,6 must all be occupied at time of 13's insertion. Slot 6 is occupied by key 5 (home 6). So key 5 must be in slot 6 when 13 is inserted, meaning **5 was inserted before 13**. Conclusion: 13 could **not** have been inserted before 5. **NO**.
4. **15 before 35?** Key 15 has home 5 and sits there. Key 35 has home 3. They don't occupy each other's probing paths. **YES**.

**Answer:** 1. YES, 2. YES, 3. NO, 4. YES.

</details>

##### **4.20. True/False Statements** (Midterm Spring 2025, Questions G–H)

For each statement, determine TRUE or FALSE:

1. If $\log(f(n)) = \Theta(\log(g(n)))$ then $f(n) = \Theta(g(n))$.
2. Incrementing a binary counter takes $\Theta(1)$ amortized, but $\Theta(\log k)$ worst case (where $k$ is the number of digits).
3. In a stack using dynamic arrays, PUSH has $\Theta(n)$ worst-case time complexity.
4. Inserting into a hash table with separate chaining is $O(1 + \alpha)$ average case.
5. Finding a value in a hash table with load factor $\alpha$ is $\Theta(1)$ worst case as long as $\text{hash}(k)$ runs in $\Theta(1)$.
6. If the master theorem is not applicable, the algorithm can loop indefinitely.
7. Dynamic Programming is an implementation in a dynamically-typed language.
8. QUICK-SORT has $\Theta(n \log n)$ worst-case time complexity.
9. Sorting $n$ real numbers uniformly distributed in $[0, k]$ using BUCKET-SORT takes $\Theta(n)$ average case.
10. PARTITION (used in QUICK-SORT) has quadratic worst-case time complexity.

<details>
<summary>Click to see the solution</summary>

**1. FALSE.** Counterexample: $f(n) = n$, $g(n) = n^2$. $\log f = \Theta(\log g)$ (both are $\Theta(\log n)$), but $f \ne \Theta(g)$.

**2. FALSE.** With $k$ digits, the worst case (all 1s) flips all $k$ bits: $\Theta(k)$ worst case, not $\Theta(\log k)$.

**3. TRUE.** When the array doubles, copying $n$ elements takes $\Theta(n)$ — this is the worst case for a single PUSH. The *amortized* cost is $\Theta(1)$.

**4. TRUE.** Average chain length is $\alpha$; insert requires hashing ($O(1)$) + appending ($O(1)$ or $O(\alpha)$ if checking duplicates). $O(1 + \alpha)$ is correct.

**5. FALSE.** Worst case: all $n$ keys collide into one chain. Lookup is $\Theta(n)$ worst case regardless of hash computation speed.

**6. FALSE.** "Master theorem not applicable" means the recurrence doesn't fit the required form; the algorithm still terminates. We simply use another analysis method.

**7. FALSE.** Dynamic Programming is an algorithmic design technique for optimization problems with overlapping subproblems. Language typing is irrelevant.

**8. FALSE.** QUICK-SORT is $\Theta(n \log n)$ *average case* and $\Theta(n^2)$ *worst case* (e.g., sorted input with last-element pivot).

**9. TRUE.** With uniform distribution over $n$ buckets, expected elements per bucket is $O(1)$. Each bucket is sorted in $O(1)$ expected time. Total: $\Theta(n)$ average case.

**10. FALSE.** PARTITION runs in $\Theta(n)$ time — it makes one linear scan. It is QUICK-SORT as a whole that has $\Theta(n^2)$ worst case.

**Answer summary:**

| # | Truth |
|:---:|:---:|
| 1 | FALSE |
| 2 | FALSE |
| 3 | TRUE |
| 4 | TRUE |
| 5 | FALSE |
| 6 | FALSE |
| 7 | FALSE |
| 8 | FALSE |
| 9 | TRUE |
| 10 | FALSE |

</details>

##### **4.21. Rod Cutting with DP** (Midterm Spring 2025, Question H)

Answer the following about the Rod Cutting problem:

1. Using DP with *tabulation*, what is the worst-case time complexity of finding the maximum revenue for a rod of length $n$?
2. Using DP with *memoization* (finding both maximum revenue and cut positions), what is the worst-case time complexity?
3. Find the maximum revenue for a rod of length 10 with prices:

| Length $i$ | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Price $p_i$ | 2 | 3 | 9 | 12 | 13 | 13 | 14 | 24 | 25 | 28 |

4. *(+1% extra credit)* Consider Rod Cutting with at most $k$ cuts. Give an $O(nk)$ pseudocode algorithm.

<details>
<summary>Click to see the solution</summary>

**Part 1 — Tabulation time complexity:**

Fill array $r[0..n]$; for each $j$ from 1 to $n$, try all $i$ from 1 to $j$ cuts: $\sum_{j=1}^{n} j = n(n+1)/2 = \Theta(n^2)$.

**Answer:** $\Theta(n^2)$.

**Part 2 — Memoization time complexity:**

Same subproblems, same recurrence: $\Theta(n^2)$.

**Answer:** $\Theta(n^2)$.

**Part 3 — Maximum revenue for $n = 10$:**

Recurrence: $r[j] = \max_{1 \le i \le j}(p[i] + r[j-i])$, $r[0] = 0$.

| $j$ | Calculation | $r[j]$ | Best cut |
|:---:|:---|:---:|:---:|
| 1 | $p[1] + r[0] = 2$ | **2** | cut 1 |
| 2 | $\max(2+2, 3+0) = \max(4, 3)$ | **4** | cut 1+1 |
| 3 | $\max(2+4, 3+2, 9+0) = \max(6,5,9)$ | **9** | cut 3 |
| 4 | $\max(2+9, 3+4, 9+2, 12+0) = \max(11,7,11,12)$ | **12** | cut 4 |
| 5 | $\max(2+12, 3+9, 9+4, 12+2, 13+0) = \max(14,12,13,14,13)$ | **14** | cut 1+4 or 2+... |
| 6 | $\max(2+14, 3+12, 9+9, 12+4, 13+2, 13+0) = \max(16,15,18,16,15,13)$ | **18** | cut 3+3 |
| 7 | $\max(2+18,3+14,9+12,12+9,13+4,13+2,14+0) = \max(20,17,21,21,17,15,14)$ | **21** | cut 3+4 or 4+3 |
| 8 | $\max(2+21,3+18,9+14,12+12,13+9,13+4,...,24+0) = \max(23,21,23,24,22,...,24)$ | **24** | cut 4+4 or 8 |
| 9 | $\max(2+24,3+21,...,9+12,12+9,...,25+0)$. Check $p[3]+r[6]=9+18=27$, $p[3]+r[6]=27$, $p[6]+r[3]=13+9=22$. Also $p[9]=25$. Best: $\max(26,25,27,23,...,25)$ | **27** | cut 3+3+3 |
| 10 | $\max(2+27, 3+24, 9+18, 12+14, 13+12, 13+9, 14+9, 24+4, 25+2, 28+0)$ $= \max(29,27,27,26,25,22,23,28,27,28)$ | **30** | cut 3+3+4 or no cut (28<30) |

Wait, let me recompute $r[10]$ carefully:

- $p[1]+r[9] = 2+27 = 29$
- $p[2]+r[8] = 3+24 = 27$
- $p[3]+r[7] = 9+21 = 30$ ✓
- $p[4]+r[6] = 12+18 = 30$ ✓
- $p[5]+r[5] = 13+14 = 27$
- $p[6]+r[4] = 13+12 = 25$
- $p[7]+r[3] = 14+9 = 23$
- $p[8]+r[2] = 24+4 = 28$
- $p[9]+r[1] = 25+2 = 27$
- $p[10]+r[0] = 28+0 = 28$

Maximum: **30**, achieved by cutting into pieces of length 3+7 or 4+6.

**Part 4 — At most $k$ cuts:**

Define $dp[j][c]$ = maximum revenue from a rod of length $j$ with at most $c$ cuts allowed.

$$dp[j][c] = \max\left(p[j],\ \max_{1 \le i < j}(p[i] + dp[j-i][c-1])\right)$$

Base cases: $dp[0][c] = 0$ for all $c$; $dp[j][0] = p[j]$ (no cuts allowed → sell the whole piece).

```
function ROD-CUTTING-K(p, n, k):
    dp[0..n][0..k] ← 0
    for j ← 1 to n:
        dp[j][0] ← p[j]           // no cuts: sell as length j
    for c ← 1 to k:
        for j ← 1 to n:
            dp[j][c] ← p[j]       // option: no cut
            for i ← 1 to j-1:
                dp[j][c] ← max(dp[j][c], p[i] + dp[j-i][c-1])
    return dp[n][k]
```

**Complexity:** Three nested loops: $k \times n \times n = O(n^2 k)$. To achieve $O(nk)$, observe that we can reformulate: instead of "which first cut to make," use a different DP where $dp[j][c]$ = max revenue from rod of length $j$ making *exactly* $c$ cuts (then take max over $c' \le k$). But the standard $O(n^2 k)$ version is more natural. An $O(nk)$ version requires a different subproblem formulation or convex hull trick.

**Answer:**
1. $\Theta(n^2)$
2. $\Theta(n^2)$
3. Maximum revenue = **30** (e.g., cut into 3+7 or 4+6)
4. DP defined above runs in $O(n^2 k)$; an $O(nk)$ solution requires advanced techniques.

</details>

##### **4.22. WKL — Sets and Maps System** (Coding Exercise 6, Problem A)

In the World Knowledge League (WKL), participants register knowledge areas and the system must answer queries in real time. You are given $Q$ commands ($1 \le Q \le 10^4$):

- `add <id> <k> <a_1> ... <a_k>` — add $k$ knowledge areas to participant `id`.
- `remove <id> <area>` — remove the area from participant `id`; remove participant if no areas remain.
- `has <id> <area>` — print `Yes` or `No`.
- `area_list <area>` — print all participant IDs with this area, sorted ascending.
- `areas_count` — print the total number of distinct areas.
- `specialists_count <k> <a_1> ... <a_k>` — print the number of unique participants having at least one of the given areas.
- `top_area` — print the area with the most participants (ties: lexicographically smallest).

Constraints: IDs are strings of length 10 (uppercase letters/digits); areas are lowercase letters + `_`, length 3–20; at most 4000 participants, 5000 areas, 100 areas per participant, 1000 participants per area.

**Example:**
```
Input:
11
areas_count
add IU483C9YW0 3 math computer_science spanish
areas_count
add DS983A7ILJ 3 math physics math
add IU483C9YW0 1 art
remove IU483C9YW0 computer_science
has IU483C9YW0 computer_science
areas_count
top_area
area_list math
specialists_count 3 french math physics

Output:
0
3
No
4
math
DS983A7ILJ IU483C9YW0
2
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Maintain two bidirectional mappings: participant → set of areas, and area → set of participants. Use ordered sets (e.g., `std::map` / `std::set` in C++) for sorted output and efficient operations.

**Data structures:**

```cpp
#include <bits/stdc++.h>
using namespace std;

// participant_id -> set of areas (ordered set for no duplicates)
map<string, set<string>> participant_areas;

// area -> set of participant IDs (ordered for sorted area_list output)
map<string, set<string>> area_participants;
```

**Command implementations:**

```cpp
// add <id> <k> <a_1> ... <a_k>
void cmd_add(const string& id, vector<string>& areas) {
    for (auto& a : areas) {
        if (participant_areas[id].insert(a).second) {
            // area was newly added for this participant
            area_participants[a].insert(id);
        }
    }
}

// remove <id> <area>
void cmd_remove(const string& id, const string& area) {
    auto& s = participant_areas[id];
    if (s.erase(area)) {
        area_participants[area].erase(id);
        if (area_participants[area].empty())
            area_participants.erase(area);
    }
    if (s.empty())
        participant_areas.erase(id);
}

// has <id> <area>
void cmd_has(const string& id, const string& area) {
    auto it = participant_areas.find(id);
    bool found = it != participant_areas.end() && it->second.count(area);
    cout << (found ? "Yes" : "No") << "\n";
}

// area_list <area>
void cmd_area_list(const string& area) {
    auto it = area_participants.find(area);
    if (it == area_participants.end()) { cout << "\n"; return; }
    bool first = true;
    for (auto& id : it->second) {
        if (!first) cout << " ";
        cout << id;
        first = false;
    }
    cout << "\n";
}

// areas_count
void cmd_areas_count() {
    cout << area_participants.size() << "\n";
}

// specialists_count <k> <a_1> ... <a_k>
void cmd_specialists_count(vector<string>& areas) {
    set<string> result;
    for (auto& a : areas) {
        auto it = area_participants.find(a);
        if (it != area_participants.end())
            for (auto& id : it->second) result.insert(id);
    }
    cout << result.size() << "\n";
}

// top_area
void cmd_top_area() {
    string best_area;
    size_t best_count = 0;
    for (auto& [area, participants] : area_participants) {
        // map iterates in lexicographic order, so first tie wins
        if (participants.size() > best_count) {
            best_count = participants.size();
            best_area = area;
        }
    }
    cout << best_area << "\n";
}
```

**Walkthrough of the example:**

1. `areas_count` → 0 (no areas yet)
2. `add IU483C9YW0 3 math computer_science spanish` → adds 3 areas; system has {math, computer_science, spanish}
3. `areas_count` → 3
4. `add DS983A7ILJ 3 math physics math` → adds math, physics (second "math" ignored as duplicate)
5. `add IU483C9YW0 1 art` → adds art to IU483C9YW0
6. `remove IU483C9YW0 computer_science` → removes computer_science from IU483C9YW0; areas now {math, spanish, art, physics}
7. `has IU483C9YW0 computer_science` → **No**
8. `areas_count` → **4**
9. `top_area` → **math** (has 2 participants; all others have 1)
10. `area_list math` → **DS983A7ILJ IU483C9YW0** (sorted)
11. `specialists_count 3 french math physics` → "french" not in system (ignored); participants with math or physics = {IU483C9YW0, DS983A7ILJ} → **2**

**Complexity:** Each `add`/`remove` is $O(\log n)$ for map/set operations. `specialists_count` is $O(k \cdot 1000 \cdot \log n)$ in the worst case. `top_area` is $O(|\text{areas}|)$ per call.

**Answer:** The solution uses two ordered maps for bidirectional participant–area mapping. See the C++ implementation above.

</details>

##### **4.23. AVL Tree Implementation** (Coding Exercise 6, Problem B)

Starting from an empty AVL tree, process $N$ commands ($1 \le N \le 10^5$):

- `ADD k v` — insert key $k$ with value $v$. If $k$ already exists, print `KEY ALREADY EXISTS`.
- `LOOKUP k` — print the value $v$ for key $k$. If not found, print `KEY NOT FOUND`.
- `DELETE k` — delete key $k$. If not found, print `KEY NOT FOUND`.
- `PRINT_ROTATIONS` — print the total number of rotations performed so far.

Keys and values are integers in $[-100, 100]$.

**Example:**
```
Input: 10
ADD 7 5 / ADD 3 2 / ADD 10 100 / ADD 2 3 / ADD 6 6
ADD 4 4 / ADD 8 7 / DELETE 7 / PRINT_ROTATIONS / LOOKUP 8
Output: 4 / 7
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Implement a full AVL tree with height-tracking at each node. After each insert or delete, walk back up the tree updating heights and applying rotations where $|\text{bf}| > 1$. Count every rotation (each single rotation = 1; each double rotation = 2).

**Node structure:**

```cpp
struct Node {
    int key, val, height;
    Node *left, *right;
    Node(int k, int v) : key(k), val(v), height(0), left(nullptr), right(nullptr) {}
};

int height(Node* n) { return n ? n->height : -1; }

void update_height(Node* n) {
    if (n) n->height = 1 + max(height(n->left), height(n->right));
}

int bf(Node* n) { return n ? height(n->left) - height(n->right) : 0; }
```

**Rotations (each increments a global counter):**

```cpp
int rotation_count = 0;

Node* rotate_right(Node* y) {
    rotation_count++;
    Node* x = y->left;
    y->left = x->right;
    x->right = y;
    update_height(y);
    update_height(x);
    return x;
}

Node* rotate_left(Node* x) {
    rotation_count++;
    Node* y = x->right;
    x->right = y->left;
    y->left = x;
    update_height(x);
    update_height(y);
    return y;
}
```

**Rebalancing after each insertion/deletion:**

```cpp
Node* rebalance(Node* n) {
    update_height(n);
    int b = bf(n);
    if (b == 2) {          // left-heavy
        if (bf(n->left) < 0)
            n->left = rotate_left(n->left);   // LR case
        return rotate_right(n);
    }
    if (b == -2) {         // right-heavy
        if (bf(n->right) > 0)
            n->right = rotate_right(n->right); // RL case
        return rotate_left(n);
    }
    return n;
}
```

**Insert:**

```cpp
Node* insert(Node* n, int key, int val, bool& already_exists) {
    if (!n) return new Node(key, val);
    if (key == n->key) { already_exists = true; return n; }
    if (key < n->key) n->left = insert(n->left, key, val, already_exists);
    else              n->right = insert(n->right, key, val, already_exists);
    return rebalance(n);
}
```

**Delete:**

```cpp
Node* find_min(Node* n) { return n->left ? find_min(n->left) : n; }

Node* delete_min(Node* n) {
    if (!n->left) return n->right;
    n->left = delete_min(n->left);
    return rebalance(n);
}

Node* remove(Node* n, int key, bool& found) {
    if (!n) return nullptr;
    if (key < n->key) n->left = remove(n->left, key, found);
    else if (key > n->key) n->right = remove(n->right, key, found);
    else {
        found = true;
        if (!n->right) { Node* tmp = n->left; delete n; return tmp; }
        Node* succ = find_min(n->right);
        n->key = succ->key; n->val = succ->val;
        n->right = delete_min(n->right);
    }
    return rebalance(n);
}
```

**Main loop:**

```cpp
int main() {
    int N; cin >> N;
    Node* root = nullptr;
    while (N--) {
        string cmd; cin >> cmd;
        if (cmd == "ADD") {
            int k, v; cin >> k >> v;
            bool exists = false;
            root = insert(root, k, v, exists);
            if (exists) cout << "KEY ALREADY EXISTS\n";
        } else if (cmd == "LOOKUP") {
            int k; cin >> k;
            Node* cur = root;
            while (cur && cur->key != k) cur = k < cur->key ? cur->left : cur->right;
            if (!cur) cout << "KEY NOT FOUND\n";
            else cout << cur->val << "\n";
        } else if (cmd == "DELETE") {
            int k; cin >> k;
            bool found = false;
            root = remove(root, k, found);
            if (!found) cout << "KEY NOT FOUND\n";
        } else { // PRINT_ROTATIONS
            cout << rotation_count << "\n";
        }
    }
}
```

**Trace of the example:**

Inserting 7, 3, 10, 2, 6, 4, 8 in order triggers rotations to maintain AVL balance. After `DELETE 7` (which may trigger rebalancing), `PRINT_ROTATIONS` outputs **4**. `LOOKUP 8` finds value **7**.

**Answer:** The AVL implementation tracks heights and balance factors at each node, applying single or double rotations as needed. For the given input, 4 total rotations occur and the value at key 8 is 7.

</details>

##### **4.24. Overlapping Slots — Interval Tree** (Coding Exercise 6, Problem C)

You are managing a tourist booking system. Bookings are intervals $[L, R]$ ($1 \le L \le R \le 10^5$). Process $Q$ queries:

- `add L R` — add a booking $[L, R]$.
- `remove L R` — remove one occurrence of $[L, R]$; ignore if not present.
- `check L R` — print `Yes` if $[L, R]$ overlaps any existing booking, `No` otherwise.
- `count L R` — print the number of bookings that start AND end within $[L, R]$ inclusively.

Two intervals $[a, b]$ and $[c, d]$ **overlap** if and only if $a \le d$ and $c \le b$.

**Example:**
```
Input: 9
count 1 10        → 0
add 1 7
check 10 20       → No
add 2 9
count 1 10        → 2
add 2 9
count 1 10        → 3
check 5 6         → Yes
remove 2 9

Output: 0 / No / 2 / 3 / Yes
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** This problem requires an efficient *interval data structure*. We need:

- Overlap detection: does any stored interval touch $[L, R]$?
- Range count: how many stored intervals $[a, b]$ satisfy $L \le a$ and $b \le R$?

**Strategy:**

Since $L, R \le 10^5$, we can use a **segment tree with lazy propagation** or a simpler approach based on the observation that overlapping intervals satisfy $a \le R$ and $b \ge L$ (equivalently, NOT ($a > R$ or $b < L$)).

For this problem with modest constraints, we can store intervals in a multimap and answer each query by iterating:

```cpp
#include <bits/stdc++.h>
using namespace std;

// Store bookings as multimap: left endpoint -> right endpoint
multimap<int, int> bookings;

// add L R
void cmd_add(int L, int R) {
    bookings.insert({L, R});
}

// remove L R
void cmd_remove(int L, int R) {
    auto range = bookings.equal_range(L);
    for (auto it = range.first; it != range.second; ++it) {
        if (it->second == R) {
            bookings.erase(it);
            return;
        }
    }
    // not found: ignore
}

// check L R: does any interval [a,b] overlap [L,R]?
// Overlap condition: a <= R and b >= L
void cmd_check(int L, int R) {
    // All intervals with a <= R: look in bookings up to key R
    auto it = bookings.begin();
    while (it != bookings.end() && it->first <= R) {
        if (it->second >= L) {
            cout << "Yes\n";
            return;
        }
        ++it;
    }
    cout << "No\n";
}

// count L R: intervals [a,b] with L <= a and b <= R
void cmd_count(int L, int R) {
    int cnt = 0;
    auto it = bookings.lower_bound(L);  // a >= L
    while (it != bookings.end() && it->first <= R) {  // a <= R
        if (it->second <= R)  // b <= R
            cnt++;
        ++it;
    }
    cout << cnt << "\n";
}
```

**Walkthrough of the example:**

1. `count 1 10` → 0 (no bookings)
2. `add 1 7` → bookings: {[1,7]}
3. `check 10 20` → any interval with $a \le 20$ and $b \ge 10$? [1,7]: $b=7 < 10$ → No. Output: **No**
4. `add 2 9` → bookings: {[1,7], [2,9]}
5. `count 1 10` → [1,7]: $1 \ge 1, 7 \le 10$ ✓; [2,9]: $2 \ge 1, 9 \le 10$ ✓ → **2**
6. `add 2 9` → bookings: {[1,7], [2,9], [2,9]}
7. `count 1 10` → all 3 are within [1,10] → **3**
8. `check 5 6` → [1,7]: $a=1 \le 6$ and $b=7 \ge 5$ → overlap! → **Yes**
9. `remove 2 9` → remove one [2,9]; bookings: {[1,7], [2,9]}

**Overlap detection for larger inputs:**

For better performance with larger $Q$, use an **interval tree** (augmented BST where each node stores the maximum right endpoint in its subtree). This allows overlap queries in $O(\log n)$ instead of $O(n)$. The `count` query requires a 2D range query, efficiently handled with a persistent segment tree or 2D BIT.

**Time complexity (naive):**

- `add`/`remove`: $O(\log n)$
- `check`: $O(n)$ worst case
- `count`: $O(n)$ worst case

**Answer:** Use a multimap keyed by left endpoint. The `check` query scans intervals with left endpoint $\le R$ looking for right endpoint $\ge L$. The `count` query counts intervals fully contained in $[L, R]$.

</details>
