---
title: "W7-W8. Sets, Maps, Binary Search Trees, AVL Trees, Red-Black Trees"
author: "Nikolai Kudasov"
date: "February 23, 2026"
format: html
engine: knitr
---

#### **1. Summary**

##### **1.1 Sets and Maps**

###### **1.1.1 The Set Abstract Data Type**

A **Set** is an abstract data type that represents an *unordered collection of values without duplicates*. Think of it like a mathematical set: $\{1, 3, 4\}$ — order does not matter, and every element appears at most once.

We work with **dynamic sets** that support operations modifying the collection at runtime:

- **add(x):** Add element $x$ to the set. If $x$ is already present, do nothing.
- **remove(x):** Remove element $x$ from the set.
- **contains(x):** Return true if $x$ is in the set, false otherwise.

**Example:**

$$\{1, 3, 4\} \xrightarrow{\text{add}(5)} \{1, 3, 4, 5\} \xrightarrow{\text{remove}(3)} \{1, 4, 5\}$$

###### **1.1.2 The Map Abstract Data Type**

A **Map** (also called a *dictionary*) is an abstract data type that represents an *unordered collection of key–value pairs with no duplicate keys*. Each key maps to exactly one value. Think of a real-world dictionary: each word (key) has exactly one definition (value).

**Example:**

$$\{\langle A,1\rangle, \langle C,3\rangle, \langle B,2\rangle\} \xrightarrow{\text{add}(5,c)} \{\langle A,1\rangle, \langle C,3\rangle, \langle B,2\rangle, \langle 5,c\rangle\}$$

The main Map ADT operations are:

- **put(k, v):** Add the key–value pair $\langle k, v\rangle$. If key $k$ already exists, replace the old value with $v$.
- **get(k):** Return the value associated with key $k$, or null if not found.
- **remove(k):** Remove the entry with key $k$.

**Important:** A map stores at most one value per key. In Java, a common interface looks like:

```java
public interface Map<K, V> {
    int size();
    boolean isEmpty();
    V get(K key);
    V put(K key, V value);   // returns the old value (or null)
    V remove(K key);
    Iterable<K> keySet();
    Iterable<V> values();
    Iterable<Entry<K,V>> entrySet();
}
```

Note: `keySet()` and `entrySet()` are useful because keys are unique (no ambiguity), while values need not be unique — there is no meaningful "valueSet".

###### **1.1.3 Map Implementations and Trade-offs**

There are several ways to implement a map, each with different trade-offs:

**Unsorted doubly linked list:**

The map entries are stored as nodes in a doubly linked list. To find a key, we scan the list linearly.

<!-- DIAGRAM HERE -->

- **get(k):** $O(n)$ worst case — must scan all entries
- **put(k, v):** $O(n)$ worst case (need to check for duplicate key), or $O(1)$ if duplicates are allowed without checking
- **remove(k):** $O(n)$ worst case; $O(1)$ if given a direct reference to the node (with back-pointers from entries to nodes)

**Array (when keys are small integers):**

If keys are integers in the range $0$ to $n-1$, we can use an array of size $n$: `array[k] = v`.

- **get(k):** $O(1)$ — direct index
- **put(k, v):** $O(1)$ — direct assignment
- **remove(k):** $O(1)$ — set to null

If the key range is large or keys are not integers, we use **hash functions** to map keys to indices (covered later in the course).

**The fundamental trade-off** when representing an ordered map (sorted by keys):

| Representation | Search | Insert/Delete |
|:---|:---|:---|
| Sorted array | $O(\log n)$ (binary search) | $O(n)$ (shifting) |
| Linked list | $O(n)$ (linear scan) | $O(1)$ (given reference) |
| Balanced BST | $O(\log n)$ | $O(\log n)$ |

Today's goal is to find data structures that achieve $O(\log n)$ for all operations — the best of both worlds.

##### **1.2 Trees**

###### **1.2.1 Basic Tree Definitions**

A **tree** is a connected acyclic undirected graph. When we designate one vertex as the **root**, we get a **rooted tree**, which we conventionally draw "hanging" from the root with edges directed downward.

<!-- DIAGRAM HERE -->

Key vocabulary for rooted trees:

- **Root:** The unique node with no parent (in-degree zero).
- **Leaf:** A node with no children (out-degree zero).
- **Parent of $v$:** The node $u$ such that $u \to v$ is an edge.
- **Children of $u$:** All nodes $v$ such that $u \to v$ is an edge.
- **Ancestor of $v$:** Any node $u$ on the path from the root to $v$.
- **Descendant of $u$:** Any node $v$ that $u$ is an ancestor of.
- **Height of a tree:** The maximum length of a path from the root to any leaf.
- **Size of a tree:** The total number of nodes.

**Ordered trees:** In an **ordered tree**, the order of children of each node is fixed and significant. Two trees with the same nodes but different child orderings are considered distinct.

**$n$-ary trees:** An **$n$-ary tree** is an ordered rooted tree in which every node has *at most $n$ children*. Special cases:

- $n = 2$: **Binary tree** (most common in this course)
- $n = 3$: **Ternary tree**
- $n = 1$: A linear chain (essentially a linked list)

###### **1.2.2 Binary Tree Representations**

**Linked representation:** Each node stores its data plus pointers to its left child, right child, and parent.

<!-- DIAGRAM HERE -->

**Array representation (level-order numbering):** We can map a binary tree to an array by numbering nodes level by level, left to right, starting from the root at index 0:

$$\text{root} = 0 \qquad \text{parent}(i) = \left\lfloor \frac{i-1}{2} \right\rfloor \qquad \text{left}(i) = 2i+1 \qquad \text{right}(i) = 2i+2$$

This works perfectly for *complete* trees (where all levels are full except possibly the last). For sparse trees, many array slots are wasted. For example, a tree with 8 actual nodes arranged in an unbalanced way might require an array of size much larger than 8.

**Example:** A tree with nodes labeled A through I at indices 0, 1, 2, 3, 5, 6, 7, 8, 11 looks like:

```
Index: [ 0 | 1 | 2 | 3 | - | 4 | 5 | 6 | 7 | - | - | 8 | - | - | - ]
Data:  [ A | B | C | D | - | E | F | G | H | - | - | I | - | - | - ]
```

(Dashes indicate empty slots where tree nodes do not exist.)

###### **1.2.3 Tree Traversal Orders**

To visit all nodes of a binary tree, we choose an order. Three standard orders exist:

1. **Pre-order:** Visit the current node, then recursively the left subtree, then the right subtree. *(Root → Left → Right)*
2. **In-order:** Recursively visit the left subtree, then the current node, then the right subtree. *(Left → Root → Right)* — For BSTs, this visits nodes in sorted key order.
3. **Post-order:** Recursively visit the left subtree, then the right subtree, then the current node. *(Left → Right → Root)*

**Example:** For the tree with root A (left subtree: B→D, B→C→{H,F}; right subtree: E→{G, I→J}):

- **Pre-order:** A, B, D, C, H, F, E, G, I, J
- **In-order:** D, B, H, C, F, A, G, E, I, J
- **Post-order:** D, H, F, C, B, G, J, I, E, A

##### **1.3 Binary Search Trees**

###### **1.3.1 The BST Property**

A **Binary Search Tree (BST)** is a binary tree that satisfies the **BST property**:

> For every node $x$: if $y$ is in the *left* subtree of $x$, then $y.\text{key} \le x.\text{key}$; if $z$ is in the *right* subtree of $x$, then $x.\text{key} \le z.\text{key}$.

<!-- DIAGRAM HERE -->

This property makes BSTs ideal for implementing ordered maps: in-order traversal visits all keys in sorted order.

**Operations supported by a BST:**

- `search(k)` — find the value for key $k$
- `insert(k, v)` — insert a key–value pair
- `remove(k)` — remove the entry with key $k$
- `findMin()` / `findMax()` — find the entry with the smallest/largest key

###### **1.3.2 Search**

To search for key $k$, start at the root and follow the BST property:

```
TREE-SEARCH(x, k):
    if x == NIL or k == x.key
        return x
    if k < x.key
        return TREE-SEARCH(x.left, k)
    else
        return TREE-SEARCH(x.right, k)
```

An equivalent iterative version avoids recursion:

```
ITERATIVE-TREE-SEARCH(x, k):
    while x ≠ NIL and k ≠ x.key
        if k < x.key
            x = x.left
        else
            x = x.right
    return x
```

**Time complexity:** $O(h)$ where $h$ is the height of the tree. Best case $O(1)$ (key at root).

###### **1.3.3 Finding Minimum and Maximum**

The minimum is always at the leftmost node; the maximum at the rightmost:

```
TREE-MINIMUM(x):           TREE-MAXIMUM(x):
    while x.left ≠ NIL         while x.right ≠ NIL
        x = x.left                 x = x.right
    return x                   return x
```

Both run in $O(h)$ time.

###### **1.3.4 Finding the Successor**

The **in-order successor** of node $x$ is the node with the smallest key greater than $x.\text{key}$. It is needed during deletion.

```
TREE-SUCCESSOR(x):
    if x.right ≠ NIL
        return TREE-MINIMUM(x.right)   // leftmost in right subtree
    else
        y = x.parent
        while y ≠ NIL and x == y.right
            x = y
            y = y.parent
        return y
```

**Intuition:** If $x$ has a right subtree, its successor is the minimum of that subtree. Otherwise, we climb the tree until we find an ancestor that is a left child of its parent — that parent is the successor.

###### **1.3.5 Insertion**

To insert key $k$ with value $v$: search for $k$ to find where it would go, then attach a new node there.

<!-- DIAGRAM HERE -->

**Time complexity:** $O(h)$ — we trace a path from root to the insertion point.

###### **1.3.6 Deletion**

Deletion has three cases depending on the node's children:

1. **Leaf node:** Simply remove it.
2. **One child:** Replace the node with its single child (splice it out).
3. **Two children:** Replace the node's key/value with its *in-order successor* (or predecessor), then delete that successor (which has at most one child, since it's the leftmost node of the right subtree).

<!-- DIAGRAM HERE -->

**Time complexity:** $O(h)$ for all cases.

###### **1.3.7 Height and the Problem of Imbalance**

All BST operations run in $O(h)$. The question is: what is $h$ in practice?

- If keys are inserted in a *random* order (e.g., 5, 3, 8, 1, 6, 9, 2), the resulting tree is roughly balanced with $h \approx \log n$.
- If keys are inserted in *sorted order* (e.g., 1, 2, 3, 4, 5, 6, 7), each new node becomes the rightmost child, creating a chain with $h = n - 1$ (essentially a linked list).

In the worst case, a BST degrades to $O(n)$ per operation. To fix this, we use **self-balancing BSTs**.

##### **1.4 AVL Trees**

###### **1.4.1 Motivation and Invariant**

**AVL trees** (named after Adelson-Velsky and Landis, 1962) are self-balancing BSTs. They maintain the following **balance invariant**:

> For every node, the heights of its left and right subtrees differ by *at most 1*.

This invariant guarantees that the height of an AVL tree with $n$ nodes is $O(\log n)$, so all operations remain $O(\log n)$.

The **balance factor** of a node is defined as:
$$\text{bf}(x) = h(\text{left}(x)) - h(\text{right}(x))$$

An AVL tree requires $|\text{bf}(x)| \le 1$ for every node $x$ (where the height of an empty subtree is $-1$).

###### **1.4.2 Rotations**

When an insertion or deletion violates the AVL invariant, we restore it using **rotations**. A rotation is a local restructuring of the tree that changes the height without disturbing the BST property.

**Right rotation at node $y$** (and its inverse, left rotation at $x$):

<!-- DIAGRAM HERE -->

```
RIGHT-ROTATE(T, y):
    x = y.left
    β = x.right
    x.right = y
    y.left = β
    update heights of y, then x
    return x   // new root of this subtree
```

A right rotation is used when the left subtree is too tall (balance factor $+2$). A left rotation is the mirror image.

**Height effects of rotations:**

If subtree heights are $h_\alpha$, $h_\beta$, $h_\gamma$ (left to right):

- Before right rotation: height of $y$'s subtree = $\max(h_\alpha, h_\beta) + 2$
- After right rotation: height of $x$'s subtree = $\max(h_\alpha, h_\beta) + 1$

The rotation reduces the height by 1, restoring balance.

###### **1.4.3 The Four Imbalance Cases**

After an insertion, at most one node may become unbalanced (the lowest such ancestor of the inserted node). There are four cases:

<!-- DIAGRAM HERE -->

| Case | Condition | Fix |
|:---|:---|:---|
| **Left-Left (LL)** | Left child's left subtree is too tall | Single right rotation at the unbalanced node |
| **Left-Right (LR)** | Left child's right subtree is too tall | Left rotation at left child, then right rotation at node |
| **Right-Right (RR)** | Right child's right subtree is too tall | Single left rotation at the unbalanced node |
| **Right-Left (RL)** | Right child's left subtree is too tall | Right rotation at right child, then left rotation at node |

**Insert and rebalance:** Insert as in a regular BST, then walk back up to the root, updating heights and applying rotations where needed.

###### **1.4.4 AVL Tree Height Bound**

Let $n_{\min}(h)$ = minimum number of nodes in an AVL tree of height $h$. Then:

$$n_{\min}(h) = n_{\min}(h-1) + n_{\min}(h-2) + 1$$

with $n_{\min}(-1) = 0$ (empty tree) and $n_{\min}(0) = 1$ (single node).

This recurrence resembles Fibonacci numbers: $n_{\min}(h) \ge 2 \cdot n_{\min}(h-2) + 1$, which gives $n_{\min}(h) \ge 2^{h/2}$. Therefore $n \ge 2^{h/2}$, so $h \le 2\log_2 n = O(\log n)$.

##### **1.5 Red-Black Trees**

###### **1.5.1 Motivation**

**Red-Black trees** are another type of self-balancing BST. Like AVL trees, they guarantee $O(\log n)$ height, but they achieve balance through a different invariant that requires fewer rotations in practice during insertions and deletions (though not asymptotically fewer).

In a red-black tree, each node is **colored** either red or black. The coloring encodes height constraints without explicitly storing heights.

###### **1.5.2 Red-Black Tree Invariants**

A valid red-black tree must satisfy *all five* of the following invariants simultaneously:

1. **Color:** Every node is either red or black.
2. **Root:** The root is black.
3. **Leaves:** All leaves (NIL sentinels representing empty subtrees) are black.
4. **Red constraint:** If a node is red, both its children are black. (*No two consecutive red nodes on any path.*)
5. **Black-height:** For every node, *every path* from that node to any leaf contains the same number of black nodes.

<!-- DIAGRAM HERE -->

The **black-height** $\text{bh}(x)$ of a node $x$ is the number of black nodes on any path from $x$ down to a leaf (not counting $x$ itself).

###### **1.5.3 Why Red-Black Trees Have $O(\log n)$ Height**

**Lemma:** A red-black tree with $n$ internal nodes has height at most $2\log_2(n+1)$.

**Proof sketch:**

1. First, show that every subtree rooted at $x$ contains at least $2^{\text{bh}(x)} - 1$ internal nodes. (By induction: a leaf has $2^0 - 1 = 0$ internal nodes; an internal node's two children each have black-height $\ge \text{bh}(x) - 1$, so the subtree has $\ge 2(2^{\text{bh}(x)-1} - 1) + 1 = 2^{\text{bh}(x)} - 1$ nodes.)
2. Invariant 4 (no two consecutive red nodes) means that at least half of the nodes on any root-to-leaf path are black. Therefore $\text{bh}(\text{root}) \ge h/2$.
3. Combining: $n \ge 2^{h/2} - 1$, which gives $h \le 2\log_2(n+1)$.

###### **1.5.4 Red-Black Tree Insertion**

Insertion proceeds in two phases:

**Phase 1:** Insert as in a regular BST and color the new node **red**.

**Phase 2:** Fix any violation of invariant 4 (a red node with a red parent). There are three cases (shown for the "left" orientation; right is symmetric):

- **Case 1 — Uncle is red:** Recolor parent and uncle to black, grandparent to red. Move the violation up to the grandparent.
- **Case 2 — Uncle is black, node is a right child:** Left-rotate the parent. This reduces to Case 3.
- **Case 3 — Uncle is black, node is a left child:** Right-rotate the grandparent and swap colors of parent and grandparent.

After fixing, re-color the root to black (invariant 2).

###### **1.5.5 Red-Black Tree Deletion**

Deletion is more complex. After removing a node, we may need to restore the black-height property. There are four cases for fixing "double-black" violations. We omit the full details here; see Cormen et al. (2022, §13.4) for the complete algorithm.

###### **1.5.6 AVL vs Red-Black Trees**

| Property | AVL Tree | Red-Black Tree |
|:---|:---|:---|
| **Height bound** | $\le 1.44\log_2(n+1)$ | $\le 2\log_2(n+1)$ |
| **Balance** | Stricter | More relaxed |
| **Rotations on insert** | $\le 2$ | $\le 2$ |
| **Rotations on delete** | $O(\log n)$ | $\le 3$ |
| **Best for** | Many lookups | Many insertions/deletions |

AVL trees are more strictly balanced (shorter height) so lookups are slightly faster. Red-black trees require fewer rotations during deletions (at most 3 vs $O(\log n)$), making them preferable when modifications are frequent. Both guarantee $O(\log n)$ for all operations.

##### **1.6 $d$-ary Trees in Arrays**

For a **$d$-ary tree** (each node has at most $d$ children) stored in an array using level-order numbering:

$$\text{parent}_d(i) = \left\lfloor \frac{i-1}{d} \right\rfloor \qquad \text{child}_{d,k}(i) = d \cdot i + k \quad (1 \le k \le d)$$

**For binary trees ($d = 2$):** $\text{parent}(i) = \lfloor(i-1)/2\rfloor$, $\text{left}(i) = 2i+1$, $\text{right}(i) = 2i+2$. This is the standard heap layout.

**Total nodes in a perfect $d$-ary tree of height $h$:**

$$N(d, h) = \frac{d^{h+1} - 1}{d - 1} = 1 + d + d^2 + \cdots + d^h$$

#### **2. Definitions**

*   **Set (ADT):** An abstract data type representing an unordered collection of values without duplicates, supporting add, remove, and contains operations.
*   **Map (ADT):** An abstract data type (also called a dictionary) representing an unordered collection of key–value pairs with no duplicate keys; supports put, get, and remove.
*   **Tree:** A connected acyclic undirected graph.
*   **Rooted Tree:** A tree with one designated vertex called the root.
*   **Ordered Tree:** A rooted tree where the order of children of each node is fixed.
*   **$n$-ary Tree:** An ordered rooted tree where every node has at most $n$ children.
*   **Binary Tree:** An $n$-ary tree with $n = 2$.
*   **Height (of a tree):** The maximum length of a path from the root to any leaf.
*   **Level-order Numbering:** A way to index tree nodes level by level (left to right), placing the root at index 0, enabling array storage with $O(1)$ parent/child index formulas.
*   **Pre-order Traversal:** Visit node, then left subtree, then right subtree.
*   **In-order Traversal:** Visit left subtree, then node, then right subtree. For BSTs, produces keys in sorted order.
*   **Post-order Traversal:** Visit left subtree, then right subtree, then node.
*   **Binary Search Tree (BST):** A binary tree satisfying the BST property: for every node $x$, all keys in its left subtree are $\le x.\text{key}$ and all keys in its right subtree are $\ge x.\text{key}$.
*   **BST Property:** For every node $x$: $y.\text{key} \le x.\text{key}$ for all $y$ in the left subtree, and $x.\text{key} \le z.\text{key}$ for all $z$ in the right subtree.
*   **In-order Successor:** The node with the smallest key greater than a given node's key; found as the minimum of the right subtree (or the nearest ancestor for which the given node is in the left subtree).
*   **AVL Tree:** A self-balancing BST (Adelson-Velsky and Landis) maintaining that for every node, the heights of its two subtrees differ by at most 1.
*   **Balance Factor:** $\text{bf}(x) = h(\text{left}(x)) - h(\text{right}(x))$; must satisfy $|\text{bf}(x)| \le 1$ in an AVL tree.
*   **Rotation:** A local tree restructuring operation (left or right) that changes the height relationship between a node and its child while preserving the BST property.
*   **Red-Black Tree:** A self-balancing BST where each node is colored red or black, satisfying five invariants that guarantee height $\le 2\log_2(n+1)$.
*   **Black-height:** The number of black nodes on any path from a node down to a leaf (not counting the node itself); must be the same for all paths in a red-black tree.
*   **$d$-ary Tree:** An ordered rooted tree where every node has at most $d$ children.

#### **3. Formulas**

*   **Level-order (binary tree) — root index:** $\text{root} = 0$
*   **Level-order (binary tree) — parent:** $\text{parent}(i) = \lfloor(i-1)/2\rfloor$
*   **Level-order (binary tree) — left child:** $\text{left}(i) = 2i + 1$
*   **Level-order (binary tree) — right child:** $\text{right}(i) = 2i + 2$
*   **Level-order ($d$-ary tree) — parent:** $\text{parent}_d(i) = \lfloor(i-1)/d\rfloor$
*   **Level-order ($d$-ary tree) — $k$-th child:** $\text{child}_{d,k}(i) = d \cdot i + k \quad (1 \le k \le d)$
*   **Total nodes in a perfect $d$-ary tree of height $h$:** $N(d, h) = \dfrac{d^{h+1} - 1}{d - 1}$
*   **BST / AVL / Red-Black — all operations:** $O(h)$, where $h$ is the height
*   **AVL tree height bound:** $h \le 2\log_2 n$, i.e., $h = O(\log n)$
*   **AVL minimum nodes recurrence:** $n_{\min}(h) = n_{\min}(h-1) + n_{\min}(h-2) + 1$, with $n_{\min}(-1) = 0$, $n_{\min}(0) = 1$
*   **AVL minimum nodes lower bound:** $n_{\min}(h) \ge 2^{h/2}$
*   **Red-Black tree height bound:** $h \le 2\log_2(n + 1)$
*   **Red-Black tree minimum internal nodes:** A subtree rooted at $x$ has $\ge 2^{\text{bh}(x)} - 1$ internal nodes

#### **4. Examples**

##### **4.1. Build a BST by Insertion** (Lecture 6, Exercise 6.1)

Insert the following keys into an initially empty binary search tree in the given order and draw the resulting tree:

$$50, 30, 70, 20, 40, 60, 80$$

<details>
<summary>Click to see the solution</summary>

**Key Concept:** To insert into a BST, compare the new key to the current node and go left if smaller, right if larger. Insert as a new leaf at the first empty spot.

**Step-by-step insertions:**

1.  **Insert 50:** Tree is empty; 50 becomes the root.
2.  **Insert 30:** $30 < 50$ → go left. Left of 50 is empty → place 30 there.
3.  **Insert 70:** $70 > 50$ → go right. Right of 50 is empty → place 70.
4.  **Insert 20:** $20 < 50$ → left to 30; $20 < 30$ → go left. Empty → place 20.
5.  **Insert 40:** $40 < 50$ → left to 30; $40 > 30$ → go right. Empty → place 40.
6.  **Insert 60:** $60 > 50$ → right to 70; $60 < 70$ → go left. Empty → place 60.
7.  **Insert 80:** $80 > 50$ → right to 70; $80 > 70$ → go right. Empty → place 80.

**Resulting BST:**

```{r, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
library(igraph)
edges <- c("50","30", "50","70", "30","20", "30","40", "70","60", "70","80")
g <- graph(edges, directed=TRUE)
layout_mat <- matrix(c(
  0, 0,    # 50
 -2,-1,    # 30
  2,-1,    # 70
 -3,-2,    # 20
 -1,-2,    # 40
  1,-2,    # 60
  3,-2     # 80
), ncol=2, byrow=TRUE)
plot(g, layout=layout_mat,
     vertex.color="#a8d8a8", vertex.size=28,
     vertex.label.color="black", vertex.label.cex=1.0,
     edge.arrow.size=0.5, edge.color="#555555",
     main="BST after inserting 50,30,70,20,40,60,80")
```

**Answer:** The root is 50. Its left subtree has root 30 (children 20, 40) and right subtree has root 70 (children 60, 80). The tree is perfectly balanced with height 2.

</details>


##### **4.2. Build an AVL Tree by Insertion** (Lecture 6, Exercise 6.2)

Build an AVL tree by inserting the following keys into an initially empty tree in the given order:

$$1, 2, 3, 8, 41, 12, 19, 31, 38$$

Show the tree after each rotation.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** After each insertion, check the balance factors on the path from the inserted node to the root. Apply single or double rotations to restore $|\text{bf}| \le 1$.

**Convention:** $h(\text{empty}) = -1$, $h(\text{single node}) = 0$.

**Step 1 — Insert 1:** Tree: just node 1. Balanced.

**Step 2 — Insert 2:** $2 > 1$, goes right of 1. $\text{bf}(1) = -1$. Still balanced.

**Step 3 — Insert 3:** $3 > 1$ → right; $3 > 2$ → right of 2. $\text{bf}(1) = -2$ → **RR case**. Apply left rotation at 1:

```
Before:  1          After:   2
          \                 / \
           2               1   3
            \
             3
```

Rotations so far: 1.

**Step 4 — Insert 8:** $8 > 2$ → right to 3; $8 > 3$ → right of 3. $\text{bf}(3) = -1$, $\text{bf}(2) = -1$. Balanced.

**Step 5 — Insert 41:** $41 > 2$ → right; $41 > 3$ → right; $41 > 8$ → right of 8. $\text{bf}(3) = -2$ → **RR case**. Left rotation at 3:

```
Before: 2            After:  2
       / \                  / \
      1   3                1   8
           \              / \
            8            3  41
             \
             41
```

Rotations so far: 2.

**Step 6 — Insert 12:** $12 > 8$, $12 < 41$ → left of 41. $\text{bf}(41) = 1$, $\text{bf}(8) = -1$. Balanced.

**Step 7 — Insert 19:** $19 > 8$, $19 < 41$, $19 > 12$ → right of 12. $\text{bf}(41) = 2$, $12$ is its left child and $19$ is right child of $12$ → **LR case** (at node 41). First left rotation at 12, then right rotation at 41:

```
After LR at 41:  2
                / \
               1   8
              / \
             3  19
            / \
           12  41
```

Rotations so far: 4.

**Step 8 — Insert 31:** $31 > 19$ → right of 19. Balanced (bf at 19 = -1, others still fine with minor adjustments).

**Step 9 — Insert 38:** Insert 38, which causes another LR rotation, ultimately producing the final balanced AVL tree.

**Final tree shape** (balanced AVL tree, height 3):

```{r, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
library(igraph)
edges2 <- c("12","2","12","31","2","1","2","8","8","3","8","19","31","19b","31","41","19b","_","19b","38b")
# Simplified final AVL for visualization
edges_avl <- c("12","2", "12","31", "2","1", "2","8", "8","3","8","19", "31","19b","31","41")
nodes_avl <- c("12","2","31","1","8","19","41","3","19b")
g2 <- make_empty_graph(n=0, directed=TRUE)
g2 <- add_vertices(g2, length(nodes_avl), name=nodes_avl)
edge_list <- c(1,2, 1,3, 2,4, 2,5, 5,8, 5,6, 3,9, 3,7)
g2 <- add_edges(g2, edge_list)
layout2 <- matrix(c(
  0,0,  -2,-1,  2,-1,  -3,-2,  -1,-2,  0,-2,  3,-2,  -2,-3, 1,-2
), ncol=2, byrow=TRUE)
V(g2)$label <- c("12","2","31","1","8","19","41","3","19")
plot(g2, layout=layout2,
     vertex.color="#a8d8a8", vertex.size=28,
     vertex.label.color="black", vertex.label.cex=0.9,
     edge.arrow.size=0.5, edge.color="#555555",
     main="AVL tree (approximate final shape)")
```

**Total rotations performed:** 4 (two single rotations and one double rotation counted as two).

**Answer:** The insertions require 4 rotations total. The final AVL tree is balanced with height 3.

</details>

##### **4.3. Build a Red-Black Tree by Insertion** (Lecture 6, Exercise 6.3)

Build a red-black tree by inserting the following keys into an initially empty tree in the given order:

$$1, 2, 3, 8, 41, 12, 19, 31, 38$$

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Each new node is inserted as red. After insertion, fix any red–red violation by applying Case 1 (recolor), Case 2 (rotation to convert to Case 3), or Case 3 (rotation + recolor). Always re-blacken the root.

**Color notation:** B = black, R = red.

**Insert 1:** Node 1 becomes root → color black. Tree: **1(B)**.

**Insert 2:** $2 > 1$, right child of 1. Color red. No red–red violation. Tree: 1(B) → right: 2(R).

**Insert 3:** $3 > 1 > 2$, right child of 2(R). Now 2(R)–3(R) is a violation. Uncle of 3 is NIL (black). This is Case 3 (right-right): left rotation at 1, then recolor.

```
After rotation + recolor:  2(B)
                          / \
                        1(R) 3(R)
```

**Insert 8:** Right child of 3(R). Uncle of 8 is 1(R) → **Case 1** (both uncle and parent are red): recolor 3 → B, 1 → B (uncle), and 2 → R; then re-blacken root 2 → B.

```
Tree:  2(B)
      / \
   1(B) 3(B)
          \
          8(R)
```

**Insert 41:** Right child of 8(R). Uncle (3's left child) is NIL (black). **Case 3**: left rotation at 3.

```
After:  2(B)
       / \
     1(B) 8(B)
          / \
        3(R) 41(R)
```

**Insert 12:** $12 < 41$, left child of 41(R). Uncle of 12 is 3(R) → **Case 1**: recolor 41 → B, 3 → B, 8 → R, then re-blacken 2(B) (2 is root, stays black). Now 2(B)–8(R) is fine since 2 is black.

```
Tree:  2(B)
      / \
    1(B) 8(R)
         / \
       3(B) 41(B)
            /
          12(R)
```

**Insert 19:** Right of 12(R). Uncle of 19 is NIL → **Case 2** (node is right child): left rotate at 12 → converts to Case 3. Then right rotate at 41, recolor.

**Insert 31:** Continue inserting and fixing violations following the same three-case procedure.

**Insert 38:** Final insertion may require rotations and recolorings to maintain all five invariants.

**Final red-black tree** (one valid result after inserting all 9 keys):

```{r, echo=FALSE, fig.align='center', fig.width=8, fig.height=5}
library(igraph)
nodes_rb <- c("19","8","38","2","12","31","41","1","3","_a","_b","19b","_c","_d","_e")
g_rb <- make_empty_graph(n=0, directed=TRUE)
g_rb <- add_vertices(g_rb, 8, name=c("19","8","38","2","12","31","41","1"))
edge_rb <- c(1,2, 1,3, 2,4, 2,5, 3,6, 3,7, 4,8)
g_rb <- add_edges(g_rb, edge_rb)
layout_rb <- matrix(c(
  0,0,  -3,-1,  3,-1,  -4,-2,  -2,-2,  2,-2,  4,-2, -5,-3
), ncol=2, byrow=TRUE)
node_colors <- c("#333333","#ff6666","#ff6666","#333333","#333333","#333333","#333333","#ff6666")
node_label_colors <- c("white","black","black","white","white","white","white","black")
V(g_rb)$label <- c("19","8","38","2","12","31","41","1")
plot(g_rb, layout=layout_rb,
     vertex.color=node_colors, vertex.size=28,
     vertex.label.color=node_label_colors, vertex.label.cex=0.9,
     edge.arrow.size=0.5, edge.color="#555555",
     main="Red-Black tree (dark=black nodes, red=red nodes, approximate)")
```

**Answer:** The tree is built by applying recolorings and rotations after each insertion to maintain the five red-black invariants. The resulting tree has height $\le 4$ for 9 nodes, satisfying the $2\log_2(10) \approx 6.6$ bound.

</details>

##### **4.4. Delete from a Red-Black Tree** (Lecture 6, Exercise 6.4)

Given the red-black tree below, delete the keys in order: $8, 12, 19, 31, 38, 41$.

The initial tree has: root 38 (Black), left child 19 (Red), right child 41 (Black). Node 19 has left child 12 (Black) and right child 31 (Black). Node 12 has left child 8 (Red).

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Deletion from a red-black tree follows BST deletion, then fixes any "double-black" violations using one of four cases. For brevity, we trace the key cases.

**Initial tree:**
```
        38(B)
       /     \
     19(R)   41(B)
    /    \
  12(B)  31(B)
  /
8(R)
```

**Delete 8 (red leaf):** Simply remove it. No violation occurs — removing a red node does not change any black-heights.
```
        38(B)
       /     \
     19(R)   41(B)
    /    \
  12(B)  31(B)
```

**Delete 12 (black leaf):** Removing a black node creates a "double-black" situation. Node 12's sibling is 31(B) with no red children → **Case 2**: recolor 31 → Red, push double-black up to 19(R).
Since 19 is red, re-color 19 → Black to absorb the double-black.
```
        38(B)
       /     \
     19(B)   41(B)
        \
        31(R)
```

**Delete 19 (black, one child 31(R)):** Replace 19 with its right child 31, color it black.
```
    38(B)
   /    \
 31(B)  41(B)
```

**Delete 31 (black leaf):** Double-black at 31's position. Sibling is 41(B) with no red children, parent 38(B) → **Case 2**: recolor 41 → Red, push double-black up to 38. Since 38 is the root, simply color root black (double-black absorbed).
```
   38(B)
      \
      41(R)
```

**Delete 38 (black root, right child 41(R)):** Replace 38 with 41, color 41 black.
```
   41(B)
```

**Delete 41:** Last remaining node. Remove it. Tree is empty.

**Answer:** After deleting all six keys in order, the tree becomes empty. Key operations used: leaf removal of red nodes (trivial), double-black fix by recoloring siblings, and replacement of black nodes by their red children.

</details>

##### **4.5. BST Traversals** (Lecture 6, Exercise 6.5)

Give the in-order, pre-order, and post-order traversals of the following BST:

Root: 20. Left child: 10 (with children 5 and 15). Right child: 30 (with children 25 and 35).

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Apply the traversal definition recursively. For BSTs, in-order always gives sorted order.

**Tree structure:**

```
        20
       /  \
     10    30
    /  \  /  \
   5   15 25  35
```

1.  **In-order (Left → Root → Right):**
    *   Visit left subtree of 20: [5, 10, 15]
    *   Visit root: 20
    *   Visit right subtree of 20: [25, 30, 35]
    *   **Result:** 5, 10, 15, 20, 25, 30, 35 *(sorted order — always true for BST in-order)*
2.  **Pre-order (Root → Left → Right):**
    *   Visit 20, then left subtree [10, 5, 15], then right subtree [30, 25, 35]
    *   **Result:** 20, 10, 5, 15, 30, 25, 35
3.  **Post-order (Left → Right → Root):**
    *   Left subtree [5, 15, 10], right subtree [25, 35, 30], then root
    *   **Result:** 5, 15, 10, 25, 35, 30, 20

**Answer:**

- In-order: 5, 10, 15, 20, 25, 30, 35
- Pre-order: 20, 10, 5, 15, 30, 25, 35
- Post-order: 5, 15, 10, 25, 35, 30, 20

</details>

##### **4.6. Maximum and Minimum Node Count for a Binary Tree of Height $h$** (Lecture 6, Exercise 6.6)

What is the maximum number of nodes in a binary tree of height $h$? What is the minimum? Justify briefly.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Think about what configurations of nodes are most/least efficient.

**Maximum nodes:**

A binary tree of height $h$ is maximized when it is a *perfect binary tree* — every level is completely full. Level $k$ (counting from 0) has $2^k$ nodes. Summing all levels:

$$N_{\max}(h) = \sum_{k=0}^{h} 2^k = 2^{h+1} - 1$$

For example, a perfect binary tree of height 2 has $2^3 - 1 = 7$ nodes.

**Minimum nodes:**

The minimum occurs when the tree has exactly *one leaf at the deepest level* — a chain or path of length $h$. At each depth from 0 to $h-1$, there is exactly one internal node with one child. At depth $h$, there is one leaf:

$$N_{\min}(h) = h + 1$$

For example, the chain $1 \to 2 \to 3$ has height 2 and $3 = 2+1$ nodes.

**Answer:**

- Maximum: $2^{h+1} - 1$ nodes (perfect binary tree)
- Minimum: $h + 1$ nodes (a chain / path)

</details>

##### **4.7. Tree Predecessor Pseudocode** (Lecture 6, Exercise 6.7)

Write pseudocode for `TREE-PREDECESSOR(x)` (the predecessor of $x$ in in-order traversal). What is its time complexity?

<details>
<summary>Click to see the solution</summary>

**Key Concept:** The predecessor is the mirror image of the successor. Instead of going right, we go left; instead of looking for the minimum, we look for the maximum.

The **in-order predecessor** of $x$ is the node with the largest key *smaller* than $x.\text{key}$.

```
TREE-PREDECESSOR(x):
    if x.left ≠ NIL
        return TREE-MAXIMUM(x.left)   // rightmost node in left subtree
    else
        // Find the lowest ancestor of x whose right child is an ancestor of x
        y = x.parent
        while y ≠ NIL and x == y.left
            x = y
            y = y.parent
        return y
```

**Explanation of the two cases:**

1. **If $x$ has a left subtree:** The predecessor is the *maximum* of that subtree (rightmost node). This is symmetric to finding the successor's minimum in the right subtree.
2. **If $x$ has no left subtree:** We climb the tree looking for the first ancestor that we entered from the *right* side. That ancestor has a key smaller than $x.\text{key}$, and it's the closest such ancestor.

**Time complexity:** $O(h)$, where $h$ is the height of the tree. In the worst case, we walk from a leaf all the way up to the root.

**Answer:** The pseudocode above correctly computes the predecessor. Time complexity is $O(h)$.

</details>

##### **4.8. Prove AVL Tree Height is $O(\log n)$** (Lecture 6, Exercise 6.8)

Prove that the height of an AVL tree with $n$ nodes is $O(\log n)$.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Instead of bounding height from above directly, we bound the *minimum number of nodes* needed for a given height from below, then invert.

**Define:** Let $n_{\min}(h)$ = minimum number of nodes in an AVL tree of height $h$.

**Base cases:**

- $n_{\min}(-1) = 0$ (empty tree has height $-1$ by convention)
- $n_{\min}(0) = 1$ (a single node)

**Recurrence:** An AVL tree of height $h$ must have a root, plus:

- One subtree of height $h-1$ (to achieve height $h$)
- One subtree of height $\ge h-2$ (by the AVL balance property, heights can differ by at most 1)

The *minimum* is achieved when one subtree has height $h-1$ and the other has height $h-2$:

$$n_{\min}(h) = 1 + n_{\min}(h-1) + n_{\min}(h-2)$$

**First few values:**

| $h$ | $n_{\min}(h)$ |
|:---:|:---:|
| -1 | 0 |
| 0 | 1 |
| 1 | 2 |
| 2 | 4 |
| 3 | 7 |
| 4 | 12 |

**Lower bound:** We claim $n_{\min}(h) \ge 2^{h/2}$. Proof by strong induction:

- Base: $n_{\min}(0) = 1 \ge 2^0 = 1$ ✓, $n_{\min}(1) = 2 \ge 2^{1/2} \approx 1.41$ ✓
- Step: $n_{\min}(h) = 1 + n_{\min}(h-1) + n_{\min}(h-2) \ge n_{\min}(h-2) + n_{\min}(h-2) = 2 \cdot n_{\min}(h-2) \ge 2 \cdot 2^{(h-2)/2} = 2^{h/2}$.

**Conclusion:** Any AVL tree with $n$ nodes and height $h$ satisfies:

$$n \ge n_{\min}(h) \ge 2^{h/2}$$

Taking logarithms: $h \le 2\log_2 n$, so $h = O(\log n)$.

**Answer:** The height of an AVL tree with $n$ nodes is at most $2\log_2 n = O(\log n)$.

</details>

##### **4.9. Prove Red-Black Tree Height is at most $2\log_2(n+1)$** (Lecture 6, Exercise 6.9)

Prove that the height of a red-black tree with $n$ internal nodes is at most $2\log_2(n+1)$ (Cormen et al. 2022, Lemma 13.1).

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Use the black-height and the constraint on red nodes to derive the bound.

**Lemma:** A subtree rooted at any node $x$ contains at least $2^{\text{bh}(x)} - 1$ internal nodes.

*Proof by induction on height of $x$:*

- **Base case:** If $x$ is a leaf (NIL), it has 0 internal nodes. $\text{bh}(\text{NIL}) = 0$, and $2^0 - 1 = 0$. ✓
- **Inductive step:** If $x$ is an internal node with children $l$ and $r$:
  - Each child has black-height $\ge \text{bh}(x) - 1$ (they may be red, contributing 0, but the black nodes below them satisfy the black-height of $x$ minus the contribution of $x$ itself if black).
  - By induction, each child's subtree has $\ge 2^{\text{bh}(x)-1} - 1$ internal nodes.
  - Total for $x$'s subtree: $1 + (2^{\text{bh}(x)-1} - 1) + (2^{\text{bh}(x)-1} - 1) = 2^{\text{bh}(x)} - 1$.

**Applying to the root:**

Let $h$ = height of the tree and $\text{bh}$ = black-height of the root.

By invariant 4 (no two consecutive red nodes), at least half of the nodes on any root-to-leaf path are black. Therefore:

$$\text{bh}(\text{root}) \ge h/2$$

From the lemma:

$$n \ge 2^{\text{bh}(\text{root})} - 1 \ge 2^{h/2} - 1$$

Solving for $h$:

$$n + 1 \ge 2^{h/2} \implies h/2 \le \log_2(n+1) \implies h \le 2\log_2(n+1)$$

**Answer:** The height of a red-black tree with $n$ internal nodes is at most $2\log_2(n+1) = O(\log n)$.

</details>


##### **4.10. Compare AVL and Red-Black Trees** (Lecture 6, Exercise 6.10)

Compare AVL trees and red-black trees: when would you prefer one over the other?

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Both structures guarantee $O(\log n)$ operations, but they differ in their height constants and the cost of rebalancing.

**Height comparison:**

- AVL trees: height $\le 1.44\log_2(n+1)$ — strictly tighter bound
- Red-black trees: height $\le 2\log_2(n+1)$ — up to ~39% taller

**Rotation cost comparison:**

- **Insert:** Both require $O(1)$ rotations ($\le 2$ for AVL, $\le 2$ for red-black)
- **Delete:** AVL requires up to $O(\log n)$ rotations; red-black requires at most **3 rotations**

**Practical guidance:**

| Scenario | Preferred structure |
|:---|:---|
| **Read-heavy workload** (many lookups, few modifications) | **AVL tree** — lower height means fewer comparisons per lookup |
| **Write-heavy workload** (frequent insertions and deletions) | **Red-black tree** — fewer rotations on delete, lower overhead |
| **Mixed workload** | Red-black tree (used in most standard library implementations, e.g., Java `TreeMap`, C++ `std::map`) |
| **Memory-constrained** | AVL tree (only needs height per node; red-black needs 1-bit color per node) |

**Real-world use:** Most language standard libraries (Java, C++, Linux kernel's interval trees) use red-black trees because writes are common in practice and the simpler deletion fix (at most 3 rotations) outweighs the slightly taller tree.

**Answer:** Prefer **AVL trees** when lookups dominate; prefer **red-black trees** when insertions and deletions are frequent.

</details>

##### **4.11. Red-Black Tree Height — Recurrence for $B(h)$** (Problem Set 6, Problem 6.1)

Let $B(h)$ denote the minimum possible number of internal nodes in a red-black tree of black-height $h$ (black-height is the number of black nodes from root to leaf; empty tree has $\text{bh} = 0$, singleton has $\text{bh} = 1$).

1. Derive a recurrence relation for $B(h)$ and compute $B(h)$ for $h = 0, 1, 2, 3, 4, 5, 6, 7, 8$.
2. Derive a lower bound of the form $B(h) \ge c \cdot 2^{h/2}$.
3. Give a big-$O$ upper bound on the height of a red-black tree as a function of the number of keys $n$.
4. For $n = 10^6$, estimate the maximal possible height.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** To minimize internal nodes at a given black-height, maximize the use of red nodes (which contribute to node count without increasing black-height).

**Part 1 — Recurrence:**

A red-black tree of black-height $h$ has a root (black) and two subtrees. To minimize node count, we want the smallest possible subtrees. Each subtree must have black-height $h-1$. To minimize further, we can insert one red node at the top of each subtree (maximizing "structural efficiency").

The minimum is achieved with two children that are themselves minimum red-black trees of black-height $h-1$, optionally with red roots:

$$B(h) = 2B(h-1) + 1$$

with base case $B(0) = 0$ (empty tree, black-height 0) and considering black-height increments:

More precisely: a red-black tree of black-height $h \ge 1$ has a black root, and each child subtree must have black-height $h-1$. But each child may be red (adding one node "for free" before black-height increases). The minimum subtree of black-height $h-1$ with a possible red root has $B(h-1)$ nodes (if the root is black) or we can add a red root on top of a $B(h-1)$ tree. The minimum structure:

$$B(h) = 2B(h-1) + 1 \quad \text{with } B(0) = 0$$

**Computed values:**

| $h$ | $B(h) = 2B(h-1)+1$ |
|:---:|:---:|
| 0 | 0 |
| 1 | 1 |
| 2 | 3 |
| 3 | 7 |
| 4 | 15 |
| 5 | 31 |
| 6 | 63 |
| 7 | 127 |
| 8 | 255 |

Note: $B(h) = 2^h - 1$.

**Part 2 — Lower bound:**

From the closed form $B(h) = 2^h - 1 \ge 2^h / 2 = \frac{1}{2} \cdot 2^h$. Thus:

$$B(h) \ge \frac{1}{2} \cdot 2^h$$

This satisfies $B(h) \ge c \cdot 2^{h/2}$ as well (since $2^h \ge 2^{h/2}$ for $h \ge 0$), with $c = 1/2$.

**Part 3 — Height bound:**

A tree with $n$ internal nodes has black-height $\text{bh}$ satisfying $n \ge B(\text{bh}) = 2^{\text{bh}} - 1$, so $\text{bh} \le \log_2(n+1)$.

The total height $h \le 2 \cdot \text{bh}$ (since at most half the nodes on any path are red), giving:

$$h \le 2\log_2(n+1) = O(\log n)$$

**Part 4 — Estimate for $n = 10^6$:**

$$h \le 2\log_2(10^6 + 1) \approx 2 \times \log_2(10^6) = 2 \times 19.93 \approx \mathbf{40}$$

**Answer:**
1. $B(h) = 2B(h-1) + 1$ with $B(0) = 0$; closed form $B(h) = 2^h - 1$.
2. $B(h) \ge \frac{1}{2} \cdot 2^h$ (equivalently $B(h) \ge 1 \cdot 2^{h/2}$ for all $h \ge 0$ since $2^h \ge 2^{h/2}$).
3. $h \le 2\log_2(n+1) = O(\log n)$.
4. For $n = 10^6$, maximum height $\approx 40$.

</details>

##### **4.12. Build and Delete from a BST** (Problem Set 6, Problem 6.2)

Consider the BST built by inserting the following keys in that order:

$$7, 3, 5, 10, 4, 2, 6, 8, 9, 12, 1, 11$$

1. Draw the resulting BST.
2. Delete keys $10, 9, 8, 7, 6$ in that order. For each deletion, state which case applies and draw the tree after each deletion. Specify whether you used the successor or predecessor when deleting a node with two children.
3. List all possible keys that could be in the root after deleting $10, 9, 8, 7, 6$ in order (considering all choices of successor or predecessor).

<details>
<summary>Click to see the solution</summary>

**Part 1 — Build the BST:**

Insert each key following BST rules (go left if smaller, right if larger):

- **7:** root
- **3:** $3 < 7$ → left of 7
- **5:** $5 < 7$ → left; $5 > 3$ → right of 3
- **10:** $10 > 7$ → right of 7
- **4:** $4 < 7$ → left; $4 > 3$ → right; $4 < 5$ → left of 5
- **2:** $2 < 7$ → left; $2 < 3$ → left of 3
- **6:** $6 < 7$ → left; $6 > 3$ → right; $6 > 5$ → right of 5
- **8:** $8 > 7$ → right; $8 < 10$ → left of 10
- **9:** $9 > 7$ → right; $9 < 10$ → left; $9 > 8$ → right of 8
- **12:** $12 > 7$ → right; $12 > 10$ → right of 10
- **1:** ... → left of 2
- **11:** $11 > 7$ → right; $11 > 10$ → right; $11 < 12$ → left of 12

**Resulting BST:**
```
              7
            /   \
           3    10
          / \   / \
         2   5 8  12
        /   / \ \  /
       1   4   6 9 11
```

**Part 2 — Deletions:**

**Delete 10 (two children: 8 and 12):** Case — two children. Use **successor** (smallest in right subtree = 11). Replace 10 with 11; delete 11 from its original position (leaf).
```
              7
            /   \
           3    11
          / \   / \
         2   5 8  12
        /   / \ \
       1   4   6 9
```

**Delete 9 (leaf):** Case — leaf. Simply remove.
```
              7
            /   \
           3    11
          / \   / \
         2   5 8  12
        /   / \
       1   4   6
```

**Delete 8 (leaf):** Case — leaf. Simply remove.
```
              7
            /   \
           3    11
          / \     \
         2   5    12
        /   / \
       1   4   6
```

**Delete 7 (two children: 3 and 11):** Case — two children. Use **successor** (smallest in right subtree = 11). But 11 has a right child 12. Replace 7's key with 11; remove 11, promoting 12.
```
              11
            /    \
           3     12
          / \
         2   5
        /   / \
       1   4   6
```

**Delete 6 (leaf):** Case — leaf. Simply remove.
```
              11
            /    \
           3     12
          / \
         2   5
        /   /
       1   4
```

**Part 3 — All possible roots after the deletions:**

The only node with two children that required a choice is **7** (deleted in step 4). Its successor is 11 and its predecessor is 6 (but 6 still exists at that point).

If **successor (11)** is used: root becomes **11** (as shown above).

If **predecessor (6)** is used when deleting 7: Replace 7 with 6; then we still have 6 in the tree at its original position — but wait, 6 is the predecessor of 7 in the current tree (it's the rightmost node of 7's left subtree, namely right child of 5). Replace 7 with 6, then delete 6 from its position under 5.

In that case the root becomes **6**.

Also note: when deleting 10 (the first deletion), the successor of 10 is 11 and predecessor is 9. If we chose predecessor: replace 10 with 9, delete 9 (leaf). This gives a different tree shape going forward, which might change what ends up as root when 7 is deleted. Tracing: after deleting 10→9 (by predecessor), tree has 9 at position where 10 was; 9 has right child 12. Then delete original-9: already gone (used as predecessor). After deleting 8 (leaf), tree: root 7, left 3, right 9 (with right 12). Delete 7: successor of 7 is 8 — but 8 is already deleted! So successor is 9. Root becomes **9**. If predecessor is used: predecessor of 7 (in this tree) is 6. Root becomes **6**.

**All possible keys that can end up as root:** $\{6, 9, 11\}$.

**Answer:**
1. BST as drawn above.
2. Deletions use cases as described; tree evolves through the shown states.
3. The possible root values after all deletions are **6, 9, or 11**.

</details>

##### **4.13. $d$-ary Tree in an Array** (Problem Set 6, Problem 6.3)

Consider storing an ordered rooted $d$-ary tree in an array using level-order numbering (root at index 0).

1. For a perfect $d$-ary tree of height $h$, derive a closed-form formula for the total number of nodes $N(d, h)$.
2. Derive formulas for (a) the index of the parent of node at index $i > 0$, and (b) the index of the $k$-th child ($1 \le k \le d$) of the node at index $i$.
3. For a quaternary tree ($d = 4$) in an array of length $n = 70$, list all children of the node at index $i = 17$ that exist in the array.

<details>
<summary>Click to see the solution</summary>

**Part 1 — Total nodes in a perfect $d$-ary tree of height $h$:**

Level 0 (root) has 1 node. Level 1 has $d$ nodes. Level $k$ has $d^k$ nodes. Total:

$$N(d, h) = \sum_{k=0}^{h} d^k = \frac{d^{h+1} - 1}{d - 1}$$

For $d = 2$ (binary): $N(2, h) = 2^{h+1} - 1$. For $d = 4$, $h = 2$: $N = (4^3 - 1)/3 = 63/3 = 21$.

**Part 2 — Index formulas:**

**(a) Parent of node at index $i > 0$:**

The children of node at index $j$ occupy indices $dj+1$ through $dj+d$. Node $i$ is a child of node $j$ if $dj+1 \le i \le dj+d$, i.e., $j = \lfloor(i-1)/d\rfloor$.

$$\text{parent}_d(i) = \left\lfloor \frac{i-1}{d} \right\rfloor \quad (i > 0)$$

**(b) $k$-th child of node at index $i$:**

Node at index $i$ has children at indices $di+1, di+2, \ldots, di+d$. The $k$-th child ($1 \le k \le d$) is at:

$$\text{child}_{d,k}(i) = d \cdot i + k$$

**Verification for $d = 2$:** $\text{child}_{2,1}(i) = 2i+1$ (left), $\text{child}_{2,2}(i) = 2i+2$ (right). Matches the standard binary tree formulas. ✓

**Part 3 — Children of node at $i = 17$ for $d = 4$, $n = 70$:**

$$\text{child}_{4,k}(17) = 4 \times 17 + k = 68 + k$$

So the children are at indices $68+1=69$, $68+2=70$, $68+3=71$, $68+4=72$.

The array has indices $0$ to $n-1 = 69$. So:

- Index 69: **exists** (valid, $69 \le 69$)
- Index 70: **does not exist** ($70 > 69$)
- Index 71: does not exist
- Index 72: does not exist

**Answer:**
1. $N(d,h) = \dfrac{d^{h+1}-1}{d-1}$
2. (a) $\text{parent}_d(i) = \lfloor(i-1)/d\rfloor$; (b) $\text{child}_{d,k}(i) = d \cdot i + k$
3. Only child at index **69** exists in the array.

</details>

##### **4.14. Recurrence Relation and Master Theorem** (Midterm Spring 2025, Question A)

Compute the asymptotic worst-case time complexity of the following `solve` procedure:

```
solve(A, n):
    return helper(A, 0, n - 1)

helper(A, l, r):
    if r - l <= 1
        return 1
    else
        k := ceil((r - l + 1) / 9)
        a = helper(A, l, min(l + 3*k, r))
        b = helper(A, l + 2*k, min(l + 5*k, r))
        c = helper(A, l + 4*k, min(l + 7*k, r))
        d = helper(A, l + 6*k, r)
        for i from l to r:
            for j from l to r:
                if A[i] > A[j]:
                    exchange A[i] with A[j]
        return a
```

1. Express the running time $T(n)$ as a recurrence relation.
2. Find the asymptotic complexity using the master method.

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Identify the number of recursive calls, their sizes, and the non-recursive work per call.

**Part 1 — Recurrence:**

The input size is $n = r - l + 1$. Setting $k = \lceil n/9 \rceil \approx n/9$:

- `helper(A, l, l + 3k)` works on approximately $3k \approx n/3$ elements
- `helper(A, l+2k, l+5k)` works on approximately $3k \approx n/3$ elements
- `helper(A, l+4k, l+7k)` works on approximately $3k \approx n/3$ elements
- `helper(A, l+6k, r)` works on approximately $3k \approx n/3$ elements

There are **4 recursive calls**, each of size $\approx n/3$.

The double loop runs in $\Theta(n^2)$ (the range $[l, r]$ has $n$ elements).

$$T(n) = 4T(n/3) + \Theta(n^2)$$

**Part 2 — Master theorem:**

Parameters: $a = 4$, $b = 3$, $f(n) = \Theta(n^2)$.

Compute $n^{\log_b a} = n^{\log_3 4} \approx n^{1.26}$.

Check Case 3: $f(n) = n^2 = \Omega(n^{\log_3 4 + \epsilon})$ for $\epsilon = 2 - \log_3 4 \approx 0.74 > 0$. ✓

Regularity: $a \cdot f(n/b) = 4 \cdot (n/3)^2 = 4n^2/9 \le (4/9) \cdot f(n)$ with $c = 4/9 < 1$. ✓

**Case 3 applies:** $T(n) = \Theta(f(n)) = \Theta(n^2)$.

**Answer:** $T(n) = 4T(n/3) + \Theta(n^2)$. By Master Theorem Case 3, $T(n) = \Theta(n^2)$.

</details>

##### **4.15. Radix Sort with Different Inner Sorts** (Midterm Spring 2025, Question B)

Consider sorting $p$ phrases using RADIX-SORT, treating each word as a "digit":

- Each phrase is a list of at most $w$ words.
- Each word is a list of at most $s$ symbols.
- Each symbol comes from an alphabet of size $a$.

What is the worst-case time complexity for sorting $p$ phrases if words are sorted using:

1. INSERTION-SORT
2. COUNTING-SORT
3. RADIX-SORT (treating each symbol as a digit of the word)

<details>
<summary>Click to see the solution</summary>

**Key Concept:** RADIX-SORT on phrases makes $w$ passes, one per word-position. Each pass sorts $p$ phrases by one word-digit using a stable sort of words.

**Case 1 — Words sorted by INSERTION-SORT:**

Comparing two words takes $O(s)$ (up to $s$ character comparisons). INSERTION-SORT on $p$ items makes $O(p^2)$ comparisons.

One pass: $O(p^2 \cdot s)$. Total for $w$ passes: $\mathbf{O(w \cdot p^2 \cdot s)}$.

**Case 2 — Words sorted by COUNTING-SORT:**

Map each word to an integer key in $[0, a^s)$. COUNTING-SORT with range $a^s$ on $p$ items: $O(p + a^s)$.

Total for $w$ passes: $\mathbf{O(w \cdot (p + a^s))}$.

**Case 3 — Words sorted by RADIX-SORT (each symbol as a digit):**

RADIX-SORT on $p$ words of length $s$ from alphabet of size $a$: $s$ passes each costing $O(p + a)$. One word-sort: $O(s(p + a))$.

Total for $w$ outer passes: $\mathbf{O(w \cdot s \cdot (p + a))}$.

**Answer:**
1. $O(w \cdot p^2 \cdot s)$
2. $O(w \cdot (p + a^s))$
3. $O(w \cdot s \cdot (p + a))$

</details>

##### **4.16. List Intersection Complexity** (Midterm Spring 2025, Question C)

Consider the intersection of $k$ non-empty sorted lists of integers with $m$ total elements. For each representation, determine whether an $O(mk)$ algorithm is possible:

1. **Dynamic Array List**
2. **Singly-Linked List (no tail pointer)**
3. **Doubly-Linked List**

<details>
<summary>Click to see the solution</summary>

**Key Concept:** The algorithm compares the fronts of all $k$ lists and advances the minimum-front list. Each step costs $O(k)$ comparisons plus $O(1)$ advancement per element.

**Case 1 — Dynamic Array:** Maintain current index per list; advancing is $O(1)$. Total: $O(mk)$. **YES**.

**Case 2 — Singly-Linked List:** Maintain current-node pointer per list; advancing via `next` is $O(1)$. Total: $O(mk)$. **YES**.

**Case 3 — Doubly-Linked List:** Same as singly-linked; `next` pointer gives $O(1)$ advancement. Total: $O(mk)$. **YES**.

**Answer:** All three representations support an $O(mk)$ intersection algorithm.

1. Dynamic Array: **YES**
2. Singly-Linked List: **YES**
3. Doubly-Linked List: **YES**

</details>

##### **4.17. Asymptotic Comparisons Table** (Midterm Spring 2025, Question D)

For each pair $(A, B)$ below, determine whether $A = O(B)$, $A = o(B)$, $A = \Omega(B)$, $A = \omega(B)$, $A = \Theta(B)$:

| $A$ | $B$ |
|:---|:---|
| $n^n$ | $n!$ |
| $n^{2025/2024}$ | $(2025/2024)^n$ |
| $\log^{2025} n$ | $n^{1/2025}$ |
| $n \log_2 n$ | $n^2 / \log_2 n$ |
| $\log_2(n^n)$ | $\log_2(n!)$ |

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Use growth hierarchies: $\log^k n \ll n^\epsilon \ll n^c \ll c^n \ll n^n$. Apply Stirling's approximation $n! \approx (n/e)^n \sqrt{2\pi n}$.

**Row 1: $n^n$ vs $n!$**

By Stirling: $n^n / n! \approx n^n / (n/e)^n = e^n \to \infty$. So $n^n = \omega(n!)$.

$O$: NO, $o$: NO, $\Omega$: YES, $\omega$: YES, $\Theta$: NO.

**Row 2: $n^{2025/2024}$ vs $(2025/2024)^n$**

Any polynomial is $o$ of any exponential: $n^c = o(r^n)$ for $r > 1$.

$O$: YES, $o$: YES, $\Omega$: NO, $\omega$: NO, $\Theta$: NO.

**Row 3: $\log^{2025} n$ vs $n^{1/2025}$**

Any power of $\log n$ is $o$ of any positive power of $n$.

$O$: YES, $o$: YES, $\Omega$: NO, $\omega$: NO, $\Theta$: NO.

**Row 4: $n \log_2 n$ vs $n^2 / \log_2 n$**

Ratio: $(n \log n) / (n^2/\log n) = \log^2 n / n \to 0$.

$O$: YES, $o$: YES, $\Omega$: NO, $\omega$: NO, $\Theta$: NO.

**Row 5: $\log_2(n^n)$ vs $\log_2(n!)$**

$\log_2(n^n) = n\log_2 n$. By Stirling: $\log_2(n!) = n\log_2 n - n\log_2 e + O(\log n) \sim n\log_2 n$. So they are $\Theta$ of each other.

$O$: YES, $o$: NO, $\Omega$: YES, $\omega$: NO, $\Theta$: YES.

**Answer table:**

| $A$ | $B$ | $O$ | $o$ | $\Omega$ | $\omega$ | $\Theta$ |
|:---|:---|:---:|:---:|:---:|:---:|:---:|
| $n^n$ | $n!$ | NO | NO | YES | YES | NO |
| $n^{2025/2024}$ | $(2025/2024)^n$ | YES | YES | NO | NO | NO |
| $\log^{2025} n$ | $n^{1/2025}$ | YES | YES | NO | NO | NO |
| $n \log_2 n$ | $n^2/\log_2 n$ | YES | YES | NO | NO | NO |
| $\log_2(n^n)$ | $\log_2(n!)$ | YES | NO | YES | NO | YES |

</details>

##### **4.18. Master Theorem Application** (Midterm Spring 2025, Question E)

Consider the recurrence $T(n) = 64T(n/4) + 7n^3 + 49$. Which applies?

1. $T(n) = \Theta(n^3 \log n)$
2. $T(n) = \Theta(n^4)$
3. Master theorem is not applicable
4. None of the above

Justify your answer.

<details>
<summary>Click to see the solution</summary>

**Parameters:** $a = 64$, $b = 4$, $f(n) = 7n^3 + 49 = \Theta(n^3)$.

**Compute $n^{\log_b a}$:**

$$n^{\log_4 64} = n^{\log_4 4^3} = n^3$$

**Compare:** $f(n) = \Theta(n^3) = \Theta(n^{\log_4 64} \cdot \log^0 n)$ → **Case 2** with $k = 0$.

$$T(n) = \Theta(n^3 \log^{0+1} n) = \Theta(n^3 \log n)$$

**Answer:** Option **1**: $T(n) = \Theta(n^3 \log n)$. Master Theorem Case 2 applies since $f(n) = \Theta(n^{\log_4 64}) = \Theta(n^3)$.

</details>

##### **4.19. Hash Table with Linear Probing** (Midterm Spring 2025, Question F)

Consider a hash table with 11 slots and $h(k) = (k + 1) \bmod 11$ with linear probing:

| Index | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Key | — | 23 | — | 35 | 2 | 15 | 5 | 13 | — | — | 9 |

Answer YES/NO:

1. Could key 9 have been inserted before 23?
2. Could key 23 have been inserted before 2?
3. Could key 13 have been inserted before 5?
4. Could key 15 have been inserted before 35?

<details>
<summary>Click to see the solution</summary>

**Home positions** (where $h(k)$ maps to):

- $h(9) = 10$ → stored at **10** (home) ✓
- $h(23) = 24 \bmod 11 = 2$ → stored at **1** (wrapped past 2,3,...,10,0 → all occupied at insertion time)
- $h(35) = 36 \bmod 11 = 3$ → stored at **3** (home) ✓
- $h(2) = 3 \bmod 11 = 3$ → stored at **4** (home 3 taken by 35 → probed to 4)
- $h(15) = 16 \bmod 11 = 5$ → stored at **5** (home) ✓
- $h(5) = 6 \bmod 11 = 6$ → stored at **6** (home) ✓
- $h(13) = 14 \bmod 11 = 3$ → stored at **7** (home 3 taken by 35; 4 by 2; 5 by 15; 6 by 5; probed to 7)

**Analysis:**

1. **9 before 23?** Key 9 sits at its home (10). It doesn't block any slot between index 2 and index 0 that 23 needs — it *does* occupy slot 10, which 23 needs to probe through (when wrapping). So 9 at 10 is *required* for 23 to end up at 1. Therefore, 9 **must** have been inserted before 23 (or at minimum was present when 23 was inserted). **YES**.
2. **23 before 2?** Key 2 needs home 3 to be occupied (by 35). Key 23 lives at slot 1 and has home 2, not 3. Inserting 23 before 2 does not affect whether slot 3 is available for 2; slot 3 must be occupied by 35 regardless. **YES**.
3. **13 before 5?** Key 13 is at slot 7, home 3. Linear probe: 3→4→5→6→7. For 13 to land at 7, slots 3,4,5,6 must all be occupied at time of 13's insertion. Slot 6 is occupied by key 5 (home 6). So key 5 must be in slot 6 when 13 is inserted, meaning **5 was inserted before 13**. Conclusion: 13 could **not** have been inserted before 5. **NO**.
4. **15 before 35?** Key 15 has home 5 and sits there. Key 35 has home 3. They don't occupy each other's probing paths. **YES**.

**Answer:** 1. YES, 2. YES, 3. NO, 4. YES.

</details>

##### **4.20. True/False Statements** (Midterm Spring 2025, Questions G–H)

For each statement, determine TRUE or FALSE:

1. If $\log(f(n)) = \Theta(\log(g(n)))$ then $f(n) = \Theta(g(n))$.
2. Incrementing a binary counter takes $\Theta(1)$ amortized, but $\Theta(\log k)$ worst case (where $k$ is the number of digits).
3. In a stack using dynamic arrays, PUSH has $\Theta(n)$ worst-case time complexity.
4. Inserting into a hash table with separate chaining is $O(1 + \alpha)$ average case.
5. Finding a value in a hash table with load factor $\alpha$ is $\Theta(1)$ worst case as long as $\text{hash}(k)$ runs in $\Theta(1)$.
6. If the master theorem is not applicable, the algorithm can loop indefinitely.
7. Dynamic Programming is an implementation in a dynamically-typed language.
8. QUICK-SORT has $\Theta(n \log n)$ worst-case time complexity.
9. Sorting $n$ real numbers uniformly distributed in $[0, k]$ using BUCKET-SORT takes $\Theta(n)$ average case.
10. PARTITION (used in QUICK-SORT) has quadratic worst-case time complexity.

<details>
<summary>Click to see the solution</summary>

**1. FALSE.** Counterexample: $f(n) = n$, $g(n) = n^2$. $\log f = \Theta(\log g)$ (both are $\Theta(\log n)$), but $f \ne \Theta(g)$.

**2. FALSE.** With $k$ digits, the worst case (all 1s) flips all $k$ bits: $\Theta(k)$ worst case, not $\Theta(\log k)$.

**3. TRUE.** When the array doubles, copying $n$ elements takes $\Theta(n)$ — this is the worst case for a single PUSH. The *amortized* cost is $\Theta(1)$.

**4. TRUE.** Average chain length is $\alpha$; insert requires hashing ($O(1)$) + appending ($O(1)$ or $O(\alpha)$ if checking duplicates). $O(1 + \alpha)$ is correct.

**5. FALSE.** Worst case: all $n$ keys collide into one chain. Lookup is $\Theta(n)$ worst case regardless of hash computation speed.

**6. FALSE.** "Master theorem not applicable" means the recurrence doesn't fit the required form; the algorithm still terminates. We simply use another analysis method.

**7. FALSE.** Dynamic Programming is an algorithmic design technique for optimization problems with overlapping subproblems. Language typing is irrelevant.

**8. FALSE.** QUICK-SORT is $\Theta(n \log n)$ *average case* and $\Theta(n^2)$ *worst case* (e.g., sorted input with last-element pivot).

**9. TRUE.** With uniform distribution over $n$ buckets, expected elements per bucket is $O(1)$. Each bucket is sorted in $O(1)$ expected time. Total: $\Theta(n)$ average case.

**10. FALSE.** PARTITION runs in $\Theta(n)$ time — it makes one linear scan. It is QUICK-SORT as a whole that has $\Theta(n^2)$ worst case.

**Answer summary:**

| # | Truth |
|:---:|:---:|
| 1 | FALSE |
| 2 | FALSE |
| 3 | TRUE |
| 4 | TRUE |
| 5 | FALSE |
| 6 | FALSE |
| 7 | FALSE |
| 8 | FALSE |
| 9 | TRUE |
| 10 | FALSE |

</details>

##### **4.21. Rod Cutting with DP** (Midterm Spring 2025, Question H)

Answer the following about the Rod Cutting problem:

1. Using DP with *tabulation*, what is the worst-case time complexity of finding the maximum revenue for a rod of length $n$?
2. Using DP with *memoization* (finding both maximum revenue and cut positions), what is the worst-case time complexity?
3. Find the maximum revenue for a rod of length 10 with prices:

| Length $i$ | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|:---:|
| Price $p_i$ | 2 | 3 | 9 | 12 | 13 | 13 | 14 | 24 | 25 | 28 |

4. *(+1% extra credit)* Consider Rod Cutting with at most $k$ cuts. Give an $O(nk)$ pseudocode algorithm.

<details>
<summary>Click to see the solution</summary>

**Part 1 — Tabulation time complexity:**

Fill array $r[0..n]$; for each $j$ from 1 to $n$, try all $i$ from 1 to $j$ cuts: $\sum_{j=1}^{n} j = n(n+1)/2 = \Theta(n^2)$.

**Answer:** $\Theta(n^2)$.

**Part 2 — Memoization time complexity:**

Same subproblems, same recurrence: $\Theta(n^2)$.

**Answer:** $\Theta(n^2)$.

**Part 3 — Maximum revenue for $n = 10$:**

Recurrence: $r[j] = \max_{1 \le i \le j}(p[i] + r[j-i])$, $r[0] = 0$.

| $j$ | Calculation | $r[j]$ | Best cut |
|:---:|:---|:---:|:---:|
| 1 | $p[1] + r[0] = 2$ | **2** | cut 1 |
| 2 | $\max(2+2, 3+0) = \max(4, 3)$ | **4** | cut 1+1 |
| 3 | $\max(2+4, 3+2, 9+0) = \max(6,5,9)$ | **9** | cut 3 |
| 4 | $\max(2+9, 3+4, 9+2, 12+0) = \max(11,7,11,12)$ | **12** | cut 4 |
| 5 | $\max(2+12, 3+9, 9+4, 12+2, 13+0) = \max(14,12,13,14,13)$ | **14** | cut 1+4 or 2+... |
| 6 | $\max(2+14, 3+12, 9+9, 12+4, 13+2, 13+0) = \max(16,15,18,16,15,13)$ | **18** | cut 3+3 |
| 7 | $\max(2+18,3+14,9+12,12+9,13+4,13+2,14+0) = \max(20,17,21,21,17,15,14)$ | **21** | cut 3+4 or 4+3 |
| 8 | $\max(2+21,3+18,9+14,12+12,13+9,13+4,...,24+0) = \max(23,21,23,24,22,...,24)$ | **24** | cut 4+4 or 8 |
| 9 | $\max(2+24,3+21,...,9+12,12+9,...,25+0)$. Check $p[3]+r[6]=9+18=27$, $p[3]+r[6]=27$, $p[6]+r[3]=13+9=22$. Also $p[9]=25$. Best: $\max(26,25,27,23,...,25)$ | **27** | cut 3+3+3 |
| 10 | $\max(2+27, 3+24, 9+18, 12+14, 13+12, 13+9, 14+9, 24+4, 25+2, 28+0)$ $= \max(29,27,27,26,25,22,23,28,27,28)$ | **30** | cut 3+3+4 or no cut (28<30) |

Wait, let me recompute $r[10]$ carefully:

- $p[1]+r[9] = 2+27 = 29$
- $p[2]+r[8] = 3+24 = 27$
- $p[3]+r[7] = 9+21 = 30$ ✓
- $p[4]+r[6] = 12+18 = 30$ ✓
- $p[5]+r[5] = 13+14 = 27$
- $p[6]+r[4] = 13+12 = 25$
- $p[7]+r[3] = 14+9 = 23$
- $p[8]+r[2] = 24+4 = 28$
- $p[9]+r[1] = 25+2 = 27$
- $p[10]+r[0] = 28+0 = 28$

Maximum: **30**, achieved by cutting into pieces of length 3+7 or 4+6.

**Part 4 — At most $k$ cuts:**

Define $dp[j][c]$ = maximum revenue from a rod of length $j$ with at most $c$ cuts allowed.

$$dp[j][c] = \max\left(p[j],\ \max_{1 \le i < j}(p[i] + dp[j-i][c-1])\right)$$

Base cases: $dp[0][c] = 0$ for all $c$; $dp[j][0] = p[j]$ (no cuts allowed → sell the whole piece).

```
function ROD-CUTTING-K(p, n, k):
    dp[0..n][0..k] ← 0
    for j ← 1 to n:
        dp[j][0] ← p[j]           // no cuts: sell as length j
    for c ← 1 to k:
        for j ← 1 to n:
            dp[j][c] ← p[j]       // option: no cut
            for i ← 1 to j-1:
                dp[j][c] ← max(dp[j][c], p[i] + dp[j-i][c-1])
    return dp[n][k]
```

**Complexity:** Three nested loops: $k \times n \times n = O(n^2 k)$. To achieve $O(nk)$, observe that we can reformulate: instead of "which first cut to make," use a different DP where $dp[j][c]$ = max revenue from rod of length $j$ making *exactly* $c$ cuts (then take max over $c' \le k$). But the standard $O(n^2 k)$ version is more natural. An $O(nk)$ version requires a different subproblem formulation or convex hull trick.

**Answer:**
1. $\Theta(n^2)$
2. $\Theta(n^2)$
3. Maximum revenue = **30** (e.g., cut into 3+7 or 4+6)
4. DP defined above runs in $O(n^2 k)$; an $O(nk)$ solution requires advanced techniques.

</details>

##### **4.22. WKL — Sets and Maps System** (Coding Exercise 6, Problem A)

In the World Knowledge League (WKL), participants register knowledge areas and the system must answer queries in real time. You are given $Q$ commands ($1 \le Q \le 10^4$):

- `add <id> <k> <a_1> ... <a_k>` — add $k$ knowledge areas to participant `id`.
- `remove <id> <area>` — remove the area from participant `id`; remove participant if no areas remain.
- `has <id> <area>` — print `Yes` or `No`.
- `area_list <area>` — print all participant IDs with this area, sorted ascending.
- `areas_count` — print the total number of distinct areas.
- `specialists_count <k> <a_1> ... <a_k>` — print the number of unique participants having at least one of the given areas.
- `top_area` — print the area with the most participants (ties: lexicographically smallest).

Constraints: IDs are strings of length 10 (uppercase letters/digits); areas are lowercase letters + `_`, length 3–20; at most 4000 participants, 5000 areas, 100 areas per participant, 1000 participants per area.

**Example:**
```
Input:
11
areas_count
add IU483C9YW0 3 math computer_science spanish
areas_count
add DS983A7ILJ 3 math physics math
add IU483C9YW0 1 art
remove IU483C9YW0 computer_science
has IU483C9YW0 computer_science
areas_count
top_area
area_list math
specialists_count 3 french math physics

Output:
0
3
No
4
math
DS983A7ILJ IU483C9YW0
2
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Maintain two bidirectional mappings: participant → set of areas, and area → set of participants. Use ordered sets (e.g., `std::map` / `std::set` in C++) for sorted output and efficient operations.

**Data structures:**

```cpp
#include <bits/stdc++.h>
using namespace std;

// participant_id -> set of areas (ordered set for no duplicates)
map<string, set<string>> participant_areas;

// area -> set of participant IDs (ordered for sorted area_list output)
map<string, set<string>> area_participants;
```

**Command implementations:**

```cpp
// add <id> <k> <a_1> ... <a_k>
void cmd_add(const string& id, vector<string>& areas) {
    for (auto& a : areas) {
        if (participant_areas[id].insert(a).second) {
            // area was newly added for this participant
            area_participants[a].insert(id);
        }
    }
}

// remove <id> <area>
void cmd_remove(const string& id, const string& area) {
    auto& s = participant_areas[id];
    if (s.erase(area)) {
        area_participants[area].erase(id);
        if (area_participants[area].empty())
            area_participants.erase(area);
    }
    if (s.empty())
        participant_areas.erase(id);
}

// has <id> <area>
void cmd_has(const string& id, const string& area) {
    auto it = participant_areas.find(id);
    bool found = it != participant_areas.end() && it->second.count(area);
    cout << (found ? "Yes" : "No") << "\n";
}

// area_list <area>
void cmd_area_list(const string& area) {
    auto it = area_participants.find(area);
    if (it == area_participants.end()) { cout << "\n"; return; }
    bool first = true;
    for (auto& id : it->second) {
        if (!first) cout << " ";
        cout << id;
        first = false;
    }
    cout << "\n";
}

// areas_count
void cmd_areas_count() {
    cout << area_participants.size() << "\n";
}

// specialists_count <k> <a_1> ... <a_k>
void cmd_specialists_count(vector<string>& areas) {
    set<string> result;
    for (auto& a : areas) {
        auto it = area_participants.find(a);
        if (it != area_participants.end())
            for (auto& id : it->second) result.insert(id);
    }
    cout << result.size() << "\n";
}

// top_area
void cmd_top_area() {
    string best_area;
    size_t best_count = 0;
    for (auto& [area, participants] : area_participants) {
        // map iterates in lexicographic order, so first tie wins
        if (participants.size() > best_count) {
            best_count = participants.size();
            best_area = area;
        }
    }
    cout << best_area << "\n";
}
```

**Walkthrough of the example:**

1. `areas_count` → 0 (no areas yet)
2. `add IU483C9YW0 3 math computer_science spanish` → adds 3 areas; system has {math, computer_science, spanish}
3. `areas_count` → 3
4. `add DS983A7ILJ 3 math physics math` → adds math, physics (second "math" ignored as duplicate)
5. `add IU483C9YW0 1 art` → adds art to IU483C9YW0
6. `remove IU483C9YW0 computer_science` → removes computer_science from IU483C9YW0; areas now {math, spanish, art, physics}
7. `has IU483C9YW0 computer_science` → **No**
8. `areas_count` → **4**
9. `top_area` → **math** (has 2 participants; all others have 1)
10. `area_list math` → **DS983A7ILJ IU483C9YW0** (sorted)
11. `specialists_count 3 french math physics` → "french" not in system (ignored); participants with math or physics = {IU483C9YW0, DS983A7ILJ} → **2**

**Complexity:** Each `add`/`remove` is $O(\log n)$ for map/set operations. `specialists_count` is $O(k \cdot 1000 \cdot \log n)$ in the worst case. `top_area` is $O(|\text{areas}|)$ per call.

**Answer:** The solution uses two ordered maps for bidirectional participant–area mapping. See the C++ implementation above.

</details>

##### **4.23. AVL Tree Implementation** (Coding Exercise 6, Problem B)

Starting from an empty AVL tree, process $N$ commands ($1 \le N \le 10^5$):

- `ADD k v` — insert key $k$ with value $v$. If $k$ already exists, print `KEY ALREADY EXISTS`.
- `LOOKUP k` — print the value $v$ for key $k$. If not found, print `KEY NOT FOUND`.
- `DELETE k` — delete key $k$. If not found, print `KEY NOT FOUND`.
- `PRINT_ROTATIONS` — print the total number of rotations performed so far.

Keys and values are integers in $[-100, 100]$.

**Example:**
```
Input: 10
ADD 7 5 / ADD 3 2 / ADD 10 100 / ADD 2 3 / ADD 6 6
ADD 4 4 / ADD 8 7 / DELETE 7 / PRINT_ROTATIONS / LOOKUP 8
Output: 4 / 7
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** Implement a full AVL tree with height-tracking at each node. After each insert or delete, walk back up the tree updating heights and applying rotations where $|\text{bf}| > 1$. Count every rotation (each single rotation = 1; each double rotation = 2).

**Node structure:**

```cpp
struct Node {
    int key, val, height;
    Node *left, *right;
    Node(int k, int v) : key(k), val(v), height(0), left(nullptr), right(nullptr) {}
};

int height(Node* n) { return n ? n->height : -1; }

void update_height(Node* n) {
    if (n) n->height = 1 + max(height(n->left), height(n->right));
}

int bf(Node* n) { return n ? height(n->left) - height(n->right) : 0; }
```

**Rotations (each increments a global counter):**

```cpp
int rotation_count = 0;

Node* rotate_right(Node* y) {
    rotation_count++;
    Node* x = y->left;
    y->left = x->right;
    x->right = y;
    update_height(y);
    update_height(x);
    return x;
}

Node* rotate_left(Node* x) {
    rotation_count++;
    Node* y = x->right;
    x->right = y->left;
    y->left = x;
    update_height(x);
    update_height(y);
    return y;
}
```

**Rebalancing after each insertion/deletion:**

```cpp
Node* rebalance(Node* n) {
    update_height(n);
    int b = bf(n);
    if (b == 2) {          // left-heavy
        if (bf(n->left) < 0)
            n->left = rotate_left(n->left);   // LR case
        return rotate_right(n);
    }
    if (b == -2) {         // right-heavy
        if (bf(n->right) > 0)
            n->right = rotate_right(n->right); // RL case
        return rotate_left(n);
    }
    return n;
}
```

**Insert:**

```cpp
Node* insert(Node* n, int key, int val, bool& already_exists) {
    if (!n) return new Node(key, val);
    if (key == n->key) { already_exists = true; return n; }
    if (key < n->key) n->left = insert(n->left, key, val, already_exists);
    else              n->right = insert(n->right, key, val, already_exists);
    return rebalance(n);
}
```

**Delete:**

```cpp
Node* find_min(Node* n) { return n->left ? find_min(n->left) : n; }

Node* delete_min(Node* n) {
    if (!n->left) return n->right;
    n->left = delete_min(n->left);
    return rebalance(n);
}

Node* remove(Node* n, int key, bool& found) {
    if (!n) return nullptr;
    if (key < n->key) n->left = remove(n->left, key, found);
    else if (key > n->key) n->right = remove(n->right, key, found);
    else {
        found = true;
        if (!n->right) { Node* tmp = n->left; delete n; return tmp; }
        Node* succ = find_min(n->right);
        n->key = succ->key; n->val = succ->val;
        n->right = delete_min(n->right);
    }
    return rebalance(n);
}
```

**Main loop:**

```cpp
int main() {
    int N; cin >> N;
    Node* root = nullptr;
    while (N--) {
        string cmd; cin >> cmd;
        if (cmd == "ADD") {
            int k, v; cin >> k >> v;
            bool exists = false;
            root = insert(root, k, v, exists);
            if (exists) cout << "KEY ALREADY EXISTS\n";
        } else if (cmd == "LOOKUP") {
            int k; cin >> k;
            Node* cur = root;
            while (cur && cur->key != k) cur = k < cur->key ? cur->left : cur->right;
            if (!cur) cout << "KEY NOT FOUND\n";
            else cout << cur->val << "\n";
        } else if (cmd == "DELETE") {
            int k; cin >> k;
            bool found = false;
            root = remove(root, k, found);
            if (!found) cout << "KEY NOT FOUND\n";
        } else { // PRINT_ROTATIONS
            cout << rotation_count << "\n";
        }
    }
}
```

**Trace of the example:**

Inserting 7, 3, 10, 2, 6, 4, 8 in order triggers rotations to maintain AVL balance. After `DELETE 7` (which may trigger rebalancing), `PRINT_ROTATIONS` outputs **4**. `LOOKUP 8` finds value **7**.

**Answer:** The AVL implementation tracks heights and balance factors at each node, applying single or double rotations as needed. For the given input, 4 total rotations occur and the value at key 8 is 7.

</details>

##### **4.24. Overlapping Slots — Interval Tree** (Coding Exercise 6, Problem C)

You are managing a tourist booking system. Bookings are intervals $[L, R]$ ($1 \le L \le R \le 10^5$). Process $Q$ queries:

- `add L R` — add a booking $[L, R]$.
- `remove L R` — remove one occurrence of $[L, R]$; ignore if not present.
- `check L R` — print `Yes` if $[L, R]$ overlaps any existing booking, `No` otherwise.
- `count L R` — print the number of bookings that start AND end within $[L, R]$ inclusively.

Two intervals $[a, b]$ and $[c, d]$ **overlap** if and only if $a \le d$ and $c \le b$.

**Example:**
```
Input: 9
count 1 10        → 0
add 1 7
check 10 20       → No
add 2 9
count 1 10        → 2
add 2 9
count 1 10        → 3
check 5 6         → Yes
remove 2 9

Output: 0 / No / 2 / 3 / Yes
```

<details>
<summary>Click to see the solution</summary>

**Key Concept:** This problem requires an efficient *interval data structure*. We need:

- Overlap detection: does any stored interval touch $[L, R]$?
- Range count: how many stored intervals $[a, b]$ satisfy $L \le a$ and $b \le R$?

**Strategy:**

Since $L, R \le 10^5$, we can use a **segment tree with lazy propagation** or a simpler approach based on the observation that overlapping intervals satisfy $a \le R$ and $b \ge L$ (equivalently, NOT ($a > R$ or $b < L$)).

For this problem with modest constraints, we can store intervals in a multimap and answer each query by iterating:

```cpp
#include <bits/stdc++.h>
using namespace std;

// Store bookings as multimap: left endpoint -> right endpoint
multimap<int, int> bookings;

// add L R
void cmd_add(int L, int R) {
    bookings.insert({L, R});
}

// remove L R
void cmd_remove(int L, int R) {
    auto range = bookings.equal_range(L);
    for (auto it = range.first; it != range.second; ++it) {
        if (it->second == R) {
            bookings.erase(it);
            return;
        }
    }
    // not found: ignore
}

// check L R: does any interval [a,b] overlap [L,R]?
// Overlap condition: a <= R and b >= L
void cmd_check(int L, int R) {
    // All intervals with a <= R: look in bookings up to key R
    auto it = bookings.begin();
    while (it != bookings.end() && it->first <= R) {
        if (it->second >= L) {
            cout << "Yes\n";
            return;
        }
        ++it;
    }
    cout << "No\n";
}

// count L R: intervals [a,b] with L <= a and b <= R
void cmd_count(int L, int R) {
    int cnt = 0;
    auto it = bookings.lower_bound(L);  // a >= L
    while (it != bookings.end() && it->first <= R) {  // a <= R
        if (it->second <= R)  // b <= R
            cnt++;
        ++it;
    }
    cout << cnt << "\n";
}
```

**Walkthrough of the example:**

1. `count 1 10` → 0 (no bookings)
2. `add 1 7` → bookings: {[1,7]}
3. `check 10 20` → any interval with $a \le 20$ and $b \ge 10$? [1,7]: $b=7 < 10$ → No. Output: **No**
4. `add 2 9` → bookings: {[1,7], [2,9]}
5. `count 1 10` → [1,7]: $1 \ge 1, 7 \le 10$ ✓; [2,9]: $2 \ge 1, 9 \le 10$ ✓ → **2**
6. `add 2 9` → bookings: {[1,7], [2,9], [2,9]}
7. `count 1 10` → all 3 are within [1,10] → **3**
8. `check 5 6` → [1,7]: $a=1 \le 6$ and $b=7 \ge 5$ → overlap! → **Yes**
9. `remove 2 9` → remove one [2,9]; bookings: {[1,7], [2,9]}

**Overlap detection for larger inputs:**

For better performance with larger $Q$, use an **interval tree** (augmented BST where each node stores the maximum right endpoint in its subtree). This allows overlap queries in $O(\log n)$ instead of $O(n)$. The `count` query requires a 2D range query, efficiently handled with a persistent segment tree or 2D BIT.

**Time complexity (naive):**

- `add`/`remove`: $O(\log n)$
- `check`: $O(n)$ worst case
- `count`: $O(n)$ worst case

**Answer:** Use a multimap keyed by left endpoint. The `check` query scans intervals with left endpoint $\le R$ looking for right endpoint $\ge L$. The `count` query counts intervals fully contained in $[L, R]$.

</details>

---

#### **5. Midterm Exam 2026**

##### **5.1. Asymptotic Complexity of SOLVE** (Midterm 2026, Question A)

Compute the asymptotic worst-case time complexity of the SOLVE procedure:

```
Solve(A, n):
  return Helper(A, 1, n)

Helper(A, l, r):
  if r - l <= 100
    return l
  else
    k := ⌈(r - l + 1) / 6⌉
    a = Helper(A, l, l + 2*k) + Helper(A, l + k, l + 3*k)
    b = Helper(A, l + 2*k, l + 4*k) + Helper(A, l + 3*k, l + 5*k)
    c = Helper(A, l + 4*k, r)
    low := l; high := r
    while low < high:
      mid := ⌊(low + high) / 2⌋
      count := 0
      for i from l to r:
        if A[i] > A[mid]: count := count + 1
      if count > (r - l + 1) / 2 then high := mid else low := mid + 1
    return low
```

1. Express the running time $T(n)$ as a recurrence relation.
2. Find the asymptotic complexity using the Master Method, specifying which case applies.

<details><summary>Click to see the solution</summary>

**Step 1: Set up the recurrence.**

Let $n = r - l + 1$ be the size of the subarray. When $n \le 101$, the function returns immediately: $T(n) = \Theta(1)$.

For larger $n$, set $k = \lceil n/6 \rceil \approx n/6$.

Count the recursive calls:
- `Helper(A, l, l + 2*k)` — subarray of size $2k + 1 \approx n/3$
- `Helper(A, l + k, l + 3*k)` — subarray of size $2k + 1 \approx n/3$
- `Helper(A, l + 2*k, l + 4*k)` — subarray of size $2k + 1 \approx n/3$
- `Helper(A, l + 3*k, l + 5*k)` — subarray of size $2k + 1 \approx n/3$
- `Helper(A, l + 4*k, r)` — subarray of size $\approx n - 4k \approx n/3$

So there are **5 recursive calls**, each on a subarray of size $\approx n/3$.

The non-recursive work: the `while` loop runs at most $O(\log n)$ iterations (binary search on a range of size $n$). Each iteration runs the inner `for` loop over all $n$ elements: $\Theta(n)$ per iteration, so total non-recursive work is $\Theta(n \log n)$.

**Recurrence:**

$$T(n) = 5T(n/3) + \Theta(n \log n)$$

**Step 2: Apply the Master Theorem.**

The Master Theorem for $T(n) = aT(n/b) + f(n)$:

- $a = 5$, $b = 3$, so $n^{\log_b a} = n^{\log_3 5}$.
- $\log_3 5 \approx 1.465$.
- $f(n) = \Theta(n \log n) = \Theta(n^1 \log n)$.

Compare $f(n)$ with $n^{\log_3 5} \approx n^{1.465}$:

Since $n \log n = o(n^{1.465})$ (polynomial beats polylog), we have $f(n) = O(n^{\log_3 5 - \varepsilon})$ for $\varepsilon \approx 0.465 > 0$.

**Case 1 of the Master Theorem applies.**

$$\boxed{T(n) = \Theta(n^{\log_3 5})}$$

</details>

---

##### **5.2. Asymptotic Comparison Table** (Midterm 2026, Question B)

Use the most precise asymptotic notation ($O$, $\Theta$, $\Omega$) or $\times$ if none applies, to express the relationship $f(n) = [?](g(n))$:

| # | $f(n)$ | $g(n)$ |
|---|--------|--------|
| 1 | $2026^n$ | $(n+2026)!$ |
| 2 | $(2026/2025)^n$ | $n^{2026/2025}$ |
| 3 | $(\log_2 n)^{2026}$ | $\sqrt[2026]{n}$ |
| 4 | $n \log_2 n$ | $n^2 / \log_2 n$ |
| 5 | $\log_2(n^n)$ | $\log_2(n!)$ |
| 6 | $\sqrt{n \log_2 n}$ | $\sqrt{n} \cdot \log_2 \sqrt{n}$ |
| 7 | $(1 + 1/n)^n$ | $n$ |
| 8 | $2^n + 2^n$ | $4^n$ |

<details><summary>Click to see the solution</summary>

**1. $2026^n$ vs $(n+2026)!$**

$(n+2026)!$ grows much faster than any exponential $c^n$. For large $n$, $(n+2026)! \gg 2026^n$.

**Answer:** $f(n) = O(g(n))$

---

**2. $(2026/2025)^n$ vs $n^{2026/2025}$**

$(2026/2025)^n$ is exponential (base $> 1$), while $n^{2026/2025}$ is polynomial. Exponential grows faster.

**Answer:** $f(n) = \Omega(g(n))$

---

**3. $(\log_2 n)^{2026}$ vs $\sqrt[2026]{n} = n^{1/2026}$**

Any positive power of $n$ grows faster than any power of $\log n$: $(\log n)^k = o(n^\varepsilon)$ for any $\varepsilon > 0$.

Here $\varepsilon = 1/2026 > 0$, so $(\log_2 n)^{2026} = o(n^{1/2026})$.

**Answer:** $f(n) = O(g(n))$

---

**4. $n \log_2 n$ vs $n^2 / \log_2 n$**

$\frac{g(n)}{f(n)} = \frac{n^2/\log n}{n \log n} = \frac{n}{\log^2 n} \to \infty$.

So $f(n) = o(g(n))$.

**Answer:** $f(n) = O(g(n))$

---

**5. $\log_2(n^n)$ vs $\log_2(n!)$**

$\log_2(n^n) = n \log_2 n$.

By Stirling's approximation: $\log_2(n!) = \Theta(n \log n)$.

More precisely, $n \log n - n/\ln 2 \le \log_2(n!) \le n \log n$, so both are $\Theta(n \log n)$.

**Answer:** $f(n) = \Theta(g(n))$

---

**6. $\sqrt{n \log_2 n}$ vs $\sqrt{n} \cdot \log_2 \sqrt{n}$**

$\sqrt{n \log_2 n} = \sqrt{n} \cdot \sqrt{\log_2 n}$.

$\sqrt{n} \cdot \log_2 \sqrt{n} = \sqrt{n} \cdot \frac{1}{2}\log_2 n$.

Ratio: $\frac{\sqrt{n} \cdot \sqrt{\log n}}{\sqrt{n} \cdot \frac{1}{2}\log n} = \frac{\sqrt{\log n}}{\frac{1}{2}\log n} = \frac{2}{\sqrt{\log n}} \to 0$.

So $f(n) = o(g(n))$.

**Answer:** $f(n) = O(g(n))$

---

**7. $(1 + 1/n)^n$ vs $n$**

$(1 + 1/n)^n \to e \approx 2.718$ as $n \to \infty$ (converges to a constant). Meanwhile $g(n) = n \to \infty$.

**Answer:** $f(n) = O(g(n))$

---

**8. $2^n + 2^n$ vs $4^n$**

$2^n + 2^n = 2 \cdot 2^n = 2^{n+1}$.

$4^n = (2^2)^n = 2^{2n}$.

$\frac{2^{2n}}{2^{n+1}} = 2^{n-1} \to \infty$, so $f(n) = o(g(n))$.

**Answer:** $f(n) = O(g(n))$

</details>

---

##### **5.3. Master Theorem Application** (Midterm 2026, Question C)

Consider $T(n) = 8T(n/3) + 5n^2 + 10$. Which of the following is correct?

1. $T(n) = \Theta(n^2 \log n)$
2. $T(n) = \Theta(n^3)$
3. Master theorem not applicable
4. None of the above

Answer and justify.

<details><summary>Click to see the solution</summary>

**Answer: Option 4 — None of the above.**

Apply the Master Theorem with $a = 8$, $b = 3$, $f(n) = 5n^2 + 10 = \Theta(n^2)$.

Compute $n^{\log_b a} = n^{\log_3 8}$.

$\log_3 8 = \frac{\ln 8}{\ln 3} = \frac{3\ln 2}{\ln 3} \approx \frac{3 \times 0.693}{1.099} \approx 1.893$.

So $n^{\log_3 8} \approx n^{1.893}$.

Compare $f(n) = \Theta(n^2)$ with $n^{1.893}$:

Since $n^2 = \Omega(n^{1.893 + \varepsilon})$ for $\varepsilon \approx 0.107 > 0$, **Case 3** of the Master Theorem applies — provided the regularity condition holds.

Regularity condition: $af(n/b) \le cf(n)$ for some $c < 1$.

$8 \cdot f(n/3) = 8 \cdot 5(n/3)^2 = 8 \cdot \frac{5n^2}{9} = \frac{40n^2}{9} \approx 4.44 \cdot n^2/1 \le c \cdot 5n^2$ requires $c \ge 8/9 < 1$. ✓

So **Case 3 applies** and:

$$T(n) = \Theta(f(n)) = \Theta(n^2)$$

None of the listed options (1), (2), (3) is correct — the correct answer is $T(n) = \Theta(n^2)$, which is option **(4) None of the above**.

</details>

---

##### **5.4. Data Structure Complexity Table** (Midterm 2026, Question D)

Collection $A$ of $n$ columns, each column $C_i$ has $n$ elements. Extend $A$ with column $C_{n+1}$ consisting of the diagonal elements $C_1[1], C_2[2], \ldots, C_n[n]$.

Fill the table of asymptotic worst-case time complexity for creating $C_{n+1}$ and appending it to $A$:

| Representation of $A$ | Representation of $C_i$ | Time to build $C_{n+1}$ and append |
|-----------------------|------------------------|--------------------------------------|
| Array List | Array List | ? |
| Array List | Singly-Linked (no tail) | ? |
| Array List | Singly-Linked (with tail) | ? |
| Singly-Linked (no tail) | Array List | ? |
| Singly-Linked (no tail) | Singly-Linked (no tail) | ? |
| Singly-Linked (no tail) | Singly-Linked (with tail) | ? |
| Singly-Linked (with tail) | Array List | ? |
| Singly-Linked (with tail) | Singly-Linked (no tail) | ? |
| Singly-Linked (with tail) | Singly-Linked (with tail) | ? |

<details><summary>Click to see the solution</summary>

**Key operations:**

1. **Access $C_i$ (the $i$-th column of $A$):** $O(1)$ for Array List, $O(i)$ for Singly-Linked (no tail) — must traverse to position $i$.
2. **Access $C_i[i]$ (the $i$-th element of $C_i$):** $O(1)$ for Array List, $O(i)$ for Singly-Linked (any) — must traverse to position $i$.
3. **Append $C_{n+1}$ to $A$:** $O(1)$ amortized for Array List (or $O(n)$ worst case resize), $O(n)$ for Singly-Linked (no tail) — must traverse to end, $O(1)$ for Singly-Linked (with tail).
4. **Building $C_{n+1}$:** We retrieve elements $C_1[1], C_2[2], \ldots, C_n[n]$. For each $i$, we access column $C_i$ then element $[i]$.

**Cost of collecting all $n$ diagonal elements:**

- If $A$ is an **Array List**: accessing $C_i$ is $O(1)$, so cost of accessing all $C_i$ is $O(n)$ total.
- If $A$ is a **Singly-Linked (no tail)**: accessing $C_i$ costs $O(i)$, total $\sum_{i=1}^n O(i) = O(n^2)$.
- If $A$ is a **Singly-Linked (with tail)**: same as no-tail for random access — $O(i)$ per access, total $O(n^2)$.

Within each $C_i$, accessing element at position $i$:
- **Array List**: $O(1)$ per element, $O(n)$ total over all $i$.
- **Singly-Linked (no tail or with tail)**: $O(i)$ per element, $\sum_{i=1}^n O(i) = O(n^2)$ total.

**Appending $C_{n+1}$ to $A$** (building $C_{n+1}$ itself as a list of $n$ elements):
- The new column must be built by inserting $n$ elements. Cost $O(n)$ regardless of representation.
- Appending the column pointer to $A$: $O(1)$ Array List (amortized), $O(n)$ Singly-Linked no tail, $O(1)$ Singly-Linked with tail.

**Combined complexities (dominant term):**

| $A$ | $C_i$ | Total |
|-----|--------|-------|
| Array List | Array List | $O(n)$ |
| Array List | Singly-Linked (no tail) | $O(n^2)$ |
| Array List | Singly-Linked (with tail) | $O(n^2)$ |
| Singly-Linked (no tail) | Array List | $O(n^2)$ (accessing $A$) + $O(n)$ append = $O(n^2)$ |
| Singly-Linked (no tail) | Singly-Linked (no tail) | $O(n^2)$ |
| Singly-Linked (no tail) | Singly-Linked (with tail) | $O(n^2)$ |
| Singly-Linked (with tail) | Array List | $O(n^2)$ (accessing $A$) + $O(1)$ append = $O(n^2)$ |
| Singly-Linked (with tail) | Singly-Linked (no tail) | $O(n^2)$ |
| Singly-Linked (with tail) | Singly-Linked (with tail) | $O(n^2)$ |

**Answer:** Only when both $A$ and $C_i$ are Array Lists is the total $O(n)$; all other combinations are $O(n^2)$.

</details>

---

##### **5.5. RADIX-SORT for Phrases** (Midterm 2026, Question E)

$p$ phrases are sorted using RADIX-SORT, treating each word as a "digit":
- Each phrase has at most $w$ words
- Each word has at most $s$ symbols
- Each symbol is from an alphabet of size $a$
- Comparing two symbols takes $\Theta(1)$

Find the worst-case time complexity for sorting $p$ phrases when the stable sort for words uses:

1. INSERTION-SORT
2. COUNTING-SORT
3. QUICK-SORT

<details><summary>Click to see the solution</summary>

**Setup:**

RADIX-SORT processes $w$ digit positions (one per word position in a phrase). At each position, we stably sort $p$ phrases by the word at that position.

Comparing two words of at most $s$ symbols costs $O(s)$ (compare symbol by symbol, each $\Theta(1)$).

**1. INSERTION-SORT for sorting words:**

- Comparing two words: $O(s)$.
- INSERTION-SORT on $p$ items: $O(p^2)$ comparisons worst case.
- Cost per RADIX-SORT pass: $O(p^2 \cdot s)$.
- Total for $w$ passes: $O(w \cdot p^2 \cdot s)$.

$$\boxed{O(w \cdot s \cdot p^2)}$$

**2. COUNTING-SORT for sorting words:**

COUNTING-SORT treats each word as a key. However, words are strings — to use COUNTING-SORT, we'd need to sort by the full word as a key. With alphabet size $a$ and word length $s$, there are $a^s$ possible words — the counting array would be huge.

Alternatively, COUNTING-SORT on words using the word as a composite key requires $\Theta(p + a^s)$ per pass.

Total for $w$ passes: $O(w \cdot (p + a^s))$.

$$\boxed{O(w \cdot (p + a^s))}$$

**3. QUICK-SORT for sorting words:**

QUICK-SORT is **not stable**, so using it directly inside RADIX-SORT breaks the algorithm's correctness. Assuming we use a stable variant or accept the instability issue:

- QUICK-SORT on $p$ items: $O(p^2)$ worst case.
- Each comparison: $O(s)$.
- Cost per pass: $O(p^2 \cdot s)$.
- Total: $O(w \cdot s \cdot p^2)$.

Note: Since QUICK-SORT is not stable, RADIX-SORT with QUICK-SORT does not correctly sort in general. This is a fundamental issue.

$$\boxed{O(w \cdot s \cdot p^2)}$$

</details>

---

##### **5.6. COUNTING-SORT Application** (Midterm 2026, Question F)

Apply COUNTING-SORT to the following array:

| key  | 3 | 1 | 2 | 0 | 2 | 1 | 4 | 0 | 2 | 1 | 0 | 3 |
|------|---|---|---|---|---|---|---|---|---|---|---|---|
| data | I | E | D | R | S | A | E | I | H | N | S | N |

1. Give the auxiliary array $C$ after the cumulative (prefix-sum) step for $k \in \{0,1,2,3,4\}$.
2. Give the output array (sorted).

<details><summary>Click to see the solution</summary>

**Step 1: Count occurrences.**

Scan the key array: keys are $3,1,2,0,2,1,4,0,2,1,0,3$.

| key $k$ | 0 | 1 | 2 | 3 | 4 |
|---------|---|---|---|---|---|
| count   | 3 | 3 | 3 | 2 | 1 |

**Step 2: Cumulative sum (prefix sum).**

$C[k] = \sum_{j=0}^{k} \text{count}[j]$

| $k$    | 0 | 1 | 2 | 3  | 4  |
|--------|---|---|---|----|----|
| $C[k]$ | 3 | 6 | 9 | 11 | 12 |

**Answer to part 1:** $C = [3, 6, 9, 11, 12]$

**Step 3: Build output array (process input right to left for stability).**

Output array of size 12, 1-indexed positions.

Process input from right to left:

| Step | key | data | $C[key]$ | place at | new $C[key]$ |
|------|-----|------|----------|----------|--------------|
| 12 | 3 | N | 11 | pos 11 | 10 |
| 11 | 0 | S | 3  | pos 3  | 2  |
| 10 | 1 | N | 6  | pos 6  | 5  |
| 9  | 2 | H | 9  | pos 9  | 8  |
| 8  | 0 | I | 2  | pos 2  | 1  |
| 7  | 4 | E | 12 | pos 12 | 11 |
| 6  | 1 | A | 5  | pos 5  | 4  |
| 5  | 2 | S | 8  | pos 8  | 7  |
| 4  | 2 | D | 7  | pos 7  | 6  |  *(wait — recount)*
| ... | | | | | |

Let me redo carefully. Input (1-indexed):

| pos | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 |
|-----|---|---|---|---|---|---|---|---|---|----|----|----|
| key | 3 | 1 | 2 | 0 | 2 | 1 | 4 | 0 | 2 | 1  | 0  | 3  |
| dat | I | E | D | R | S | A | E | I | H | N  | S  | N  |

$C$ after cumulative step: $C[0]=3, C[1]=6, C[2]=9, C[3]=11, C[4]=12$.

Process right to left (positions 12 down to 1):

- pos=12: key=3, data=N → output[C[3]]=output[11]=N, C[3]=10
- pos=11: key=0, data=S → output[C[0]]=output[3]=S, C[0]=2
- pos=10: key=1, data=N → output[C[1]]=output[6]=N, C[1]=5
- pos=9:  key=2, data=H → output[C[2]]=output[9]=H, C[2]=8
- pos=8:  key=0, data=I → output[C[0]]=output[2]=I, C[0]=1
- pos=7:  key=4, data=E → output[C[4]]=output[12]=E, C[4]=11
- pos=6:  key=1, data=A → output[C[1]]=output[5]=A, C[1]=4
- pos=5:  key=2, data=S → output[C[2]]=output[8]=S, C[2]=7
- pos=4:  key=0, data=R → output[C[0]]=output[1]=R, C[0]=0
- pos=3:  key=2, data=D → output[C[2]]=output[7]=D, C[2]=6
- pos=2:  key=1, data=E → output[C[1]]=output[4]=E, C[1]=3
- pos=1:  key=3, data=I → output[C[3]]=output[10]=I, C[3]=9

**Output array:**

| pos | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 |
|-----|---|---|---|---|---|---|---|---|---|----|----|----|
| key | 0 | 0 | 0 | 1 | 1 | 1 | 2 | 2 | 2 | 3  | 3  | 4  |
| dat | R | I | S | E | A | N | D | S | H | I  | N  | E  |

**Answer to part 2:** The sorted array is: $R, I, S, E, A, N, D, S, H, I, N, E$ (keys: $0,0,0,1,1,1,2,2,2,3,3,4$).

Note: Reading the data in sorted order spells out **RISE AND SHINE** — a cute exam easter egg!

</details>

---

##### **5.7. True or False** (Midterm 2026, Question G)

Determine TRUE or FALSE for each statement:

1. If $\log(f(n)) = \Theta(\log(g(n)))$ then $f(n) = \Theta(g(n))$
2. Incrementing a binary counter takes $\Theta(\log k)$ worst-case time where $k$ is the number of digits
3. In ARRAYSTACK, PUSH(x) has $\Theta(n)$ worst-case time complexity
4. MERGE-SORT has $\Theta(n \log n)$ worst-case time complexity
5. COUNTING-SORT sorts $n$ integers in $[0, n^2]$ in $\Theta(n)$ worst-case time
6. If the master theorem doesn't apply, the algorithm may loop indefinitely
7. Dynamic Programming involves a dynamically-typed programming language (e.g., Python)
8. QUICK-SORT has $\Theta(n \log n)$ worst-case time complexity
9. In a BST, the node with minimum key cannot have a right child
10. Height of a red-black tree with $n$ internal nodes is $\Theta(\log n)$

<details><summary>Click to see the solution</summary>

**1. FALSE.**

Counterexample: $f(n) = n^2$, $g(n) = n^3$. Then $\log f(n) = 2\log n$, $\log g(n) = 3\log n$, so $\log f(n) = \Theta(\log g(n))$. But $f(n) = O(g(n))$, not $\Theta(g(n))$.

---

**2. TRUE.**

In the worst case, incrementing a $k$-bit binary counter flips all $k$ bits (e.g., $0111\ldots1 \to 1000\ldots0$), which takes $\Theta(k)$ time. So worst case is $\Theta(k)$... but the question says $\Theta(\log k)$, which is wrong — it should be $\Theta(k)$.

**FALSE.** Worst case is $\Theta(k)$, not $\Theta(\log k)$.

---

**3. FALSE.**

ARRAYSTACK PUSH(x) appends to the end of an array. This is $O(1)$ amortized (with dynamic resizing), and $O(1)$ worst case if no resize is needed. Worst case with resizing is $O(n)$, but typically we say PUSH is $O(1)$ amortized. As a strict worst-case statement: if resize happens, it's $\Theta(n)$, so this is **TRUE** for worst case (a single PUSH can cost $\Theta(n)$).

**TRUE** (a single PUSH can take $\Theta(n)$ in the worst case due to array resizing).

---

**4. TRUE.**

MERGE-SORT always divides the array in half and merges, giving $T(n) = 2T(n/2) + \Theta(n)$, which solves to $\Theta(n \log n)$ in all cases (best, average, worst).

---

**5. FALSE.**

COUNTING-SORT on $n$ integers in $[0, n^2]$ requires an auxiliary array of size $n^2 + 1$, so it takes $\Theta(n + n^2) = \Theta(n^2)$ time, not $\Theta(n)$.

---

**6. FALSE.**

If the master theorem doesn't apply, it simply means the theorem cannot be used — the algorithm still terminates normally. We just need another method (e.g., substitution, recursion tree) to find the complexity.

---

**7. FALSE.**

Dynamic Programming (DP) is an algorithmic technique for solving problems with overlapping subproblems and optimal substructure. It has nothing to do with dynamically-typed programming languages.

---

**8. FALSE.**

QUICK-SORT has worst-case $\Theta(n^2)$ time (when pivots are always the smallest or largest element). Its average-case is $\Theta(n \log n)$.

---

**9. FALSE.**

The minimum node in a BST has no **left** child (no smaller key exists). It can have a right child. For example, a BST with root 5, left child 3, and 3 has a right child 4: node 3 is not the minimum (1 might not exist), but the minimum node can have a right child.

More precisely: the minimum node has no left child, but may have a right child.

---

**10. TRUE.**

A red-black tree with $n$ internal nodes has height $h \le 2\log_2(n+1)$, so $h = O(\log n)$. Also $h = \Omega(\log n)$ since it's a binary tree. Therefore height is $\Theta(\log n)$.

</details>

---

##### **5.8. BST Array Representation** (Midterm 2026, Question H)

A BST is stored in an array representation (indices 0–15). The initial tree $T$:

| idx | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 | 13 | 14 | 15 |
|-----|---|---|---|---|---|---|---|---|---|---|----|----|----|----|----|----|
| key | 8 | 4 | 12| 2 | 6 |10 | – | – | – | – | –  | 9  | 11 | –  | –  | –  |

The array uses level-order (BFS) indexing: root at index 0, left child of $i$ at $2i+1$, right child at $2i+2$.

1. Array after inserting 20, 5, 17 (in order) into initial tree $T$
2. Array after deleting key 10 from initial tree $T$
3. Array after deleting key 8 from initial tree $T$

<details><summary>Click to see the solution</summary>

**Tree structure from initial array:**

```
Index 0: 8 (root)
Index 1: 4 (left child of 8)
Index 2: 12 (right child of 8)
Index 3: 2 (left child of 4)
Index 4: 6 (right child of 4)
Index 5: 10 (left child of 12)
Index 6: – (right child of 12, empty)
Index 11: 9 (left child of 10, at 2*5+1=11)
Index 12: 11 (right child of 10, at 2*5+2=12)
```

**Part 1: Insert 20, 5, 17**

*Insert 20:* 20 > 8 → right (12) → 20 > 12 → right of 12 (index 6). Place 20 at index 6.

*Insert 5:* 5 < 8 → left (4) → 5 > 4 → right (6) → 5 < 6 → left of 6. Left of 6 is index 13. Place 5 at index 13.

*Insert 17:* 17 > 8 → right (12) → 17 > 12 → right (20 at idx 6) → 17 < 20 → left of 20. Left of 20 is index 13... wait, index 6 is 20, left child = 2*6+1=13, right = 2*6+2=14.

But we placed 5 at index 13 already — conflict! Let me re-examine.

Actually index 13 is the left child of node at index 6. After inserting 20 at index 6, node 20's left child would be at index 13. But we also placed 5 at index 13 when inserting 5 (going 8→4→6→left=13). These are different subtrees: index 13 as child of node at index 4 (key=6) is left child of 6, while index 13 as child of node at index 6 (key=20) is left child of 20. **They cannot both be index 13 — this is a conflict in the array representation.**

This means the array representation runs out of indices for certain insertions. In a flat array of size 16 with BFS indexing, some insertions may exceed the allocated space.

*Insert 5:* Path: 8→4→6→left. Left child of 6 (index 4) is index $2 \times 4 + 1 = 9$. Place 5 at index **9**.

*Insert 17:* Path: 8→12→20→left. Left child of 20 (index 6) is index $2 \times 6 + 1 = 13$. Place 17 at index **13**.

| idx | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 | 13 | 14 | 15 |
|-----|---|---|---|---|---|---|---|---|---|---|----|----|----|----|----|----|
| key | 8 | 4 | 12| 2 | 6 |10 |20 | – | – | 5 | –  | 9  | 11 | 17 | –  | –  |

**Part 2: Delete key 10 from initial $T$**

Node 10 is at index 5. It has two children: 9 (index 11) and 11 (index 12).

Standard BST deletion with two children: replace with in-order successor (smallest in right subtree). In-order successor of 10 is 11 (at index 12, no left child).

Replace key at index 5 with 11. Remove 11 from index 12.

| idx | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 | 13 | 14 | 15 |
|-----|---|---|---|---|---|---|---|---|---|---|----|----|----|----|----|----|
| key | 8 | 4 | 12| 2 | 6 |11 | – | – | – | – | –  | 9  | –  | –  | –  | –  |

**Part 3: Delete key 8 from initial $T$**

Node 8 is at index 0 (root). It has two children: 4 (left, index 1) and 12 (right, index 2).

In-order successor of 8 is 9 (leftmost in right subtree: 12→10→9). Node 9 is at index 11, it has no children.

Replace root key with 9. Remove 9 from index 11.

| idx | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 | 11 | 12 | 13 | 14 | 15 |
|-----|---|---|---|---|---|---|---|---|---|---|----|----|----|----|----|----|
| key | 9 | 4 | 12| 2 | 6 |10 | – | – | – | – | –  | –  | 11 | –  | –  | –  |

</details>

---

##### **5.9. Rod Cutting Problem** (Midterm 2026, Question I)

The Rod Cutting problem: given a rod of length $n$ and a table of prices $p_i$ for lengths $i = 1, \ldots, n$, find the maximum revenue $r_n$ obtainable by cutting the rod.

1. Worst-case time complexity of:
   - DP with tabulation (bottom-up)
   - DP with memoization (top-down)
2. Find the maximum revenue for a rod of length 10 with prices:

| length $i$ | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9 | 10 |
|------------|---|---|---|---|---|----|----|----|----|-----|
| price $p_i$| 1 | 5 | 5 | 8 |10 | 14 | 20 | 18 | 22 | 25 |

Present DP table values $r[0], r[1], \ldots, r[10]$.

<details><summary>Click to see the solution</summary>

**Part 1: Time complexities**

The recurrence is:
$$r[n] = \max_{1 \le i \le n} (p_i + r[n-i])$$

- **DP with tabulation (bottom-up):** For each subproblem of size $j$ from 1 to $n$, we try all $j$ cuts. Total work: $\sum_{j=1}^{n} j = \frac{n(n+1)}{2} = \Theta(n^2)$.

$$\boxed{\Theta(n^2)}$$

- **DP with memoization (top-down):** Each subproblem is solved at most once. Same analysis applies: $\Theta(n^2)$ in worst case.

$$\boxed{\Theta(n^2)}$$

**Part 2: DP Table**

$r[0] = 0$ (empty rod, no revenue).

$$r[j] = \max_{1 \le i \le j}(p_i + r[j - i])$$

- $r[0] = 0$
- $r[1] = p_1 + r[0] = 1 + 0 = 1$
- $r[2] = \max(p_1+r[1], p_2+r[0]) = \max(1+1, 5+0) = \max(2, 5) = 5$
- $r[3] = \max(p_1+r[2], p_2+r[1], p_3+r[0]) = \max(1+5, 5+1, 5+0) = \max(6,6,5) = 6$
- $r[4] = \max(p_1+r[3], p_2+r[2], p_3+r[1], p_4+r[0]) = \max(1+6, 5+5, 5+1, 8+0) = \max(7,10,6,8) = 10$
- $r[5] = \max(p_1+r[4], p_2+r[3], p_3+r[2], p_4+r[1], p_5+r[0])$
  $= \max(1+10, 5+6, 5+5, 8+1, 10+0) = \max(11,11,10,9,10) = 11$
- $r[6] = \max(p_1+r[5], p_2+r[4], p_3+r[3], p_4+r[2], p_5+r[1], p_6+r[0])$
  $= \max(1+11, 5+10, 5+6, 8+5, 10+1, 14+0) = \max(12,15,11,13,11,14) = 15$
- $r[7] = \max(p_1+r[6], p_2+r[5], p_3+r[4], p_4+r[3], p_5+r[2], p_6+r[1], p_7+r[0])$
  $= \max(1+15, 5+11, 5+10, 8+6, 10+5, 14+1, 20+0) = \max(16,16,15,14,15,15,20) = 20$
- $r[8] = \max(p_i + r[8-i])$ for $i=1..8$:
  $= \max(1+20, 5+15, 5+11, 8+10, 10+6, 14+5, 20+1, 18+0) = \max(21,20,16,18,16,19,21,18) = 21$
- $r[9] = \max(p_i + r[9-i])$ for $i=1..9$:
  $= \max(1+21, 5+20, 5+15, 8+11, 10+10, 14+6, 20+5, 18+1, 22+0)$
  $= \max(22,25,20,19,20,20,25,19,22) = 25$
- $r[10] = \max(p_i + r[10-i])$ for $i=1..10$:
  $= \max(1+25, 5+21, 5+20, 8+15, 10+11, 14+10, 20+6, 18+5, 22+1, 25+0)$
  $= \max(26,26,25,23,21,24,26,23,23,25) = 26$

**DP Table:**

| $j$ | 0 | 1 | 2 | 3 | 4  | 5  | 6  | 7  | 8  | 9  | 10 |
|-----|---|---|---|---|----|----|----|----|----|----|-----|
| $r[j]$ | 0 | 1 | 5 | 6 | 10 | 11 | 15 | 20 | 21 | 25 | 26 |

**Answer:** Maximum revenue for rod of length 10 is $r[10] = \mathbf{26}$.

One optimal cut: cut into length 7 + length 3 gives $p_7 + p_3 = 20 + 5 = 25$. Better: length 1 + length 9 gives $1 + 22 = 23$. Let's check $r[10]=26$: achieved by, e.g., $p_1 + r[9] = 1 + 25 = 26$, and $r[9]=25$ from $p_2 + r[7] = 5 + 20 = 25$. So: cut 10 = 1 + 2 + 7, giving $1 + 5 + 20 = 26$. ✓

</details>

---

#### **6. Mock Midterm Exam 2026**

##### **6.1. Asymptotic Complexity of Solve** (Mock Midterm 2026, Question A)

Compute the asymptotic worst-case time complexity of Solve:

```
Solve(A, n):
    return Helper(A, 1, n)

Helper(A, l, r):
    if r - l <= 50
        return l
    else
        low := l; high := r
        while low < high:
            mid := ⌊(low + high) / 2⌋
            count := 0
            for i from l to r:
                if A[i] > A[mid]: count := count + 1
            if count > (r - l + 1) / 2 then high := mid else low := mid + 1
        k := ⌈(r - l + 1) / 3⌉
        a := Helper(A, l,       l + k - 1)
        b := Helper(A, l + k,   l + 2*k - 1)
        c := Helper(A, l + 2*k, r)
        return low
```

1. Express $T(n)$ as a recurrence relation.
2. Find the asymptotic complexity using the Master Method, specifying the case and justification.

<details><summary>Click to see the solution</summary>

**Step 1: Set up the recurrence.**

Let $n = r - l + 1$. Base case: if $n \le 51$, return immediately.

For larger $n$, set $k = \lceil n/3 \rceil \approx n/3$.

The three recursive calls:
- `Helper(A, l, l + k - 1)` — size $k \approx n/3$
- `Helper(A, l + k, l + 2k - 1)` — size $k \approx n/3$
- `Helper(A, l + 2k, r)` — size $n - 2k \approx n/3$

So there are **3 recursive calls**, each on $\approx n/3$ elements.

Non-recursive work: the `while` loop runs $O(\log n)$ iterations (binary search). Each iteration has the inner `for` loop: $\Theta(n)$. Total non-recursive work: $\Theta(n \log n)$.

**Recurrence:**

$$T(n) = 3T(n/3) + \Theta(n \log n)$$

**Step 2: Apply the Master Theorem.**

$a = 3$, $b = 3$, $f(n) = \Theta(n \log n)$.

$n^{\log_b a} = n^{\log_3 3} = n^1 = n$.

Compare $f(n) = n \log n$ with $n^{\log_3 3} = n$:

$f(n) = n \log n = n^1 \cdot \log n$.

This is the boundary case: $f(n) = \Theta(n^{\log_b a} \cdot \log^k n)$ with $k = 1$.

**Case 2 of the Master Theorem (extended)** applies with $k = 1$:

$$T(n) = \Theta(n^{\log_b a} \cdot \log^{k+1} n) = \Theta(n \cdot \log^2 n)$$

$$\boxed{T(n) = \Theta(n \log^2 n)}$$

</details>

---

##### **6.2. Asymptotic Comparison Table** (Mock Midterm 2026, Question B)

Use the most precise asymptotic notation ($O$, $\Theta$, $\Omega$) or $\times$ if none applies:

| # | $f(n)$ | $g(n)$ |
|---|--------|--------|
| 1 | $\sqrt{n} \cdot \log_2 n$ | $n^{3/4}$ |
| 2 | $n^{1.001}$ | $n \log^2 n$ |
| 3 | $\log_2(n!)$ | $n \log_2 n$ |
| 4 | $\sqrt[4]{n}$ | $\log_2 n$ |
| 5 | $(1 + 2/n)^n$ | $n^2$ |
| 6 | $3^n + 2^n$ | $3^n$ |
| 7 | $n^2/\log n$ | $n\sqrt{n}$ |
| 8 | $(n+1)!$ | $n!$ |

<details><summary>Click to see the solution</summary>

**1. $\sqrt{n} \cdot \log_2 n$ vs $n^{3/4}$**

$f(n) = n^{1/2} \log n$. Compare with $g(n) = n^{3/4}$.

$\frac{f(n)}{g(n)} = \frac{n^{1/2}\log n}{n^{3/4}} = \frac{\log n}{n^{1/4}} \to 0$ as $n \to \infty$.

So $f(n) = o(g(n))$.

**Answer:** $f(n) = O(g(n))$

---

**2. $n^{1.001}$ vs $n \log^2 n$**

$\frac{f(n)}{g(n)} = \frac{n^{1.001}}{n \log^2 n} = \frac{n^{0.001}}{\log^2 n} \to \infty$.

Polynomial beats polylog.

**Answer:** $f(n) = \Omega(g(n))$

---

**3. $\log_2(n!)$ vs $n \log_2 n$**

By Stirling: $\log(n!) = \Theta(n \log n)$. Both are $\Theta(n \log n)$.

**Answer:** $f(n) = \Theta(g(n))$

---

**4. $\sqrt[4]{n} = n^{1/4}$ vs $\log_2 n$**

Any positive power of $n$ grows faster than $\log n$.

$\frac{n^{1/4}}{\log n} \to \infty$.

**Answer:** $f(n) = \Omega(g(n))$

---

**5. $(1 + 2/n)^n$ vs $n^2$**

$(1 + 2/n)^n \to e^2 \approx 7.389$ (constant). Meanwhile $g(n) = n^2 \to \infty$.

**Answer:** $f(n) = O(g(n))$

---

**6. $3^n + 2^n$ vs $3^n$**

$3^n + 2^n = 3^n(1 + (2/3)^n)$. As $n \to \infty$, $(2/3)^n \to 0$, so $3^n + 2^n \sim 3^n$.

$\frac{3^n + 2^n}{3^n} = 1 + (2/3)^n \to 1$.

**Answer:** $f(n) = \Theta(g(n))$

---

**7. $n^2/\log n$ vs $n\sqrt{n} = n^{3/2}$**

$\frac{f(n)}{g(n)} = \frac{n^2/\log n}{n^{3/2}} = \frac{n^{1/2}}{\log n} \to \infty$.

**Answer:** $f(n) = \Omega(g(n))$

---

**8. $(n+1)!$ vs $n!$**

$(n+1)! = (n+1) \cdot n!$, so $\frac{(n+1)!}{n!} = n+1 \to \infty$.

**Answer:** $f(n) = \Omega(g(n))$

</details>

---

##### **6.3. Brute-Force Subset Sum** (Mock Midterm 2026, Question C)

Given a set $A$ of $n$ integers and a target $k$, find a subset of $A$ of maximum size whose elements sum to exactly $k$. A brute-force algorithm enumerates all subsets.

1. Give pseudocode for a brute-force algorithm.
2. Worst-case time complexity in terms of $n$.
3. Brief justification.
4. Worst-case space complexity (excluding input) using iterative enumeration.

<details><summary>Click to see the solution</summary>

**Part 1: Pseudocode**

```
BruteForceMaxSubset(A, n, k):
    best := ∅
    for mask from 0 to 2^n - 1:
        S := ∅
        total := 0
        for i from 0 to n - 1:
            if (mask >> i) & 1 == 1:
                S := S ∪ {A[i]}
                total := total + A[i]
        if total == k and |S| > |best|:
            best := S
    return best
```

**Part 2: Worst-case time complexity**

$$\Theta(n \cdot 2^n)$$

**Part 3: Justification**

There are $2^n$ subsets. For each subset (represented as a bitmask), we iterate through all $n$ bits to reconstruct the subset and compute its sum. Each subset takes $\Theta(n)$ time, giving total $\Theta(n \cdot 2^n)$.

**Part 4: Space complexity (iterative enumeration)**

Using iterative enumeration with a bitmask, we only need:
- One variable for the current mask: $O(1)$ (or $O(n)$ bits)
- Variables for the current sum and size: $O(1)$
- The best subset found so far: $O(n)$

**Space complexity: $O(n)$** (excluding input).

</details>

---

##### **6.4. Merging $n$ Sorted Lists** (Mock Midterm 2026, Question D)

$n$ sorted lists contain $m$ elements in total. Merge them into one sorted list of $m$ elements.

Fill the table with worst-case time complexity depending on the representation of input lists and output list:

| Input / Output | Array | Singly-linked (with tail) | Doubly-linked |
|----------------|-------|--------------------------|---------------|
| Array | ? | ? | ? |
| Singly-linked (with tail) | ? | ? | ? |
| Doubly-linked | ? | ? | ? |

<details><summary>Click to see the solution</summary>

**Merging Strategy:**

Use a $k$-way merge: repeatedly extract the minimum element from the fronts of $n$ lists using a min-heap. Total steps: $m$ extractions, each costs $O(\log n)$. So the merge itself takes $O(m \log n)$.

**Key operations:**

- **Reading front of a list:** $O(1)$ for all representations (array: index 0; singly-linked: head; doubly-linked: head).
- **Removing front of a list:** $O(1)$ for all (array with pointer offset, linked list head removal).
- **Appending to output:**
  - **Array:** $O(1)$ amortized (dynamic array), but each element written: $O(1)$.
  - **Singly-linked with tail:** $O(1)$ per append (use tail pointer).
  - **Doubly-linked:** $O(1)$ per append (use tail pointer).

**Total complexity:**

The merge algorithm itself is $O(m \log n)$ regardless of representation (assuming efficient front access and removal). The output construction is $O(m)$ for all output types.

However, if we use a naive approach (no heap, sequential scan of list heads): each of $m$ elements requires scanning $n$ list fronts, giving $O(mn)$.

With a **min-heap** of size $n$:

| Input / Output | Array | Singly-linked (with tail) | Doubly-linked |
|----------------|-------|--------------------------|---------------|
| Array | $O(m \log n)$ | $O(m \log n)$ | $O(m \log n)$ |
| Singly-linked (with tail) | $O(m \log n)$ | $O(m \log n)$ | $O(m \log n)$ |
| Doubly-linked | $O(m \log n)$ | $O(m \log n)$ | $O(m \log n)$ |

**Answer:** All cells are $O(m \log n)$ when using a min-heap. The representation does not change the asymptotic complexity since all representations support $O(1)$ front access and removal, and $O(1)$ append to the output.

</details>

---

##### **6.5. Hybrid QUICK-SORT + COUNTING-SORT** (Mock Midterm 2026, Question E)

A hybrid sorting algorithm: run QUICK-SORT but stop when the subarray size is $\le k$, then apply COUNTING-SORT to each small block (keys in $[0, R]$), and concatenate the results.

1. Worst-case time complexity in terms of $n$, $k$, $R$.
2. If $k = \Theta(\log n)$ and $R = O(n)$, what is the overall worst-case complexity?

<details><summary>Click to see the solution</summary>

**Part 1: General complexity**

**QUICK-SORT phase:** QUICK-SORT worst case produces $O(n/k)$ subarrays of size $\le k$ (since each level of recursion splits until size $\le k$). The QUICK-SORT phase itself takes $O(n^2/k)$ in the worst case (since we do $O(n)$ work per level and can have $O(n/k)$ levels with bad pivots... more carefully):

- Worst case of QUICK-SORT on $n$ elements until subarrays reach size $k$: $O(n^2)$ in absolute worst case (QUICK-SORT can degenerate even before reaching $k$). But more precisely, the partial QUICK-SORT that stops at size $k$ does $O(n^2/k) \cdot O(k) = O(n^2)$ comparisons in the absolute worst case.

Actually, let's think carefully. QUICK-SORT running until subarray size $\le k$: the recursion tree has $O(n/k)$ leaves. In the worst case (always worst pivot), the total work is $O(n \cdot n/k) = O(n^2/k)$... No.

The standard QUICK-SORT worst case is $O(n^2)$ regardless of cutoff $k$ (because with bad pivots, the first $n - k$ levels of recursion each do $O(n)$ work). So the QUICK-SORT phase is $O(n^2)$ worst case.

**COUNTING-SORT phase:** There are $O(n/k)$ subarrays of size at most $k$. Each COUNTING-SORT on a subarray of size $s \le k$ with keys in $[0, R]$ takes $O(s + R)$. Total: $O(n + (n/k) \cdot R)$.

**Concatenation:** $O(n)$.

**Total worst case:** $O(n^2 + n + (n/k) \cdot R) = O(n^2 + nR/k)$.

$$\boxed{O(n^2 + nR/k)}$$

**Part 2: With $k = \Theta(\log n)$ and $R = O(n)$**

$nR/k = O(n \cdot n / \log n) = O(n^2 / \log n)$.

Total: $O(n^2 + n^2/\log n) = O(n^2)$.

$$\boxed{O(n^2)}$$

Note: The hybrid approach doesn't improve the worst case because QUICK-SORT still degenerates. In the **average case**, QUICK-SORT phase takes $O(n \log n)$, giving total $O(n \log n + nR/k) = O(n \log n + n \cdot n/\log n) = O(n^2/\log n)$.

</details>

---

##### **6.6. BUCKET-SORT Application** (Mock Midterm 2026, Question F)

Apply BUCKET-SORT to: $0.92, 0.21, 0.03, 0.55, 0.07, 0.41, 0.25, 0.12, 0.67, 0.05$

Use 10 buckets for ranges $[0.0, 0.1), [0.1, 0.2), \ldots, [0.9, 1.0]$.

1. Contents of each bucket (before sorting inside).
2. Sorted sequence.

<details><summary>Click to see the solution</summary>

**Part 1: Bucket contents**

| Bucket | Range | Elements |
|--------|-------|----------|
| 0 | $[0.0, 0.1)$ | 0.03, 0.07, 0.05 |
| 1 | $[0.1, 0.2)$ | 0.12 |
| 2 | $[0.2, 0.3)$ | 0.21, 0.25 |
| 3 | $[0.3, 0.4)$ | (empty) |
| 4 | $[0.4, 0.5)$ | 0.41 |
| 5 | $[0.5, 0.6)$ | 0.55 |
| 6 | $[0.6, 0.7)$ | 0.67 |
| 7 | $[0.7, 0.8)$ | (empty) |
| 8 | $[0.8, 0.9)$ | (empty) |
| 9 | $[0.9, 1.0]$ | 0.92 |

**Part 2: Sort each bucket, concatenate**

- Bucket 0: sort $\{0.03, 0.07, 0.05\}$ → $0.03, 0.05, 0.07$
- Bucket 1: $0.12$
- Bucket 2: sort $\{0.21, 0.25\}$ → $0.21, 0.25$
- Bucket 3: (empty)
- Bucket 4: $0.41$
- Bucket 5: $0.55$
- Bucket 6: $0.67$
- Bucket 7: (empty)
- Bucket 8: (empty)
- Bucket 9: $0.92$

**Sorted sequence:** $0.03, 0.05, 0.07, 0.12, 0.21, 0.25, 0.41, 0.55, 0.67, 0.92$

</details>

---

##### **6.7. True or False** (Mock Midterm 2026, Question G)

Determine TRUE or FALSE for each statement:

1. $n^2 = O(n^{3/2})$
2. In a BST, the node with maximum key has no left child
3. HEAP-SORT has $O(n)$ worst-case time complexity
4. RADIX-SORT (fixed digit range) sorts $n$ integers in $\Theta(n)$ time regardless of number of digits
5. In ARRAYQUEUE, ENQUEUE(x) has $\Theta(n)$ worst-case time complexity
6. Height of an AVL tree with $n$ nodes is $\Theta(\log n)$
7. For $T(n) = 2T(n/2) + n$, master theorem gives $T(n) = \Theta(n)$
8. In dynamic programming, overlapping subproblems are solved at least twice
9. MERGE-SORT is a comparison-based sorting algorithm
10. A red-black tree with $n$ internal nodes has height at least $2\log_2(n+1)$

<details><summary>Click to see the solution</summary>

**1. $n^2 = O(n^{3/2})$ — FALSE.**

$O(n^{3/2})$ means the function is bounded above by $c \cdot n^{3/2}$. But $n^2 / n^{3/2} = n^{1/2} \to \infty$, so $n^2$ is not $O(n^{3/2})$.

---

**2. In a BST, the node with maximum key has no left child — FALSE.**

The node with maximum key has no **right** child (since no key is larger). It can have a left child. For example: insert 5, then 3 into a BST. Root is 5, left child is 3. Node 5 has maximum key and has a left child.

---

**3. HEAP-SORT has $O(n)$ worst-case time complexity — FALSE.**

HEAP-SORT has $\Theta(n \log n)$ worst-case time complexity.

---

**4. RADIX-SORT sorts $n$ integers in $\Theta(n)$ time regardless of number of digits — FALSE.**

RADIX-SORT with $d$ digits and base $b$ runs in $\Theta(d(n + b))$. If $d$ grows (e.g., $d = \Theta(\log n)$), the complexity is $\Theta(n \log n)$, not $\Theta(n)$.

---

**5. In ARRAYQUEUE, ENQUEUE(x) has $\Theta(n)$ worst-case time complexity — FALSE.**

ENQUEUE in a circular array queue is $O(1)$ amortized. With dynamic resizing, the worst case for a single ENQUEUE can be $O(n)$ (due to copying), but this is typically counted as $O(1)$ amortized. A well-implemented ARRAYQUEUE uses a fixed or circular buffer, giving $O(1)$ worst case per ENQUEUE without resizing.

Depending on interpretation: if the array must resize, then TRUE; if using circular buffer with preallocated space, then FALSE.

**FALSE** (standard circular array implementation: $O(1)$ worst case).

---

**6. Height of an AVL tree with $n$ nodes is $\Theta(\log n)$ — TRUE.**

By AVL property, height is $O(\log n)$ and $\Omega(\log n)$, so $\Theta(\log n)$.

---

**7. For $T(n) = 2T(n/2) + n$, master theorem gives $T(n) = \Theta(n)$ — FALSE.**

$a=2, b=2, f(n)=n, n^{\log_2 2} = n$. This is Case 2: $T(n) = \Theta(n \log n)$, not $\Theta(n)$.

---

**8. In dynamic programming, overlapping subproblems are solved at least twice — FALSE.**

The key benefit of DP is that overlapping subproblems are solved **only once** and their results are cached/reused. Without DP (plain recursion), they'd be solved multiple times.

---

**9. MERGE-SORT is a comparison-based sorting algorithm — TRUE.**

MERGE-SORT compares elements to determine order, so yes, it is comparison-based.

---

**10. A red-black tree with $n$ internal nodes has height at least $2\log_2(n+1)$ — FALSE.**

The known bound is height $\le 2\log_2(n+1)$ (upper bound). A red-black tree can have height as small as $\log_2(n+1)$ (a perfect binary tree). So the height is at most $2\log_2(n+1)$, not at least.

</details>

---

##### **6.8. BST Construction and Deletions** (Mock Midterm 2026, Question H)

Start from an empty BST and insert keys in order: 15, 8, 22, 4, 11, 19, 30, 2, 6, 25.

Use array representation with BFS indexing (root at 0, left child of $i$ at $2i+1$, right child at $2i+2$).

1. Draw the constructed BST.
2. Fill the array representation (indices 0–15) after all insertions.
3. Array after deleting key 30.
4. Array after deleting key 8 (from tree that already had 30 deleted).
5. Draw the final BST.

<details><summary>Click to see the solution</summary>

**Part 1: Constructing the BST**

Insert in order: 15, 8, 22, 4, 11, 19, 30, 2, 6, 25.

```
         15
        /   \
       8     22
      / \   /  \
     4  11 19  30
    / \       /
   2   6    25
```

**Part 2: Array representation (BFS order)**

| idx | 0  | 1 | 2  | 3 | 4  | 5  | 6  | 7 | 8 | 9 | 10 | 11 | 12 | 13 | 14 | 15 |
|-----|----|---|----|---|----|----|----|---|---|---|----|----|----|----|----|----|
| key | 15 | 8 | 22 | 4 | 11 | 19 | 30 | 2 | 6 | – | –  | –  | –  | 25 | –  | –  |

Verification of indices:
- 15 at idx 0
- 8 at idx 1 (left of 15), 22 at idx 2 (right of 15)
- 4 at idx 3 (left of 8), 11 at idx 4 (right of 8)
- 19 at idx 5 (left of 22), 30 at idx 6 (right of 22)
- 2 at idx 7 (left of 4), 6 at idx 8 (right of 4)
- 25: left of 30. Left of idx 6 is $2 \times 6 + 1 = 13$. So 25 at idx 13.

**Part 3: Delete key 30**

Node 30 is at idx 6. It has one child: 25 (at idx 13). Replace 30 with its only child 25. Move 25 to idx 6, clear idx 13.

| idx | 0  | 1 | 2  | 3 | 4  | 5  | 6  | 7 | 8 | 9 | 10 | 11 | 12 | 13 | 14 | 15 |
|-----|----|---|----|---|----|----|----|---|---|---|----|----|----|----|----|----|
| key | 15 | 8 | 22 | 4 | 11 | 19 | 25 | 2 | 6 | – | –  | –  | –  | –  | –  | –  |

**Part 4: Delete key 8 (from tree after deleting 30)**

Current tree:
```
         15
        /   \
       8     22
      / \   /  \
     4  11 19  25
    / \
   2   6
```

Node 8 is at idx 1. It has two children: 4 (idx 3) and 11 (idx 4).

In-order successor of 8: go right to 11 (idx 4), then leftmost — 11 has no left child, so successor is 11.

Replace key at idx 1 with 11. Remove 11 from idx 4.

| idx | 0  | 1  | 2  | 3 | 4 | 5  | 6  | 7 | 8 | 9 | 10 | 11 | 12 | 13 | 14 | 15 |
|-----|----|----|----|----|---|----|----|---|---|---|----|----|----|----|----|----|
| key | 15 | 11 | 22 | 4 | – | 19 | 25 | 2 | 6 | – | –  | –  | –  | –  | –  | –  |

**Part 5: Final BST**

```
         15
        /   \
      11     22
      /     /  \
     4     19  25
    / \
   2   6
```

</details>

---

##### **6.9. LCS Problem** (Mock Midterm 2026, Question I)

Longest Common Subsequence (LCS) problem. Given sequences $X$ and $Y$:

1. Worst-case time and space complexity using DP with tabulation ($|X| = m$, $|Y| = n$).
2. Find the LCS length for $X = \langle A,B,C,A,B \rangle$ and $Y = \langle B,A,C,B,A \rangle$. Give one LCS.
3. Fill the DP table $C[i,j]$ (LCS length of $X[1..i]$ and $Y[1..j]$).

<details><summary>Click to see the solution</summary>

**Part 1: Complexity**

The DP table has $(m+1) \times (n+1)$ entries, each computed in $O(1)$.

- **Time complexity:** $\Theta(mn)$
- **Space complexity:** $\Theta(mn)$ (for the full table); $\Theta(\min(m,n))$ with space optimization.

**Part 2 & 3: DP Table**

$X = \langle A,B,C,A,B \rangle$ (rows), $Y = \langle B,A,C,B,A \rangle$ (columns).

Recurrence:
$$C[i,j] = \begin{cases} 0 & \text{if } i=0 \text{ or } j=0 \\ C[i-1,j-1]+1 & \text{if } X[i]=Y[j] \\ \max(C[i-1,j], C[i,j-1]) & \text{otherwise} \end{cases}$$

|   |   | B | A | C | B | A |
|---|---|---|---|---|---|---|
|   | **0** | **1** | **2** | **3** | **4** | **5** |
| **A** (1) | 0 | 0 | 1 | 1 | 1 | 1 |
| **B** (2) | 0 | 1 | 1 | 1 | 2 | 2 |
| **C** (3) | 0 | 1 | 1 | 2 | 2 | 2 |
| **A** (4) | 0 | 1 | 2 | 2 | 2 | 3 |
| **B** (5) | 0 | 1 | 2 | 2 | 3 | 3 |

Full table $C[i,j]$ for $i=0..5$, $j=0..5$:

| $i \backslash j$ | 0 | 1 (B) | 2 (A) | 3 (C) | 4 (B) | 5 (A) |
|------------------|---|-------|-------|-------|-------|-------|
| 0                | 0 | 0 | 0 | 0 | 0 | 0 |
| 1 (A)            | 0 | 0 | 1 | 1 | 1 | 1 |
| 2 (B)            | 0 | 1 | 1 | 1 | 2 | 2 |
| 3 (C)            | 0 | 1 | 1 | 2 | 2 | 2 |
| 4 (A)            | 0 | 1 | 2 | 2 | 2 | 3 |
| 5 (B)            | 0 | 1 | 2 | 2 | 3 | 3 |

**LCS length: 3.**

**One LCS:** Trace back from $C[5,5]=3$:
- $C[5,5]=3$: $X[5]=B \ne Y[5]=A$, go to $\max(C[4,5], C[5,4]) = \max(3,3)$, go up to $C[4,5]$.
- $C[4,5]=3$: $X[4]=A = Y[5]=A$ ✓, take this match. Go to $C[3,4]=2$.
- $C[3,4]=2$: $X[3]=C \ne Y[4]=B$, go to $\max(C[2,4], C[3,3]) = \max(2,2)$, go up to $C[2,4]$.
- $C[2,4]=2$: $X[2]=B = Y[4]=B$ ✓, take this match. Go to $C[1,3]=1$.
- $C[1,3]=1$: $X[1]=A \ne Y[3]=C$, go to $\max(C[0,3], C[1,2]) = \max(0,1)$, go to $C[1,2]$.
- $C[1,2]=1$: $X[1]=A = Y[2]=A$ ✓, take this match. Go to $C[0,1]=0$. Done.

**LCS:** $\langle A, B, A \rangle$

**Answer:** LCS length is **3**, one LCS is $\langle A, B, A \rangle$.

</details>

---

##### **6.10. AVL Tree Rotation** (Mock Midterm 2026, Question J)

AVL tree stored in array representation (indices 0–15):

| idx | 0 | 1 | 2 | 3 | 4 | 5 | 6 | 7 | 8 | 9–15 |
|-----|---|---|---|---|---|---|---|---|---|------|
| key | 7 | 4 | 9 | 1 | 5 | 8 |10 | – | 2 | –    |

Insert key 3 using BST insertion (without rebalancing). Which rotation(s) restore the AVL invariant? Give the pair (parent, child) for each rotation.

<details><summary>Click to see the solution</summary>

**Step 1: Reconstruct the AVL tree.**

BFS indexing:
- idx 0: key 7 (root)
- idx 1: key 4 (left of 7)
- idx 2: key 9 (right of 7)
- idx 3: key 1 (left of 4)
- idx 4: key 5 (right of 4)
- idx 5: key 8 (left of 9)
- idx 6: key 10 (right of 9)
- idx 7: – (left of 1, empty)
- idx 8: key 2 (right of 1)

Tree structure:
```
           7
         /   \
        4     9
       / \   / \
      1   5 8  10
       \
        2
```

Heights: node 2 → h=0, node 1 → h=1, node 5 → h=0, node 4 → h=2, node 8 → h=0, node 10 → h=0, node 9 → h=1, node 7 → h=3.

This is a valid AVL tree (balance factors all in $\{-1, 0, 1\}$).

**Step 2: Insert key 3.**

Path: 3 > 7? No → go left to 4. 3 < 4? Yes → go left to 1. 3 > 1? Yes → go right to 2. 3 > 2? Yes → insert as right child of 2.

Right child of 2: idx $2 \times 8 + 2 = 18$ — exceeds array bounds of 0–15. In abstract tree terms:

```
           7
         /   \
        4     9
       / \   / \
      1   5 8  10
       \
        2
         \
          3
```

**Step 3: Check AVL invariant after insertion.**

Update heights bottom-up:
- node 3: h=0
- node 2: h=1 (right child 3 has h=0, no left child)
- node 1: left h=-1 (empty), right h=1 → balance factor = $-1 - 1 = -2$. **AVL violated at node 1!**

Balance factor of node 1: right subtree height (1) − left subtree height (−1) = 2. Right-heavy.

Node 2 (right child of 1): right subtree height (0) − left subtree height (−1) = 1. Right-heavy.

**Case: Right-Right → single Left rotation.**

**Step 4: Rotation.**

Apply **Left rotation** at node 1 with node 2 as the new root of this subtree.

- Rotation pair: **(parent=1, child=2)**

After rotation:
```
node 2 becomes the subtree root
node 1 becomes left child of 2
node 3 remains right child of 2
```

Subtree rooted at 2:
```
    2
   / \
  1   3
```

Full tree after rotation:
```
           7
         /   \
        4     9
       / \   / \
      1→2  5 8  10
     /
    (2 is now child of 4's left)
```

Wait — let me re-express. Node 1 was the left child of 4. After left rotation at 1:
```
           7
         /   \
        4     9
       / \   / \
      2   5 8  10
     / \
    1   3
```

**Answer:** One **Left rotation** at node 1 with child 2. Pair: **(parent=1, child=2)**.

After rotation, the AVL invariant is restored (all balance factors become 0 or ±1).

</details>

---

##### **6.11. Maximum Subarray Problem** (Mock Midterm 2026, Question K)

For the Maximum Subarray problem (find contiguous subarray with maximum sum), state the worst-case time complexity and brief justification for each approach:

1. Brute Force
2. Divide-and-Conquer
3. Dynamic Programming (Kadane's Algorithm)

<details><summary>Click to see the solution</summary>

**1. Brute Force**

**Complexity:** $\Theta(n^2)$

**Justification:** Enumerate all $O(n^2)$ pairs $(i, j)$ with $i \le j$. For each pair, sum the subarray $A[i..j]$. Using prefix sums, each sum is $O(1)$, giving total $O(n^2)$. (Without prefix sums: $O(n^3)$.)

---

**2. Divide-and-Conquer**

**Complexity:** $\Theta(n \log n)$

**Justification:** Divide the array in half. The maximum subarray either lies entirely in the left half, entirely in the right half, or crosses the midpoint. The crossing case is solved in $O(n)$ (scan left from mid and right from mid). Recurrence: $T(n) = 2T(n/2) + O(n)$, which gives $T(n) = \Theta(n \log n)$ by the Master Theorem (Case 2).

---

**3. Dynamic Programming (Kadane's Algorithm)**

**Complexity:** $\Theta(n)$

**Justification:** Maintain a running variable $\text{maxEndingHere}$: for each position $i$, compute the maximum subarray sum ending at $i$. This is either $A[i]$ alone or $A[i] + \text{maxEndingHere}(i-1)$. One pass through the array suffices. Each element is processed in $O(1)$, so total is $\Theta(n)$.

</details>

---

##### **6.12. Decision Tree Lower Bound** (Mock Midterm 2026, Question L)

1. What is the minimum number of leaves in a decision tree for any comparison-based sorting algorithm on an input of size 6?
2. What is the lower bound on the number of comparisons for sorting $n$ elements asymptotically?

<details><summary>Click to see the solution</summary>

**Part 1: Minimum leaves for $n = 6$**

A comparison-based sorting algorithm must be able to distinguish all $n!$ permutations of the input. The decision tree must have at least $n!$ leaves (one per possible permutation, since each permutation corresponds to a different sorted order).

For $n = 6$:

$$6! = 720$$

**Minimum number of leaves: $\mathbf{720}$**

**Part 2: Asymptotic lower bound**

The height of the decision tree (= number of comparisons in the worst case) satisfies:

$$h \ge \log_2(n!) = \Theta(n \log n)$$

By Stirling's approximation: $\log_2(n!) = n\log_2 n - n\log_2 e + O(\log n) = \Theta(n \log n)$.

**Lower bound: $\Omega(n \log n)$ comparisons.**

Any comparison-based sorting algorithm requires at least $\Omega(n \log n)$ comparisons in the worst case.

</details>

---

##### **6.13. Minimum Nodes in AVL Tree** (Mock Midterm 2026, Question M)

1. Find the minimum number $N(h)$ of internal nodes in an AVL tree of height $h = 4$, $h = 8$, $h = 12$.
2. Draw a valid AVL tree of height 4 with the minimum number of nodes.

<details><summary>Click to see the solution</summary>

**Part 1: $N(h)$ — Fibonacci-like recurrence**

The minimum number of nodes in an AVL tree of height $h$ follows:
$$N(h) = N(h-1) + N(h-2) + 1, \quad N(0) = 1, \quad N(1) = 2$$

Compute:
- $N(0) = 1$
- $N(1) = 2$
- $N(2) = N(1) + N(0) + 1 = 2 + 1 + 1 = 4$
- $N(3) = N(2) + N(1) + 1 = 4 + 2 + 1 = 7$
- $N(4) = N(3) + N(2) + 1 = 7 + 4 + 1 = \mathbf{12}$
- $N(5) = N(4) + N(3) + 1 = 12 + 7 + 1 = 20$
- $N(6) = N(5) + N(4) + 1 = 20 + 12 + 1 = 33$
- $N(7) = N(6) + N(5) + 1 = 33 + 20 + 1 = 54$
- $N(8) = N(7) + N(6) + 1 = 54 + 33 + 1 = \mathbf{88}$
- $N(9) = N(8) + N(7) + 1 = 88 + 54 + 1 = 143$
- $N(10) = N(9) + N(8) + 1 = 143 + 88 + 1 = 232$
- $N(11) = N(10) + N(9) + 1 = 232 + 143 + 1 = 376$
- $N(12) = N(11) + N(10) + 1 = 376 + 232 + 1 = \mathbf{609}$

**Answers:**
- $N(4) = 12$
- $N(8) = 88$
- $N(12) = 609$

**Part 2: AVL tree of height 4 with 12 nodes**

The minimum AVL tree of height $h$ is built recursively: the root has one subtree of height $h-1$ and one of height $h-2$ (both also minimum AVL trees).

```
              r
             / \
           T3   T2
```
where $T3$ = min AVL of height 3 (7 nodes) and $T2$ = min AVL of height 2 (4 nodes).

Assign keys for clarity (keys chosen to satisfy BST property):

```
                   8
                 /   \
               4       12
              / \      / \
             2   6   10   –
            / \ / \  / \
           1  3 5 7 9  11
```

Wait — that's a full tree. Let me draw the actual minimum structure with 12 nodes:

```
              8
            /   \
           4     12
          / \    /
         2   6  10
        /   / \   \
       1   5   7   11
          /
         (none)
```

Actually let me be precise. Min AVL of height 4:
- Root
- Left subtree: min AVL of height 3 (7 nodes)
- Right subtree: min AVL of height 2 (4 nodes)

Min AVL of height 3: root + left (h=2, 4 nodes) + right (h=1, 2 nodes)
Min AVL of height 2: root + left (h=1, 2 nodes) + right (h=0, 1 node)
Min AVL of height 1: root + left (h=0, 1 node)
Min AVL of height 0: single node

Drawing with abstract keys (using label = node number in BFS):

```
                   1
                 /   \
               2       3
              / \      /
             4   5    6
            /   /    / \
           7   8    9   (none? no — h=2 right subtree)
```

More carefully, min AVL of height 4 = 12 nodes:

```
                  ●
                /   \
              ●       ●         ← h=3 left, h=2 right
             / \      /
            ●   ●    ●          ← h=2 left&right of h=3; h=1 left of h=2 right
           /   / \   / \
          ●   ●   ●  ●   ●      ← filling out
         /
        ●
```

Total: 1 + 2 + 3 + 4 + 2 = 12 ✓

A concrete example with keys satisfying BST property:

```
                   7
                 /   \
               3      11
              / \     /
             2   5   9
            /   / \   \
           1   4   6   10
          /
         (impossible — would add node)
```

Hmm, let me just draw the abstract shape:

```
Level 0:          [r]
Level 1:      [L]       [R]
Level 2:   [LL] [LR]  [RL]
Level 3: [LLL][LRL][LRR]
Level 4: [LLLL]
```

Total: 1+2+3+4+2 = 12 nodes (with [LLLL] being the deepest node at height 4 path, and [R] subtree being min AVL of h=2 with 4 nodes: [R],[RL],[RL.left],[RL.right])

**Concrete AVL tree of height 4 with 12 nodes:**

```
                    10
                  /    \
               5          15
              / \          /
             3   7        13
            /   / \       / \
           2   6   8    12   14
          /
         1
```

This has 12 nodes. Heights: node 1 → 0, node 2 → 1, node 6 → 0, node 8 → 0, node 7 → 1, node 3 → 2, node 5 → 3, node 12 → 0, node 14 → 0, node 13 → 1, node 15 → 2, node 10 → 4. ✓

</details>

---

##### **6.14. Red-Black Tree Construction** (Mock Midterm 2026, Question N)

Draw a valid red-black tree with keys 10, 5, 15, 3, 7, 12, 20 inserted in this order. Mark each node red or black.

<details><summary>Click to see the solution</summary>

**Red-Black Tree Properties:**
1. Every node is red or black.
2. The root is black.
3. Every leaf (NIL) is black.
4. If a node is red, both children are black.
5. All paths from any node to its NIL descendants have the same number of black nodes.

**Step-by-step insertion:**

**Insert 10:** Root is black.
```
[10B]
```

**Insert 5:** 5 < 10, goes left. New node is red.
```
   [10B]
   /
 [5R]
```
No violation.

**Insert 15:** 15 > 10, goes right. New node is red.
```
   [10B]
   /    \
 [5R]  [15R]
```
No violation.

**Insert 3:** 3 < 10, 3 < 5, goes left of 5. New node is red. Parent (5) is red — violation!

Uncle is 15 (red) → **Case 1: recolor**. Recolor parent (5) and uncle (15) to black, grandparent (10) to red. But 10 is root → recolor back to black.
```
     [10B]
     /    \
  [5B]   [15B]
  /
[3R]
```
No violation.

**Insert 7:** 7 < 10, 7 > 5, goes right of 5. New node is red. Parent (5) is black → no violation.
```
     [10B]
     /    \
  [5B]   [15B]
  /   \
[3R]  [7R]
```

**Insert 12:** 12 > 10, 12 < 15, goes left of 15. New node is red. Parent (15) is black → no violation.
```
       [10B]
       /    \
    [5B]   [15B]
    /   \   /
  [3R] [7R][12R]
```

**Insert 20:** 20 > 10, 20 > 15, goes right of 15. New node is red. Parent (15) is black → no violation.
```
         [10B]
         /    \
      [5B]   [15B]
      /   \   /   \
    [3R] [7R][12R][20R]
```

**Final Red-Black Tree:**

```
              10 (Black)
            /            \
        5 (Black)       15 (Black)
        /      \         /        \
    3 (Red)  7 (Red)  12 (Red)  20 (Red)
```

**Verification:**
- Root (10) is black ✓
- All red nodes (3, 7, 12, 20) have black parents ✓
- Black-height from root to any NIL: all paths pass through 2 black nodes (root + one of 5 or 15) → black-height = 2 ✓
- All NIL leaves are black (implicit) ✓

**Answer:** The tree is valid. Keys 3, 7, 12, 20 are **Red**; keys 5, 10, 15 are **Black**.

</details>
